---
title: "PASS1b_MOFA_figure panels"
output: html_document
date: "2026-05-21"
---


Load libraries, MOFA output, and create figure directory. 
```{r}
# ---- Libraries ----
library(MOFA2)
library(dplyr)
library(ggplot2)
library(tidyr)
library(forcats)
library(stringr)
library(patchwork)


# -----------------------------------------
# Load model 
# -----------------------------------------
model <- load_model("mofa_trained.out")

# -----------------------------------------
# Output directory
# -----------------------------------------
dir.create("MOFA_figures/outputs", recursive = TRUE, showWarnings = FALSE)
```



This code makes the heatmap which explains the variance per view/ome in each factor.

```{r}
############################################
# Figure 6A
############################################

# -----------------------------------------
# Base plots
# -----------------------------------------
p1 <- plot_variance_explained(model)


# -----------------------------------------
# Clean theme 
# -----------------------------------------
nice_theme <- theme_classic(base_size = 14) +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(color = "black"),
    legend.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5),
    strip.background = element_blank(),
    strip.text = element_blank()
  )

# -----------------------------------------
# Muted color scale
# -----------------------------------------
muted_scale <- scale_fill_gradient(
  low = "#f2f2f2",
  high = "#4d6a7f"
)

# -----------------------------------------
# Apply styling
# -----------------------------------------
p1 <- p1 + nice_theme + muted_scale



# -----------------------------------------
# Save outputs
# -----------------------------------------
ggsave(
  filename = "MOFA_figures/outputs/MOFA_variance_explained_factor.pdf",
  plot = combined,
  width = 8,
  height = 7,
  device = cairo_pdf
)

ggsave(
  filename = "MOFA_figures/outputs/MOFA_variance_explained_factor.png",
  plot = p1,
  width = 4,
  height = 4,
  dpi = 300
)
############################################
```

The following code plots the variece per view


```{r}

############################################
# Figure S6A
############################################


var_exp <- get_variance_explained(model)
df_view <- var_exp$r2_total %>%
  as.data.frame() %>%
  rownames_to_column("view") %>%
  pivot_longer(-view, names_to = "group", values_to = "variance_explained")
df_view <- df_view %>%
  filter(group == "group1")
p_view <- ggplot(df_view,
                 aes(x = fct_reorder(view, variance_explained),
                     y = variance_explained)) +
  geom_col(fill = "#0f6b82", width = 0.8) +
  
  labs(
    x = NULL,
    y = "Variance explained (%)"
  ) +
  
  scale_y_continuous(expand = c(0, 0)) +
  
  theme_classic(base_size = 14) +
  theme(
    axis.title.y = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),  # 👈 tilt here
    axis.line = element_line(size = 0.8),
    axis.ticks = element_line(size = 0.6)
  )

ggsave(
  filename = "MOFA_figures/outputs/MOFA_variance_explained_view.png",
  plot = p_view,
  width = 3,
  height = 4,
  dpi = 600
)

ggsave(
  filename = "MOFA_figures/outputs/MOFA_variance_explained_view.pdf",
  plot = p_view,
  width = 3,
  height = 4,
  dpi = 600
)
```



This code plots the factor weights over time separated by sex, to identify kep patterns in the data.

```{r}

############################################
# Figure 6B/ S6B
############################################

# ---- Extract factors + metadata ----
Z <- get_factors(model, factors = "all")$group1
meta <- samples_metadata(model)

df <- cbind(meta, Z)

# ---- Name factor columns ----
colnames(df)[(ncol(meta) + 1):ncol(df)] <- paste0("Factor", seq_len(ncol(Z)))

# ---- Long format ----
df_long <- df %>%
  pivot_longer(
    cols = starts_with("Factor"),
    names_to = "Factor",
    values_to = "Value"
  )

keep_factors <- paste0("Factor", c(1,4,5,10)) #choose factors to plot, to plot all put "1:10"

df_long$Factor <- factor(df_long$Factor, levels = keep_factors)

# ---- Plot function ----
plot_factors <- function(data,
                         xvar,
                         title,
                         keep_factors = paste0("Factor", 1:10),
                         time_levels = c("SED", "1W", "2W", "4W", "8W"),
                         palette = c(Male = "#5555ff", Female = "#ff6eff")) {
  
  stopifnot(xvar %in% names(data))
  stopifnot(all(c("Factor", "Value", "sex") %in% names(data)))
  
  data <- data |>
    filter(Factor %in% keep_factors) |>
    mutate(
      Factor = factor(Factor, levels = keep_factors),
      "{xvar}" := factor(.data[[xvar]], levels = time_levels)
    )
  
  ggplot(data, aes(x = .data[[xvar]], y = Value, fill = sex)) +
    geom_boxplot(alpha = 0.6,
                 width = 0.7,
                 outlier.shape = NA,
                 position = position_dodge(width = 0.75)) +
    geom_point(aes(color = sex),
               position = position_jitterdodge(jitter.width = 0.12,
                                               dodge.width = 0.75),
               size = 1.3,
               alpha = 0.65,
               show.legend = FALSE) +
    facet_wrap(~Factor, scales = "free_y", ncol = 2) + #set number of col/ rows for figure layout
    scale_fill_manual(values = palette) +
    scale_color_manual(values = palette) +
    labs(title = title,
         x = xvar,
         y = "Factor value",
         fill = "Sex") +
    theme_classic(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold"),
      axis.text.x = element_text(angle = 45, hjust = 1),
      strip.background = element_rect(fill = "grey92", color = NA),
      strip.text = element_text(face = "bold"),
      legend.position = "right"
    )
}

# ---- Generate plot ----
p <- plot_factors(df_long, "timepoint", "MOFA Factors by Timepoint")

# ---- Save output ----
dir.create("figures/mofa", recursive = TRUE, showWarnings = FALSE)

ggsave(
  filename = "MOFA_figures/outputs/MOFA_factors_by_timepoint_main_panel.png",
  plot = p,
  width = 4.7, #4.7 for main, 6 for supp figure panels
  height = 4,
  dpi = 600
)

ggsave(
  filename = "MOFA_figures/outputs/MOFA_factors_by_timepoint_main_panel.pdf",
  plot = p,
  width = 4.7, #4.7 for main, 6 for supp figure panels
  height = 4,
  dpi = 600
)
```


This code plots the enrichment for selected factors which were manually curated. You will need the enrichment files for each view/ome and a selected list of terms with Factor and View included in the format below. 

```{r}
############################################
# Figure 6C/ S6C
############################################

# ----------------------------
# MANUAL CURATION TABLE
# ----------------------------
curated_terms <- tibble::tribble(
  ~view, ~factor, ~feature_set,
  
  "metabolomics", "Factor1", "PC",
  
  "phosphoproteomics", "Factor1", "GOBP_RESPONSE_TO_HORMONE",
  "phosphoproteomics", "Factor1", "GOBP_RESPONSE_TO_INSULIN",
  "phosphoproteomics", "Factor1", "GOBP_RESPONSE_TO_NITROGEN_COMPOUND",
  "phosphoproteomics", "Factor1", "GOCC_SARCOLEMMA",
  "phosphoproteomics", "Factor1", "GOBP_CELLULAR_RESPONSE_TO_PEPTIDE",
  
  "proteomics", "Factor1", "GOCC_BLOOD_MICROPARTICLE",
  "proteomics", "Factor1", "GOCC_VESICLE_LUMEN",
  "proteomics", "Factor1", "GOBP_TISSUE_HOMEOSTASIS",
  "proteomics", "Factor1", "GOBP_ACUTE_INFLAMMATORY_RESPONSE",
  "proteomics", "Factor1", "GOBP_NEGATIVE_REGULATION_OF_PROTEOLYSIS",
  "proteomics", "Factor1", "GOBP_REPRODUCTION",
  
  "redox", "Factor1", "GOBP_RESPONSE_TO_LIPID",
  "redox", "Factor1", "GOBP_CARBOHYDRATE_HOMEOSTASIS",
  "redox", "Factor1", "GOBP_REACTIVE_OXYGEN_SPECIES_METABOLIC_PROCESS",
  "redox", "Factor1", "GOBP_IMMUNE_EFFECTOR_PROCESS",
  "redox", "Factor1", "GOBP_HORMONE_TRANSPORT",
  
  #"proteomics", "Factor3", "GOCC_MITOCHONDRIAL_ENVELOPE",
  #"proteomics", "Factor3", "GOBP_ATP_METABOLIC_PROCESS",
  #"proteomics", "Factor3", "GOBP_GLUCOSE_METABOLIC_PROCESS",
  #"proteomics", "Factor3", "GOCC_RESPIRASOME",
  #"proteomics", "Factor3", "GOCC_MITOCHONDRION",
  #"proteomics", "Factor3", "GOBP_CARBOHYDRATE_BIOSYNTHETIC_PROCESS",
  #"proteomics", "Factor3", "GOBP_REGULATION_OF_APOPTOTIC_SIGNALING_PATHWAY",
  #"proteomics", "Factor3", "GOBP_ATP_BIOSYNTHETIC_PROCESS",
  
  #"phosphoproteomics", "Factor3", "GOBP_GLYCOLYTIC_PROCESS",
  #"phosphoproteomics", "Factor3", "GOCC_SUPRAMOLECULAR_COMPLEX",
  #"phosphoproteomics", "Factor3", "GOMF_ACTIN_BINDING",
  #"phosphoproteomics", "Factor3", "GOCC_CONTRACTILE_FIBER",
  #"phosphoproteomics", "Factor3", "GOBP_REGULATION_OF_MUSCLE_SYSTEM_PROCESS",
  
  #"redox", "Factor3", "GOBP_LIPID_LOCALIZATION",
  #"redox", "Factor3", "GOBP_ORGANIC_ANION_TRANSPORT",
  #"redox", "Factor3", "GOBP_MONOSACCHARIDE_BIOSYNTHETIC_PROCESS",
  
  #"metabolomics", "Factor3", "TG",
  
  #"transcriptomics", "Factor3", "GOBP_REGULATION_OF_LIPID_METABOLIC_PROCESS",
  #"transcriptomics", "Factor3", "GOBP_RESPONSE_TO_BACTERIUM",
  #"transcriptomics", "Factor3", "GOBP_FEEDING_BEHAVIOR",
  
  "metabolomics", "Factor4", "TG",
  "metabolomics", "Factor4", "PC",
  
  "phosphoproteomics", "Factor4", "GOBP_ESTABLISHMENT_OF_PROTEIN_LOCALIZATION_TO_MEMBRANE",
  
  "proteomics", "Factor4", "GOBP_INFLAMMATORY_RESPONSE",
  "proteomics", "Factor4", "GOBP_MONOATOMIC_ION_TRANSPORT",
  "proteomics", "Factor4", "GOBP_PROTEIN_FOLDING",
  
  "transcriptomics", "Factor4", "GOBP_CARBOHYDRATE_BIOSYNTHETIC_PROCESS",
  
  
  "phosphoproteomics", "Factor5", "GOBP_MACROAUTOPHAGY",
  "phosphoproteomics", "Factor5", "GOBP_REGULATION_OF_MACROAUTOPHAGY",
  
  "proteomics", "Factor5", "GOCC_MITOCHONDRION",
  "proteomics", "Factor5", "RUBENSTEIN_SKELETAL_MUSCLE_SMOOTH_MUSCLE_CELLS",
  "proteomics", "Factor5", "GOCC_MITOCHONDRIAL_MATRIX",
  "proteomics", "Factor5", "GOCC_MITOCHONDRIAL_ENVELOPE",
  
  "transcriptomics", "Factor5", "GOMF_PEPTIDE_BINDING",
  "transcriptomics", "Factor5", "GOMF_AMIDE_BINDING",
  "transcriptomics", "Factor5", "GOCC_CILIUM",
  
  "redox", "Factor5", "GOBP_CALCIUM_MEDIATED_SIGNALING",
  
  "metabolomics", "Factor5", "LPC",
  "metabolomics", "Factor5", "PC",
  
  "proteomics", "Factor10", "GOBP_AEROBIC_RESPIRATION",
  "proteomics", "Factor10", "GOBP_CELLULAR_OXIDANT_DETOXIFICATION",
  "proteomics", "Factor10", "GOBP_ATP_BIOSYNTHETIC_PROCESS",
  "proteomics", "Factor10", "GOCC_INNER_MITOCHONDRIAL_MEMBRANE_PROTEIN_COMPLEX",
  
  "redox", "Factor10", "GOMF_EXTRACELLULAR_MATRIX_STRUCTURAL_CONSTITUENT",
  "redox", "Factor10", "GOBP_SKELETAL_SYSTEM_DEVELOPMENT",
  "redox", "Factor10", "GOCC_SARCOLEMMA",
  "redox", "Factor10", "GOBP_BLOOD_VESSEL_MORPHOGENESIS",
  
  "metabolomics", "Factor10", "PC",
  
)
# ----------------------------
# LOAD + COMBINE
# ----------------------------
files <- list(
  metabolomics      = "enrichment_metabolomics.csv",
  phosphoproteomics = "enrichment_phosphoproteomics.csv",
  proteomics        = "enrichment_proteomics.csv",
  redox             = "enrichment_redox.csv",
  transcriptomics   = "enrichment_transcriptomics.csv"
)

MOFA_enrichments <- files %>%
  map(read_csv) %>%
  bind_rows(.id = "view") %>%
  distinct(view, factor, feature_set, .keep_all = TRUE)

# ----------------------------
# PARAMETERS
# ----------------------------
truncate_with_dots <- function(x, n = 40) {
  ifelse(nchar(x) > n,
         paste0(substr(x, 1, n - 3), "..."),
         x)
}
keep_factors <- c("Factor1", "Factor4","Factor5", "Factor10")
padj_cutoff <- 0.05

omic_cols <- c(transcriptomics = "#377EB8", proteomics = "#4DAF4A",phosphoproteomics = "#FF7F00",redox =  "#74add1",metabolomics = "#E41A1C")

omic_order <- c(
  "transcriptomics",
  "proteomics",
  "phosphoproteomics",
  "redox",
  "metabolomics"
)

# ----------------------------
# PROCESS DATA (CURATED ONLY)
# ----------------------------
plot_df <- MOFA_enrichments %>%
  filter(factor %in% keep_factors) %>%
  
  inner_join(
    curated_terms %>% distinct(view, factor, feature_set),
    by = c("view", "factor", "feature_set")
  ) %>%
  
  mutate(
    logp = -log10(padj),
    is_significant = padj < padj_cutoff,
    direction = sign,
    factor = factor(factor, levels = keep_factors),,
    view = factor(view, levels = omic_order), 
    
    # NEW: truncated labels for plotting
    feature_set_plot = truncate_with_dots(feature_set, 35)
  ) %>%
  arrange(view, factor, desc(logp))

# stable ordering
plot_df <- plot_df %>%
  mutate(feature_set_plot = factor(feature_set_plot,
                                   levels = rev(unique(feature_set_plot))))

# ----------------------------
# STRIP DATA
# ----------------------------
strip_df <- plot_df %>%
  distinct(view, feature_set_plot) %>%
  mutate(x = 0.6)

# ----------------------------
# PLOT
# ----------------------------
p <- ggplot(plot_df, aes(x = factor, y = feature_set_plot)) +
  
  geom_tile(
    data = strip_df,
    aes(x = x, y = feature_set_plot, fill = view),
    width = 0.3,
    inherit.aes = FALSE
  ) +
  scale_fill_manual(values = omic_cols, name = "Omic layer") +
  
  geom_point(
    aes(
      size = logp,
      color = direction,
      alpha = is_significant
    )
  ) +
  
  scale_color_manual(
    values = c(
      positive = "#D55E00",
      negative = "#0072B2"
    )
  ) +
  
  scale_alpha_manual(
    values = c(`TRUE` = 1, `FALSE` = 0.3),
    guide = "none"
  ) +
  
  scale_size(range = c(1.5, 5), name = expression(-log[10](padj))) +
  
  scale_x_discrete(expand = expansion(add = c(0.6, 0.4))) +
  
  facet_grid(view ~ ., scales = "free_y", space = "free_y") +
  
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    
    axis.text.y = element_text(size = 9, color = "black"),
    axis.text.x = element_text(
      size = 11,
      face = "bold",
      angle = 45,
      hjust = 1,
      vjust = 1
    ),
    
    strip.text.y = element_blank(),
    strip.background = element_blank(),
    
    legend.position = "right",
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10),
    
    legend.key.size = unit(0.5, "cm"),
    panel.spacing = unit(0.8, "lines")
  ) +
  
  labs(
    x = NULL,
    y = NULL,
    color = "Direction"
  )

p

# ----------------------------
# SAVE
# ----------------------------
ggsave(filename = "MOFA_figures/outputs/MOFA_curated_enrichments.png",
       p,
       width = 6.2,
       height = 8,
       dpi = 600)

ggsave("MOFA_figures/outputs/MOFA_curated_enrichments.pdf",
       p,
       width = 6.2,
       height = 8,
       dpi = 600)
```


