---
title: "MOFA Exploration"
output: html_document
date: "2026-03-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MOFA2)
library(Biobase)
library(dplyr)
library(edgeR)
library(tidyr)
library(here)
library(MotrpacRatTraining6moMuscleData)
source(file.path(here(), "Other_Figures", "MOFA", "mofa_helpers.R"))

mofa_out = file.path(here(), "Other_Figures", "MOFA", "saved_outputs")
```

# Overview

Multi-Omics Factor Analysis (MOFA) omic layers: transcriptomics, global proteomics, and
(protein-normalized) phosphoproteomics.

# Prepare each ome

Features with any missing values are removed, and where multiple features map
to the same symbol, the first is kept (or highest-variance for phosphosites, to 
meet the recommendations of the MOFA tool in picking higher variance features).

**Note: possible complete-case bias:** MOFA requires complete data across all
samples for each feature. Proteomics and phosphoproteomics contain missing
values, so we restrict to complete cases before selecting highly variable
features. This may bias the feature set toward higher-abundance proteins or
phosphosites that are detected in all samples, and enrichment results should
be interpreted with this in mind.


## Transcriptomics

Log-CPM values are computed from raw counts. Ensembl IDs are mapped to human
gene symbols via `RAT_TO_HUMAN`. 
```{r}
describe_eset(TRNSCRPT_GN)

gene_symbol_map = RAT_TO_HUMAN %>% select(ensembl, human_symbol)
trnscrpt_mat = cpm(exprs(TRNSCRPT_GN), log = TRUE) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("ensembl") %>%
  left_join(gene_symbol_map, by = "ensembl") %>%
  group_by(human_symbol) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  filter(!is.na(human_symbol)) %>%
  tibble::column_to_rownames("human_symbol") %>%
  select(-"ensembl") %>%
  as.matrix()

cat("Complete features:", nrow(trnscrpt_mat), "\n")
```

## Proteomics

Normalized log2 abundance. Rows with any missing values are dropped.

```{r}
describe_eset(PROT_GN)

gene_symbol_map = fData(PROT_GN) %>% select(protein_id, gene_symbol)

prot_mat = exprs(PROT_GN) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("protein_id") %>%
  left_join(gene_symbol_map, by = "protein_id") %>%
  group_by(gene_symbol) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  filter(!is.na(gene_symbol)) %>%
  tibble::column_to_rownames("gene_symbol") %>%
  select(-"protein_id") %>%
  as.matrix()
prot_mat = prot_mat[complete.cases(prot_mat), ]

cat("Complete features:", nrow(prot_mat), "\n")
```

## Phosphoproteomics (Protein-Normalized)

See vignette for KSEA gene set testing for mapping info. 

```{r}
describe_eset(PHOSPHO_GN_NORM)

# Map rat phospho features to human UniProt-site format (matches kinase_sets)
site_map = fData(PHOSPHO_GN_NORM) %>%
  tibble::rownames_to_column("feature_id") %>%
  select(feature_id) %>%
  left_join(RAT_TO_HUMAN_SITE,
            by = c("feature_id" = "ptm_id_rat_refseq")) %>%
  rename(human_feature = ptm_id_human_uniprot) %>%
  mutate(
    is_human = !is.na(human_feature),
    human_feature = ifelse(is_human, human_feature, feature_id),
    human_uniprot = ifelse(is_human,
                           sub("_.*", "", human_feature),
                           sub("([A-Z]P_[^_]+).*", "\\1", human_feature)),
    human_site = sub(".*_", "", human_feature),
    human_site = strsplit(human_site, split = ";")
  ) %>%
  unnest(cols = human_site) %>%
  mutate(featureID = paste0(human_uniprot, "-", human_site))

phospho_mat = exprs(PHOSPHO_GN_NORM) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("feature_id") %>%
  left_join(site_map %>% select(feature_id, featureID), by = "feature_id") %>%
  select(-feature_id) %>%
  # For duplicate sites, keep the row with highest variance
  mutate(row_var = apply(select(., -featureID), 1, var, na.rm = TRUE)) %>%
  arrange(featureID, -row_var) %>%
  filter(!duplicated(featureID)) %>%
  select(-row_var) %>%
  tibble::column_to_rownames("featureID") %>%
  as.matrix()
phospho_mat = phospho_mat[complete.cases(phospho_mat), ]

cat("Complete features:", nrow(phospho_mat), "\n")
```

# Select top variance features

Filter to same number of features per ome

```{r}
n_features = min(nrow(trnscrpt_mat), nrow(prot_mat), nrow(phospho_mat))
cat("Target feature count (set by smallest ome):", n_features, "\n")
```

```{r}
trnscrpt_hvf = select_top_var(trnscrpt_mat, n_features)
prot_hvf = select_top_var(prot_mat, n_features)
phospho_hvf = select_top_var(phospho_mat, n_features)

cat("Transcriptomics HVF:", nrow(trnscrpt_hvf), "\n")
cat("Proteomics HVF:", nrow(prot_hvf), "\n")
cat("Phospho HVF:", nrow(phospho_hvf), "\n")
```

# Match samples across omes by pid

Filter to the shared set of animals for the multiple omes (transcrpt has 5 male + 5 female/timepoint, prot has 6 + 6)

```{r}
trnscrpt_hvf = remap_to_pid(trnscrpt_hvf, TRNSCRPT_GN)
prot_hvf = remap_to_pid(prot_hvf, PROT_GN)
phospho_hvf = remap_to_pid(phospho_hvf, PHOSPHO_GN_NORM)

shared_pids = Reduce(intersect, list(
  colnames(trnscrpt_hvf),
  colnames(prot_hvf),
  colnames(phospho_hvf)
))
cat("Shared samples:", length(shared_pids), "\n")

trnscrpt_hvf = trnscrpt_hvf[, shared_pids]
prot_hvf = prot_hvf[, shared_pids]
phospho_hvf = phospho_hvf[, shared_pids]
```

# Build MOFA object

```{r}
mofa_data = list(
  Transcriptomics = trnscrpt_hvf,
  Proteomics = prot_hvf,
  Phosphoproteomics = phospho_hvf
)

mofa_obj = create_mofa(mofa_data)
mofa_obj
```

```{r}
plot_data_overview(mofa_obj)
```

# Add sample metadata

```{r}
# Build metadata from one of the esets (all share the same pid->sex/timepoint mapping)
pd = pData(TRNSCRPT_GN) %>%
  mutate(pid = as.character(pid)) %>%
  filter(pid %in% shared_pids) %>%
  select(sample = pid, sex, timepoint) %>%
  distinct()

samples_metadata(mofa_obj) = pd
```

# Set MOFA options and train

```{r}
data_opts = MOFA2::get_default_data_options(mofa_obj)
data_opts$scale_views = TRUE

model_opts = MOFA2::get_default_model_options(mofa_obj)
model_opts$num_factors = 10

train_opts = MOFA2::get_default_training_options(mofa_obj)

mofa_obj = MOFA2::prepare_mofa(
  mofa_obj,
  data_options = data_opts,
  model_options = model_opts,
  training_options = train_opts
)
```

```{r, results='hide', message=FALSE}
mofa_trained = MOFA2::run_mofa(mofa_obj,
                               outfile = file.path(here(), "Other_Figures", "MOFA", "mofa_trained.out"),
                               use_basilisk = TRUE)
```

# Downstream exploration

## Variance explained

Total and per-factor variance explained by each view.

```{r}
plot_variance_explained(mofa_trained, plot_total = TRUE)
```

```{r}
plot_variance_explained(mofa_trained, max_r2 = 15)
```

## Factor plots

Visualize sample loadings on the first four factors, colored by experimental
variables.

```{r, fig.width=8, fig.height=6}
plot_factor(mofa_trained, factors = 1:4, color_by = "timepoint", shape_by = "sex")
```

```{r, fig.width=8, fig.height=6}
plot_factor(mofa_trained, factors = 1:4, color_by = "sex", shape_by = "timepoint")
```

Factor 2 appears to separate well by sex.

## Factor correlations

```{r}
plot_factor_cor(mofa_trained)
```

## Top weights

Top-weighted features per factor for each view.

```{r, fig.width=8, fig.height=6}
plot_weights(mofa_trained, view = "Transcriptomics", factors = 1:3, nfeatures = 5)
```

```{r, fig.width=8, fig.height=6}
plot_weights(mofa_trained, view = "Proteomics", factors = 1:3, nfeatures = 5)
```

```{r, fig.width=8, fig.height=6}
plot_weights(mofa_trained, view = "Phosphoproteomics", factors = 1:3, nfeatures = 5)
```

## Enrichment: Transcriptomics

```{r}
names = toupper(features_names(mofa_trained)[["Transcriptomics"]])
features_names(mofa_trained)[["Transcriptomics"]] = gsub("_TRANSCRIPTOMICS", "", names)

enrichment_trnscrpt = run_mofa_enrichment(mofa_trained,
  view = "Transcriptomics",
  feature_sets = trnscrpt_GN_gene_sets,
  factors = 1:10,
  csv_path = file.path(mofa_out, "enrichment_transcriptomics.csv")
)
```

```{r}
plot_enrichment_detailed(enrichment_trnscrpt$down,
  factor = 1,
  max.genes = 8,
  max.pathways = 5
)
```

## Enrichment: Proteomics

```{r}
names = toupper(features_names(mofa_trained)[["Proteomics"]])
features_names(mofa_trained)[["Proteomics"]] = gsub("_PROTEOMICS", "", names)

enrichment_prot = run_mofa_enrichment(mofa_trained,
  view = "Proteomics",
  feature_sets = prot_GN_gene_sets,
  factors = 1:10,
  csv_path = file.path(mofa_out, "enrichment_proteomics.csv")
)
```

## Enrichment: Phosphoproteomics

```{r}
features_names(mofa_trained)[["Phosphoproteomics"]] = toupper(features_names(mofa_trained)[["Phosphoproteomics"]])

enrichment_phospho = run_mofa_enrichment(mofa_trained,
  view = "Phosphoproteomics",
  feature_sets = phospho_GN_NORM_kinase_sets,
  factors = 1:10,
  csv_path = file.path(mofa_out, "enrichment_phosphoproteomics.csv")
)
```
