---
title: "MOFA Exploration"
output: html_document
date: "2026-03-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MOFA2)
library(Biobase)
library(dplyr)
library(edgeR)
library(tidyr)
library(ggplot2)
library(here)
library(MotrpacRatTraining6moMuscleData)
library(ComplexHeatmap)
source(file.path(here(), "Other_Figures", "MOFA", "mofa_helpers.R"))

mofa_out = file.path(here(), "Other_Figures", "MOFA", "saved_outputs")
```

# Overview

Multi-Omics Factor Analysis (MOFA) omic layers: transcriptomics, global proteomics,
(protein-normalized) phosphoproteomics, (protein-normalized) redox proteomics, and
metabolomics.

# Prepare each ome

Features with any missing values are removed, and where multiple features map
to the same symbol, the first is kept (or highest-variance for phosphosites, to 
meet the recommendations of the MOFA tool in picking higher variance features).

**Note: possible complete-case bias:** MOFA input is currently complete data across all
samples for each feature. Proteomics and phosphoproteomics contain missing
values, so we restrict to complete cases before selecting highly variable
features. This may bias the feature set toward higher-abundance proteins or
phosphosites that are detected in all samples, and enrichment results should
be interpreted with this in mind.


## Transcriptomics

Log-CPM values are computed from raw counts. Ensembl IDs are mapped to human
gene symbols via `RAT_TO_HUMAN`. 
```{r}
describe_eset(TRNSCRPT_GN)

gene_symbol_map = RAT_TO_HUMAN %>% select(ensembl, human_symbol)
trnscrpt_mat = cpm(exprs(TRNSCRPT_GN), log = TRUE) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("ensembl") %>%
  left_join(gene_symbol_map, by = "ensembl") %>%
  group_by(human_symbol) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  filter(!is.na(human_symbol)) %>%
  tibble::column_to_rownames("human_symbol") %>%
  select(-"ensembl") %>%
  as.matrix()

cat("Complete features:", nrow(trnscrpt_mat), "\n")
```

## Proteomics

Normalized log2 abundance. Rows with any missing values are dropped.

```{r}
describe_eset(PROT_GN)

gene_symbol_map = fData(PROT_GN) %>% select(protein_id, gene_symbol)

prot_mat = exprs(PROT_GN) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("protein_id") %>%
  left_join(gene_symbol_map, by = "protein_id") %>%
  group_by(gene_symbol) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  filter(!is.na(gene_symbol)) %>%
  tibble::column_to_rownames("gene_symbol") %>%
  select(-"protein_id") %>%
  as.matrix()
prot_mat = prot_mat[complete.cases(prot_mat), ]

cat("Complete features:", nrow(prot_mat), "\n")
```

## Phosphoproteomics (Protein-Normalized)

Log2 abundance adjusted for global protein changes. Features mapped to rat gene symbols;
duplicate sites per gene are resolved by highest variance.

```{r}
describe_eset(PHOSPHO_GN_NORM)

site_map = fData(PHOSPHO_GN_NORM) %>%
  tibble::rownames_to_column("feature_id") %>%
  mutate(gene_symbol = toupper(gene_symbol),
         gene_site = paste0(gene_symbol, "_", ptm_id)) %>%
  select(feature_id, gene_symbol, gene_site) 

phospho_mat = exprs(PHOSPHO_GN_NORM) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("feature_id") %>%
  left_join(site_map, by = "feature_id") %>%
  select(-feature_id) %>%
  filter(!is.na(gene_symbol)) %>%
  # compute per-row variance across expression columns only (- excludes string ID cols)
  # then pick the one site per gene with the highest variance
  mutate(row_var = apply(select(., -gene_site, -gene_symbol), 1, var, na.rm = TRUE)) %>%
  group_by(gene_symbol) %>%
  slice_max(row_var, n=1, with_ties=FALSE) %>%
  ungroup() %>%
  select(-row_var, -gene_symbol) %>%
  tibble::column_to_rownames("gene_site") %>%
  as.matrix()

site_map = site_map %>% filter(gene_site %in% rownames(phospho_mat))
phospho_mat = phospho_mat[complete.cases(phospho_mat), ]

cat("Complete features:", nrow(phospho_mat), "\n")
```


## Redox Proteomics (Protein-Normalized)

Log2 abundance adjusted for global protein changes. Features mapped to rat gene symbols;
duplicate sites per gene are resolved by highest variance.

```{r}
describe_eset(REDOX_GN_NORM)

redox_feature_map = fData(REDOX_GN_NORM) %>%
  tibble::rownames_to_column("feature_id") %>%
  mutate(gene_symbol = toupper(gene_symbol),
         gene_site = paste0(gene_symbol, "_", ptm_id)) %>%
  select(feature_id, gene_symbol, gene_site) 

redox_mat = exprs(REDOX_GN_NORM) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("feature_id") %>%
  left_join(redox_feature_map, by = "feature_id") %>%
  select(-feature_id) %>%
  filter(!is.na(gene_symbol)) %>%
  # compute per-row variance across expression columns only (- excludes string ID cols)
  # then pick the one site per gene with the highest variance
  mutate(row_var = apply(select(., -gene_site, -gene_symbol), 1, var, na.rm = TRUE)) %>%
  group_by(gene_symbol) %>%
  slice_max(row_var, n=1, with_ties=FALSE) %>%
  ungroup() %>%
  select(-row_var, -gene_symbol) %>%
  tibble::column_to_rownames("gene_site") %>%
  as.matrix()

redox_feature_map = redox_feature_map %>% filter(gene_site %in% rownames(redox_mat))

redox_mat = redox_mat[complete.cases(redox_mat), ]

cat("Complete features:", nrow(redox_mat), "\n")
```

## Metabolomics

Normalized log2 abundance. Features are mapped to RefMet standardized names; where a
metabolite appears in multiple datasets, the higher variance one is picked

```{r}
describe_eset(METAB_GN)

metab_feature_map = fData(METAB_GN) %>%
  tibble::rownames_to_column("feature_id") %>%
  select(feature_id, refmet_name = metabolite_refmet) %>%
  mutate(refmet_name = toupper(refmet_name))

# How many raw features map to the same refmet name?
repeated_samples = metab_feature_map %>%
  filter(!is.na(refmet_name)) %>%
  count(refmet_name, name = "n_features") %>%
  filter(n_features > 1) %>%
  arrange(desc(n_features)) 

metab_mat = exprs(METAB_GN) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("feature_id") %>%
  left_join(metab_feature_map, by = "feature_id") %>%
  # select(-feature_id) %>%
  filter(!is.na(refmet_name)) %>%
  mutate(row_var = apply(select(., -feature_id, -refmet_name), 1, var, na.rm = TRUE)) %>%
  group_by(refmet_name) %>%
  slice_max(row_var, n=1, with_ties=FALSE) %>%
  ungroup() %>%
  select(-row_var) %>%
  tibble::column_to_rownames("refmet_name") 

metab_da_selected_plat = METAB_GN_DA[["trained_vs_SED"]] %>% filter(featureName %in% metab_mat$feature_id)

metab_mat = metab_mat %>% select(-feature_id) %>% as.matrix()

metab_mat = metab_mat[complete.cases(metab_mat), ]

cat("Complete features:", nrow(metab_mat), "\n")
```

# Select top variance features

Filter to same number of features per ome

```{r}
n_features = min(nrow(trnscrpt_mat), nrow(prot_mat), nrow(phospho_mat),
                 nrow(redox_mat), nrow(metab_mat))
cat("Target feature count (set by smallest ome):", n_features, "\n")
```

```{r}
trnscrpt_hvf = select_top_var(trnscrpt_mat, n_features)
prot_hvf = select_top_var(prot_mat, n_features)
phospho_hvf = select_top_var(phospho_mat, n_features)
redox_hvf = select_top_var(redox_mat, n_features)
metab_hvf = select_top_var(metab_mat, n_features)

cat("Transcriptomics HVF:", nrow(trnscrpt_hvf), "\n")
cat("Proteomics HVF:", nrow(prot_hvf), "\n")
cat("Phospho HVF:", nrow(phospho_hvf), "\n")
cat("Redox HVF:", nrow(redox_hvf), "\n")
cat("Metabolomics HVF:", nrow(metab_hvf), "\n")
```

# Match samples across omes by pid

Filter to the shared set of animals across all omes (transcriptomics has 5M+5F/timepoint,
proteomics/phospho/redox have 6+6; metabolomics has its own sample set).

```{r}
trnscrpt_hvf = remap_to_pid(trnscrpt_hvf, TRNSCRPT_GN)
prot_hvf = remap_to_pid(prot_hvf, PROT_GN)
phospho_hvf = remap_to_pid(phospho_hvf, PHOSPHO_GN_NORM)
redox_hvf = remap_to_pid(redox_hvf, REDOX_GN_NORM)
# METAB_GN columns are already pid values; no remap needed

shared_pids = Reduce(intersect, list(
  colnames(trnscrpt_hvf),
  colnames(prot_hvf),
  colnames(phospho_hvf),
  colnames(redox_hvf),
  colnames(metab_hvf)
))
cat("Shared samples:", length(shared_pids), "\n")

trnscrpt_hvf = trnscrpt_hvf[, shared_pids]
prot_hvf = prot_hvf[, shared_pids]
phospho_hvf = phospho_hvf[, shared_pids]
redox_hvf = redox_hvf[, shared_pids]
metab_hvf = metab_hvf[, shared_pids]
```

# Build MOFA object

```{r}
mofa_data = list(
  Transcriptomics = trnscrpt_hvf,
  Proteomics = prot_hvf,
  Phosphoproteomics = phospho_hvf,
  Redox = redox_hvf,
  Metabolomics = metab_hvf
)

mofa_obj = create_mofa(mofa_data)
mofa_obj
```

```{r}
plot_data_overview(mofa_obj)
```

# Add sample metadata

```{r}
# Build metadata from transcriptomics pData (all omes share pid->sex/timepoint mapping)
pd = pData(TRNSCRPT_GN) %>%
  mutate(pid = as.character(pid)) %>%
  filter(pid %in% shared_pids) %>%
  select(sample = pid, sex, timepoint) %>%
  distinct()

samples_metadata(mofa_obj) = pd
```

# Set MOFA options and train

```{r}
data_opts = MOFA2::get_default_data_options(mofa_obj)
data_opts$scale_views = TRUE

model_opts = MOFA2::get_default_model_options(mofa_obj)
model_opts$num_factors = 10

train_opts = MOFA2::get_default_training_options(mofa_obj)

mofa_obj = MOFA2::prepare_mofa(
  mofa_obj,
  data_options = data_opts,
  model_options = model_opts,
  training_options = train_opts
)
```

```{r, results='hide', message=FALSE, warning=FALSE}
# mofa_trained = MOFA2::run_mofa(mofa_obj,
#                                outfile = file.path(here(), "Other_Figures", "MOFA", "mofa_trained.out"),
#                                use_basilisk = TRUE)

mofa_trained = load_model(
  file.path(here(), "Other_Figures", "MOFA", "mofa_trained.out")
)
```

```{r}
get_weights(mofa_trained, as.data.frame = TRUE) %>%
  write.csv(., file = file.path(here(), "Other_Figures", "MOFA", "all_feature_weights.csv"), row.names = FALSE)
```

# Downstream exploration

## Variance explained

Total and per-factor variance explained by each view.

```{r}
ve_plots = plot_variance_explained(mofa_trained, plot_total = TRUE)
p_ve = patchwork::wrap_plots(ve_plots)
ggsave(file.path(here(), "plots", "MOFA", "variance_explained.png"),
       plot = p_ve, width = 12, height = 5, units = "in", dpi = 600)
p_ve
```

## Factor plots

Visualize sample loadings on the first four factors, colored by experimental
variables. I'm not sure why the plot_factors function is not working well for SEDS. 

Just plot through ggplot

```{r, fig.width=8, fig.height=6}
factor_df = get_factors(mofa_trained, factors = 1:4, as.data.frame = TRUE) %>%
  left_join(samples_metadata(mofa_trained), by = "sample") %>%
  mutate(
    factor = factor(factor, levels = paste0("Factor", 1:4)),
    timepoint = factor(timepoint, levels = c("SED", "1W", "2W", "4W", "8W"))
  )

p_factor_sex = ggplot(factor_df, aes(x = "", y = value, fill = sex, shape = timepoint)) +
  geom_jitter(width = 0.4, height = 0, size = 2.5, color = "black", stroke = 0.4) +
  facet_wrap(~factor, nrow = 1) +
  scale_shape_manual(values = c("SED" = 25, "1W" = 21, "2W" = 24, "4W" = 22, "8W" = 23)) +
  scale_fill_manual(
    values = c(Female = MotrpacBicQC::sex_cols[["Female"]],
               Male   = MotrpacBicQC::sex_cols[["Male"]])
  ) +
  labs(x = NULL, y = "Factor value", fill = "sex", shape = "timepoint") +
  theme_bw() +
  theme(
    axis.text  = element_text(size = 12),
    axis.title = element_text(size = 13),
    strip.text = element_text(size = 12),
    legend.text  = element_text(size = 11),
    legend.title = element_text(size = 12)
  )
ggsave(file.path(here(), "plots", "MOFA", "factor_plot_by_sex.png"),
       plot = p_factor_sex, width = 8, height = 6, units = "in", dpi = 600)
p_factor_sex
```

Factor 1 appears to separate well by sex.

## Factor correlations

```{r}
plot_factor_cor(mofa_trained)
```

## Feature weight plots

```{r feature_weights, fig.width=12, fig.height=10}
for (f in 1:10) {
  p = plot_feature_weights(mofa_trained,
                           factors = f,
                           n_features = 8,
                           png_path = file.path(here(), "plots", "MOFA", paste0("factor", f, "_feature_weights.png")))
  print(p)
}

```

```{r}
#custom plots for panel 2, 3
p = plot_feature_weights(mofa_trained,
                         factors = 2,
                         views = "Redox",
                         n_features = 20,
                         png_path = file.path(here(), "plots", "MOFA", "custom_panels", paste0("factor", 2, "_redox_feature_weights.png")))

p = plot_feature_weights(mofa_trained,
                         factors = 3,
                         views = c("Proteomics",
                                   "Phosphoproteomics",
                                   "Redox"),
                         n_features = 10,
                         png_path = file.path(here(), "plots", "MOFA", "custom_panels", paste0("factor", 3, "_prot_feature_weights.png")))
```

## Enrichment: Transcriptomics

```{r}
names = toupper(features_names(mofa_trained)[["Transcriptomics"]])
features_names(mofa_trained)[["Transcriptomics"]] = gsub("_TRANSCRIPTOMICS", "", names)

enrichment_trnscrpt = run_mofa_enrichment(mofa_trained,
                                          view = "Transcriptomics",
                                          feature_sets = trnscrpt_GN_gene_sets,
                                          factors = 1:10,
                                          csv_path = file.path(mofa_out, "enrichment_transcriptomics.csv")
)
```

## Enrichment: Proteomics

```{r}
names = toupper(features_names(mofa_trained)[["Proteomics"]])
features_names(mofa_trained)[["Proteomics"]] = gsub("_PROTEOMICS", "", names)

enrichment_prot = run_mofa_enrichment(mofa_trained,
                                      view = "Proteomics",
                                      feature_sets = prot_GN_gene_sets,
                                      factors = 1:10,
                                      csv_path = file.path(mofa_out, "enrichment_proteomics.csv")
)
```

## Enrichment: Phosphoproteomics

```{r}
#trim the sites off the names for initial enrichment. then put them back after
phospho_site_names = features_names(mofa_trained)[["Phosphoproteomics"]]
features_names(mofa_trained)[["Phosphoproteomics"]] = sub("_.*", "", toupper(features_names(mofa_trained)[["Phosphoproteomics"]]))

enrichment_phospho = run_mofa_enrichment(mofa_trained,
                                         view = "Phosphoproteomics",
                                         feature_sets = prot_GN_gene_sets,
                                         factors = 1:10,
                                         csv_path = file.path(mofa_out, "enrichment_phosphoproteomics.csv")
)

#and put back the names 
features_names(mofa_trained)[["Phosphoproteomics"]] = phospho_site_names
```

## Enrichment: Redox Proteomics

Feature names are rat gene symbols (touppered to match gene sets).

```{r}
redox_site_names = features_names(mofa_trained)[["Redox"]]
features_names(mofa_trained)[["Redox"]] = sub("_.*", "", toupper(features_names(mofa_trained)[["Redox"]]))

enrichment_redox = run_mofa_enrichment(mofa_trained,
                                       view = "Redox",
                                       feature_sets = prot_GN_gene_sets,
                                       factors = 1:10,
                                       csv_path = file.path(mofa_out, "enrichment_redox.csv")
)

features_names(mofa_trained)[["Redox"]] = redox_site_names
```

## Enrichment: Metabolomics

Feature names are RefMet standardized names (touppered to match refmet sets).

```{r}
features_names(mofa_trained)[["Metabolomics"]] = toupper(features_names(mofa_trained)[["Metabolomics"]])

enrichment_metab = run_mofa_enrichment(mofa_trained,
                                       view = "Metabolomics",
                                       feature_sets = metab_GN_refmet_sets,
                                       factors = 1:10,
                                       csv_path = file.path(mofa_out, "enrichment_metabolomics.csv")
)
```

## Cross-ome enrichment: faceted per factor

Show the top enriched pathways/sets from every ome for a single factor in one panel.
Edit `factor_to_plot` to explore other factors.

```{r, fig.width=9, fig.height=14}
all_enrichments = list(
  Transcriptomics = enrichment_trnscrpt,
  Proteomics = enrichment_prot,
  Phosphoproteomics = enrichment_phospho,
  Redox = enrichment_redox,
  Metabolomics = enrichment_metab
)
```


```{r}
sig_all = load_sig_enrichments(mofa_out)
```

```{r}
for (f in 1:10) {
  p = plot_enrichment_faceted(all_enrichments,
                              factor = f,
                              max_pathways = 8,
                              png_path = file.path(here(), "plots", "MOFA", paste0("factor", f, ".png")))
  print(p)
}

```

Custom factor 2,3 plots
```{r}
p = plot_enrichment_faceted(list(Redox = enrichment_redox),
                            factor = 2,
                            max_pathways = 7,
                            width = 6,
                            png_path = file.path(here(), "plots", "MOFA", "custom_panels", paste0("factor", 2, "_redox_pathways.png")))
print(p)

p = plot_enrichment_faceted(list(Proteomics = enrichment_prot,
                                 Phosphoproteomics = enrichment_phospho,
                                 Redox = enrichment_redox),
                            factor = 3,
                            max_pathways = 7,
                            width = 12,
                            png_path = file.path(here(), "plots", "MOFA","custom_panels",  paste0("factor", 3, "_protein_pathways.png")))
print(p)
```

## Combined enrichment bubble heatmap

Bubble heatmap across all views and factors.  Each row is a top significant
gene set; columns are factors.  Circle size = -log10(padj) (capped at 5 mm),
circle color = signed -log10(padj) (red = up-weighted, blue = down-weighted),
grey background = padj < 0.05.

```{r mofa_enrichment_bubble, message=FALSE}
# Compute global signed -log10(padj) range for a shared colour scale
compute_signed_logp = function(enrich, n_top=8, padj_threshold=0.05) {
  factor_cols = intersect(colnames(enrich$up$pval.adj),
                          colnames(enrich$down$pval.adj))
  pu = enrich$up$pval.adj[, factor_cols, drop=FALSE]
  pd = enrich$down$pval.adj[, factor_cols, drop=FALSE]
  min_p = pmin(apply(pu, 1, min, na.rm=TRUE), apply(pd, 1, min, na.rm=TRUE))
  sig = names(sort(min_p[min_p < padj_threshold]))[seq_len(min(n_top, sum(min_p < padj_threshold)))]
  if (length(sig) == 0) return(numeric(0))
  as.vector(ifelse(pu[sig, , drop=FALSE] <= pd[sig, , drop=FALSE],
                   -log10(pu[sig, , drop=FALSE]),
                   log10(pd[sig, , drop=FALSE])))
}

all_vals = unlist(lapply(all_enrichments, compute_signed_logp))
max_abs_logp = min(max(abs(all_vals), na.rm=TRUE), 20)

col_fun_enrich = circlize::colorRamp2(
  c(-max_abs_logp, 0, max_abs_logp),
  c("#3366ff", "white", "darkred")
)

ome_cols_mofa = c(
  Transcriptomics="#377EB8",
  Proteomics="#4DAF4A",
  Phosphoproteomics="#FF7F00",
  Redox="#74add1",
  Metabolomics="#E41A1C"
)

color_lg = ComplexHeatmap::Legend(
  title="-log10(padj)\n(+ up, \u2212 down)",
  col_fun=col_fun_enrich,
  at=c(-round(max_abs_logp), 0, round(max_abs_logp)),
  legend_height=grid::unit(4, "cm")
)
ome_lg = ComplexHeatmap::Legend(
  title="View",
  at=names(ome_cols_mofa),
  legend_gp=grid::gpar(fill=ome_cols_mofa)
)
sig_lg = ComplexHeatmap::Legend(
  title="Significance",
  at="padj < 0.05",
  legend_gp=grid::gpar(fill="grey90", col=NA)
)

enrich_heatmaps = list()
for (v in names(all_enrichments)) {
  ht = make_mofa_enrichment_heatmap(
    enrichment=all_enrichments[[v]],
    view_name=v,
    ome_cols=ome_cols_mofa,
    col_fun=col_fun_enrich,
    n_top=8
  )
  if (!is.null(ht)) enrich_heatmaps[[v]] = ht
}

if (length(enrich_heatmaps) > 0) {
  combined_enrich = Reduce(`%v%`, enrich_heatmaps)
  png_path = file.path(here(), "plots", "MOFA", "enrichment_bubble_heatmap.png")
  png(png_path, width=12, height=10, units="in", res=600)
  ComplexHeatmap::draw(
    combined_enrich,
    heatmap_legend_list=list(color_lg, ome_lg, sig_lg),
    merge_legend=TRUE,
    heatmap_legend_side="left"
  )
  dev.off()
  cat("Saved:", png_path, "\n")
}
```

## Feature heatmaps

```{r feature_heatmaps, fig.width=10, fig.height=6, results='hide', message=FALSE}
for (f in 1:10) {
  ht = mofa_feature_heatmap_zscore(mofa_trained, factor_num=f, n=8, scale=3)
  if (!is.null(ht)) {
    png_path = file.path(here(), "plots", "MOFA", paste0("factor", f, "_heatmap.png"))
    png(png_path, width=10, height=12, units="in", res=600)
    ComplexHeatmap::draw(ht, merge_legend=TRUE,
                         column_title=paste0("Factor ", f),
                         column_title_gp=grid::gpar(fontsize=14, fontface="bold"))
    dev.off()
    cat("Saved:", png_path, "\n")
  }
}
```


```{r}
ht = mofa_feature_heatmap_zscore(mofa_trained, 
                                 views = c("Proteomics",
                                           "Phosphoproteomics",
                                           "Redox"),
                                 factor_num=3, n=8, scale=3)

if (!is.null(ht)) {
    png_path = file.path(here(), "plots", "MOFA", "custom_panels", paste0("factor", 3, "_prot_heatmap.png"))
    png(png_path, width=10, height=12, units="in", res=600)
    ComplexHeatmap::draw(ht, merge_legend=TRUE,
                         column_title=paste0("Factor ", 3),
                         column_title_gp=grid::gpar(fontsize=14, fontface="bold"))
    dev.off()
    cat("Saved:", png_path, "\n")
  }

```

```{r}
phospho_mofa_to_da = function(mofa_features) {
  m = site_map %>%
    dplyr::filter(gene_site %in% mofa_features)
  setNames(m$feature_id, m$gene_site)
}

redox_mofa_to_da = function(mofa_features) {
  m = redox_feature_map %>%
    dplyr::filter(gene_site %in% mofa_features)
  setNames(m$feature_id, m$gene_site)
}

# deep dive specifically on Factor 2 redox
ht = mofa_feature_heatmap_da(mofa_trained,
                             factor_num=2,
                             view_da=list(Redox = REDOX_GN_NORM_DA[["trained_vs_SED"]]),
                             view_gene_map=list(Redox = redox_mofa_to_da),
                             n=20,
                             scale=3)
if (!is.null(ht)) {
  png_path = file.path(here(), "plots", "MOFA", "custom_panels", paste0("factor", 2, "_deepdive_heatmap.png"))
  png(png_path, width=10, height=10, units="in", res=600)
  ComplexHeatmap::draw(ht, merge_legend=TRUE,
                       column_title=paste0("Factor ", 2),
                       column_title_gp=grid::gpar(fontsize=14, fontface="bold"))
  dev.off()
  cat("Saved:", png_path, "\n")
}
```

