---
title: "MOFA Exploration"
output: html_document
date: "2026-03-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MOFA2)
library(Biobase)
library(dplyr)
library(edgeR)
library(tidyr)
library(here)
library(MotrpacRatTraining6moMuscleData)
source(file.path(here(), "Other_Figures", "MOFA", "mofa_helpers.R"))

mofa_out = file.path(here(), "Other_Figures", "MOFA", "saved_outputs")
```

# Overview

Multi-Omics Factor Analysis (MOFA) is applied to three gastrocnemius (GN)
-omic layers: transcriptomics, global proteomics, and
protein-normalized phosphoproteomics.

# Prepare each ome

Each ome is reformatted to a features-by-samples matrix with human gene
symbols (or human UniProt-site IDs for phosphoproteomics) as row names.
Features with any missing values are removed, and where multiple features map
to the same symbol, the first is kept (or highest-variance for phosphosites, to 
meet the recommendations of the MOFA tool in picking higher variance features.).

**Caveat -- complete-case bias:** MOFA requires complete data across all
samples for each feature. Proteomics and phosphoproteomics contain missing
values, so we restrict to complete cases before selecting highly variable
features. This may bias the feature set toward higher-abundance proteins or
phosphosites that are detected in all samples, and enrichment results should
be interpreted with this in mind.


## Transcriptomics

Log-CPM values are computed from raw counts. Ensembl IDs are mapped to human
gene symbols via `RAT_TO_HUMAN`. RNA-seq has no missing values.

```{r}
describe_eset(TRNSCRPT_GN)

gene_symbol_map = RAT_TO_HUMAN %>% select(ensembl, human_symbol)
trnscrpt_mat = cpm(exprs(TRNSCRPT_GN), log = TRUE) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("ensembl") %>%
  left_join(gene_symbol_map, by = "ensembl") %>%
  group_by(human_symbol) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  filter(!is.na(human_symbol)) %>%
  tibble::column_to_rownames("human_symbol") %>%
  select(-"ensembl") %>%
  as.matrix()

cat("Complete features:", nrow(trnscrpt_mat), "\n")
```

## Proteomics

Normalized log2 abundance. Protein IDs are mapped to gene symbols via
`fData(PROT_GN)`. Rows with any missing values are dropped.

```{r}
describe_eset(PROT_GN)

gene_symbol_map = fData(PROT_GN) %>% select(protein_id, gene_symbol)

prot_mat = exprs(PROT_GN) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("protein_id") %>%
  left_join(gene_symbol_map, by = "protein_id") %>%
  group_by(gene_symbol) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  filter(!is.na(gene_symbol)) %>%
  tibble::column_to_rownames("gene_symbol") %>%
  select(-"protein_id") %>%
  as.matrix()
prot_mat = prot_mat[complete.cases(prot_mat), ]

cat("Complete features:", nrow(prot_mat), "\n")
```

## Phosphoproteomics (Protein-Normalized)

Phosphoproteomics abundance adjusted for global protein-level changes
(`PHOSPHO_GN_NORM`). Rat phosphosite IDs are mapped to human UniProt-site
format (e.g., `Q9Y6K9-S473`) via `RAT_TO_HUMAN_SITE` so that feature names
match `phospho_GN_NORM_kinase_sets`. Multi-site peptides are split into
individual sites; when multiple peptides map to the same site, the row with
the highest variance is kept.

```{r}
describe_eset(PHOSPHO_GN_NORM)

# Map rat phospho features to human UniProt-site format (matches kinase_sets)
site_map = fData(PHOSPHO_GN_NORM) %>%
  tibble::rownames_to_column("feature_id") %>%
  select(feature_id) %>%
  left_join(RAT_TO_HUMAN_SITE,
            by = c("feature_id" = "ptm_id_rat_refseq")) %>%
  rename(human_feature = ptm_id_human_uniprot) %>%
  mutate(
    is_human = !is.na(human_feature),
    human_feature = ifelse(is_human, human_feature, feature_id),
    human_uniprot = ifelse(is_human,
                           sub("_.*", "", human_feature),
                           sub("([A-Z]P_[^_]+).*", "\\1", human_feature)),
    human_site = sub(".*_", "", human_feature),
    human_site = strsplit(human_site, split = ";")
  ) %>%
  unnest(cols = human_site) %>%
  mutate(featureID = paste0(human_uniprot, "-", human_site))

phospho_mat = exprs(PHOSPHO_GN_NORM) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("feature_id") %>%
  left_join(site_map %>% select(feature_id, featureID), by = "feature_id") %>%
  select(-feature_id) %>%
  # For duplicate sites, keep the row with highest variance
  mutate(row_var = apply(select(., -featureID), 1, var, na.rm = TRUE)) %>%
  arrange(featureID, -row_var) %>%
  filter(!duplicated(featureID)) %>%
  select(-row_var) %>%
  tibble::column_to_rownames("featureID") %>%
  as.matrix()
phospho_mat = phospho_mat[complete.cases(phospho_mat), ]

cat("Complete features:", nrow(phospho_mat), "\n")
```

# Select top variance features

The ome with the fewest complete features sets a common feature count. The
top-variance features from each ome are selected to match this count, so that
MOFA receives balanced views.

```{r}
n_features = min(nrow(trnscrpt_mat), nrow(prot_mat), nrow(phospho_mat))
cat("Target feature count (set by smallest ome):", n_features, "\n")
```

```{r}
trnscrpt_hvf = select_top_var(trnscrpt_mat, n_features)
prot_hvf = select_top_var(prot_mat, n_features)
phospho_hvf = select_top_var(phospho_mat, n_features)

cat("Transcriptomics HVF:", nrow(trnscrpt_hvf), "\n")
cat("Proteomics HVF:", nrow(prot_hvf), "\n")
cat("Phospho HVF:", nrow(phospho_hvf), "\n")
```

# Match samples across omes by pid

Each ome uses different viallabels for the same animal. We remap column names
from viallabel to pid, then intersect to keep only samples present in all
three omes. Transcriptomics has 5 samples per sex/timepoint while proteomics
has 6, so this step also reconciles the unequal sample sizes and removes any
outlier samples that were dropped from individual omes.

```{r}
trnscrpt_hvf = remap_to_pid(trnscrpt_hvf, TRNSCRPT_GN)
prot_hvf = remap_to_pid(prot_hvf, PROT_GN)
phospho_hvf = remap_to_pid(phospho_hvf, PHOSPHO_GN_NORM)

shared_pids = Reduce(intersect, list(
  colnames(trnscrpt_hvf),
  colnames(prot_hvf),
  colnames(phospho_hvf)
))
cat("Shared samples:", length(shared_pids), "\n")

trnscrpt_hvf = trnscrpt_hvf[, shared_pids]
prot_hvf = prot_hvf[, shared_pids]
phospho_hvf = phospho_hvf[, shared_pids]
```

# Build MOFA object

```{r}
mofa_data = list(
  Transcriptomics = trnscrpt_hvf,
  Proteomics = prot_hvf,
  Phosphoproteomics = phospho_hvf
)

mofa_obj = create_mofa(mofa_data)
mofa_obj
```

```{r}
plot_data_overview(mofa_obj)
```

# Add sample metadata

```{r}
# Build metadata from one of the esets (all share the same pid->sex/timepoint mapping)
pd = pData(TRNSCRPT_GN) %>%
  mutate(pid = as.character(pid)) %>%
  filter(pid %in% shared_pids) %>%
  select(sample = pid, sex, timepoint) %>%
  distinct()

samples_metadata(mofa_obj) = pd
```

# Set MOFA options and train

Views are scaled (`scale_views = TRUE`) so that omes with larger overall
variance do not dominate factor inference.

```{r}
data_opts = MOFA2::get_default_data_options(mofa_obj)
data_opts$scale_views = TRUE

model_opts = MOFA2::get_default_model_options(mofa_obj)
model_opts$num_factors = 10

train_opts = MOFA2::get_default_training_options(mofa_obj)

mofa_obj = MOFA2::prepare_mofa(
  mofa_obj,
  data_options = data_opts,
  model_options = model_opts,
  training_options = train_opts
)
```

```{r, results='hide', message=FALSE}
mofa_trained = MOFA2::run_mofa(mofa_obj,
                               outfile = file.path(here(), "Other_Figures", "MOFA", "mofa_trained.out"),
                               use_basilisk = TRUE)
```

# Downstream exploration

## Variance explained

Total and per-factor variance explained by each view.

```{r}
plot_variance_explained(mofa_trained, plot_total = TRUE)
```

```{r}
plot_variance_explained(mofa_trained, max_r2 = 15)
```

## Factor plots

Visualize sample loadings on the first four factors, colored by experimental
variables.

```{r, fig.width=8, fig.height=6}
plot_factor(mofa_trained, factors = 1:4, color_by = "timepoint", shape_by = "sex")
```

```{r, fig.width=8, fig.height=6}
plot_factor(mofa_trained, factors = 1:4, color_by = "sex", shape_by = "timepoint")
```

Factor 2 appears to separate well by sex.

## Factor correlations

```{r}
plot_factor_cor(mofa_trained)
```

## Top weights

Top-weighted features per factor for each view.

```{r, fig.width=8, fig.height=6}
plot_weights(mofa_trained, view = "Transcriptomics", factors = 1:3, nfeatures = 5)
```

```{r, fig.width=8, fig.height=6}
plot_weights(mofa_trained, view = "Proteomics", factors = 1:3, nfeatures = 5)
```

```{r, fig.width=8, fig.height=6}
plot_weights(mofa_trained, view = "Phosphoproteomics", factors = 1:3, nfeatures = 5)
```

## Enrichment: Transcriptomics

Gene-set enrichment (GO + skeletal muscle gene sets from `trnscrpt_GN_gene_sets`)
tested against MOFA factor weights using a parametric test. Feature names are
uppercased to match the gene-set format. Both positive- and negative-weight
enrichments are tested for factors 1--10.

```{r}
names = toupper(features_names(mofa_trained)[["Transcriptomics"]])
features_names(mofa_trained)[["Transcriptomics"]] = gsub("_TRANSCRIPTOMICS", "", names)

enrichment_trnscrpt = run_mofa_enrichment(mofa_trained,
  view = "Transcriptomics",
  feature_sets = trnscrpt_GN_gene_sets,
  factors = 1:10,
  csv_path = file.path(mofa_out, "enrichment_transcriptomics.csv")
)
```

```{r}
plot_enrichment_detailed(enrichment_trnscrpt$down,
  factor = 1,
  max.genes = 8,
  max.pathways = 5
)
```

## Enrichment: Proteomics

Gene-set enrichment using `prot_GN_gene_sets`.

```{r}
names = toupper(features_names(mofa_trained)[["Proteomics"]])
features_names(mofa_trained)[["Proteomics"]] = gsub("_PROTEOMICS", "", names)

enrichment_prot = run_mofa_enrichment(mofa_trained,
  view = "Proteomics",
  feature_sets = prot_GN_gene_sets,
  factors = 1:10,
  csv_path = file.path(mofa_out, "enrichment_proteomics.csv")
)
```

## Enrichment: Phosphoproteomics

Kinase-substrate enrichment using `phospho_GN_NORM_kinase_sets`
(PhosphoSitePlus). Feature names are in human UniProt-site format to match the
kinase set definitions.

```{r}
features_names(mofa_trained)[["Phosphoproteomics"]] = toupper(features_names(mofa_trained)[["Phosphoproteomics"]])

enrichment_phospho = run_mofa_enrichment(mofa_trained,
  view = "Phosphoproteomics",
  feature_sets = phospho_GN_NORM_kinase_sets,
  factors = 1:10,
  csv_path = file.path(mofa_out, "enrichment_phosphoproteomics.csv")
)
```
