---
title: "MOFA Exploration"
output: html_document
date: "2026-03-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MOFA2)
library(Biobase)
library(dplyr)
library(edgeR)
library(tidyr)
library(ggplot2)
library(here)
library(MotrpacRatTraining6moMuscleData)
source(file.path(here(), "Other_Figures", "MOFA", "mofa_helpers.R"))

mofa_out = file.path(here(), "Other_Figures", "MOFA", "saved_outputs")
```

# Overview

Multi-Omics Factor Analysis (MOFA) omic layers: transcriptomics, global proteomics,
(protein-normalized) phosphoproteomics, (protein-normalized) redox proteomics, and
metabolomics.

# Prepare each ome

Features with any missing values are removed, and where multiple features map
to the same symbol, the first is kept (or highest-variance for phosphosites, to 
meet the recommendations of the MOFA tool in picking higher variance features).

**Note: possible complete-case bias:** MOFA input is currently complete data across all
samples for each feature. Proteomics and phosphoproteomics contain missing
values, so we restrict to complete cases before selecting highly variable
features. This may bias the feature set toward higher-abundance proteins or
phosphosites that are detected in all samples, and enrichment results should
be interpreted with this in mind.


## Transcriptomics

Log-CPM values are computed from raw counts. Ensembl IDs are mapped to human
gene symbols via `RAT_TO_HUMAN`. 
```{r}
describe_eset(TRNSCRPT_GN)

gene_symbol_map = RAT_TO_HUMAN %>% select(ensembl, human_symbol)
trnscrpt_mat = cpm(exprs(TRNSCRPT_GN), log = TRUE) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("ensembl") %>%
  left_join(gene_symbol_map, by = "ensembl") %>%
  group_by(human_symbol) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  filter(!is.na(human_symbol)) %>%
  tibble::column_to_rownames("human_symbol") %>%
  select(-"ensembl") %>%
  as.matrix()

cat("Complete features:", nrow(trnscrpt_mat), "\n")
```

## Proteomics

Normalized log2 abundance. Rows with any missing values are dropped.

```{r}
describe_eset(PROT_GN)

gene_symbol_map = fData(PROT_GN) %>% select(protein_id, gene_symbol)

prot_mat = exprs(PROT_GN) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("protein_id") %>%
  left_join(gene_symbol_map, by = "protein_id") %>%
  group_by(gene_symbol) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  filter(!is.na(gene_symbol)) %>%
  tibble::column_to_rownames("gene_symbol") %>%
  select(-"protein_id") %>%
  as.matrix()
prot_mat = prot_mat[complete.cases(prot_mat), ]

cat("Complete features:", nrow(prot_mat), "\n")
```

## Phosphoproteomics (Protein-Normalized)

See vignette for KSEA gene set testing for mapping info. 

```{r}
describe_eset(PHOSPHO_GN_NORM)

# Map rat phospho features to human UniProt-site format (matches kinase_sets)
site_map = fData(PHOSPHO_GN_NORM) %>%
  tibble::rownames_to_column("feature_id") %>%
  select(feature_id) %>%
  left_join(RAT_TO_HUMAN_SITE,
            by = c("feature_id" = "ptm_id_rat_refseq")) %>%
  rename(human_feature = ptm_id_human_uniprot) %>%
  mutate(
    is_human = !is.na(human_feature),
    human_feature = ifelse(is_human, human_feature, feature_id),
    human_uniprot = ifelse(is_human,
                           sub("_.*", "", human_feature),
                           sub("([A-Z]P_[^_]+).*", "\\1", human_feature)),
    human_site = sub(".*_", "", human_feature),
    human_site = strsplit(human_site, split = ";")
  ) %>%
  unnest(cols = human_site) %>%
  mutate(featureID = paste0(human_uniprot, "-", human_site))

phospho_mat = exprs(PHOSPHO_GN_NORM) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("feature_id") %>%
  left_join(site_map %>% select(feature_id, featureID), by = "feature_id") %>%
  select(-feature_id) %>%
  # For duplicate sites, keep the row with highest variance
  mutate(row_var = apply(select(., -featureID), 1, var, na.rm = TRUE)) %>%
  arrange(featureID, -row_var) %>%
  filter(!duplicated(featureID)) %>%
  select(-row_var) %>%
  tibble::column_to_rownames("featureID") %>%
  as.matrix()
phospho_mat = phospho_mat[complete.cases(phospho_mat), ]

cat("Complete features:", nrow(phospho_mat), "\n")
```

## Redox Proteomics (Protein-Normalized)

Log2 abundance adjusted for global protein changes. Features mapped to rat gene symbols;
duplicate sites per gene are resolved by highest variance.

```{r}
describe_eset(REDOX_GN_NORM)

redox_feature_map = fData(REDOX_GN_NORM) %>%
  tibble::rownames_to_column("feature_id") %>%
  select(feature_id, gene_symbol)

redox_mat = exprs(REDOX_GN_NORM) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("feature_id") %>%
  left_join(redox_feature_map, by = "feature_id") %>%
  select(-feature_id) %>%
  filter(!is.na(gene_symbol)) %>%
  mutate(row_var = apply(select(., -gene_symbol), 1, var, na.rm = TRUE)) %>%
  arrange(gene_symbol, -row_var) %>%
  filter(!duplicated(gene_symbol)) %>%
  select(-row_var) %>%
  tibble::column_to_rownames("gene_symbol") %>%
  as.matrix()
redox_mat = redox_mat[complete.cases(redox_mat), ]

cat("Complete features:", nrow(redox_mat), "\n")
```

## Metabolomics

Normalized log2 abundance. Features are mapped to RefMet standardized names; where a
metabolite appears in multiple datasets the highest-variance row is retained.

```{r}
describe_eset(METAB_GN)

metab_feature_map = fData(METAB_GN) %>%
  tibble::rownames_to_column("feature_id") %>%
  select(feature_id, refmet_name = metabolite_refmet)

metab_mat = exprs(METAB_GN) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("feature_id") %>%
  left_join(metab_feature_map, by = "feature_id") %>%
  select(-feature_id) %>%
  filter(!is.na(refmet_name)) %>%
  mutate(row_var = apply(select(., -refmet_name), 1, var, na.rm = TRUE)) %>%
  arrange(refmet_name, -row_var) %>%
  filter(!duplicated(refmet_name)) %>%
  select(-row_var) %>%
  tibble::column_to_rownames("refmet_name") %>%
  as.matrix()
metab_mat = metab_mat[complete.cases(metab_mat), ]

cat("Complete features:", nrow(metab_mat), "\n")
```

# Select top variance features

Filter to same number of features per ome

```{r}
n_features = min(nrow(trnscrpt_mat), nrow(prot_mat), nrow(phospho_mat),
                 nrow(redox_mat), nrow(metab_mat))
cat("Target feature count (set by smallest ome):", n_features, "\n")
```

```{r}
trnscrpt_hvf = select_top_var(trnscrpt_mat, n_features)
prot_hvf = select_top_var(prot_mat, n_features)
phospho_hvf = select_top_var(phospho_mat, n_features)
redox_hvf = select_top_var(redox_mat, n_features)
metab_hvf = select_top_var(metab_mat, n_features)

cat("Transcriptomics HVF:", nrow(trnscrpt_hvf), "\n")
cat("Proteomics HVF:", nrow(prot_hvf), "\n")
cat("Phospho HVF:", nrow(phospho_hvf), "\n")
cat("Redox HVF:", nrow(redox_hvf), "\n")
cat("Metabolomics HVF:", nrow(metab_hvf), "\n")
```

# Match samples across omes by pid

Filter to the shared set of animals across all omes (transcriptomics has 5M+5F/timepoint,
proteomics/phospho/redox have 6+6; metabolomics has its own sample set).

```{r}
trnscrpt_hvf = remap_to_pid(trnscrpt_hvf, TRNSCRPT_GN)
prot_hvf = remap_to_pid(prot_hvf, PROT_GN)
phospho_hvf = remap_to_pid(phospho_hvf, PHOSPHO_GN_NORM)
redox_hvf = remap_to_pid(redox_hvf, REDOX_GN_NORM)
# METAB_GN columns are already pid values; no remap needed

shared_pids = Reduce(intersect, list(
  colnames(trnscrpt_hvf),
  colnames(prot_hvf),
  colnames(phospho_hvf),
  colnames(redox_hvf),
  colnames(metab_hvf)
))
cat("Shared samples:", length(shared_pids), "\n")

trnscrpt_hvf = trnscrpt_hvf[, shared_pids]
prot_hvf = prot_hvf[, shared_pids]
phospho_hvf = phospho_hvf[, shared_pids]
redox_hvf = redox_hvf[, shared_pids]
metab_hvf = metab_hvf[, shared_pids]
```

# Build MOFA object

```{r}
mofa_data = list(
  Transcriptomics = trnscrpt_hvf,
  Proteomics = prot_hvf,
  Phosphoproteomics = phospho_hvf,
  Redox = redox_hvf,
  Metabolomics = metab_hvf
)

mofa_obj = create_mofa(mofa_data)
mofa_obj
```

```{r}
plot_data_overview(mofa_obj)
```

# Add sample metadata

```{r}
# Build metadata from transcriptomics pData (all omes share pid->sex/timepoint mapping)
pd = pData(TRNSCRPT_GN) %>%
  mutate(pid = as.character(pid)) %>%
  filter(pid %in% shared_pids) %>%
  select(sample = pid, sex, timepoint) %>%
  distinct()

samples_metadata(mofa_obj) = pd
```

# Set MOFA options and train

```{r}
data_opts = MOFA2::get_default_data_options(mofa_obj)
data_opts$scale_views = TRUE

model_opts = MOFA2::get_default_model_options(mofa_obj)
model_opts$num_factors = 10

train_opts = MOFA2::get_default_training_options(mofa_obj)

mofa_obj = MOFA2::prepare_mofa(
  mofa_obj,
  data_options = data_opts,
  model_options = model_opts,
  training_options = train_opts
)
```

```{r, results='hide', message=FALSE}
# mofa_trained = MOFA2::run_mofa(mofa_obj,
#                                outfile = file.path(here(), "Other_Figures", "MOFA", "mofa_trained.out"),
#                                use_basilisk = TRUE)

mofa_trained = load_model(
  file.path(here(), "Other_Figures", "MOFA", "mofa_trained.out")
)
```

# Downstream exploration

## Variance explained

Total and per-factor variance explained by each view.

```{r}
plot_variance_explained(mofa_trained, plot_total = TRUE)
```

## Factor plots

Visualize sample loadings on the first four factors, colored by experimental
variables.

```{r, fig.width=8, fig.height=6}
plot_factor(mofa_trained, factors = 1:4, color_by = "timepoint", shape_by = "sex")
```

```{r, fig.width=8, fig.height=6}
plot_factor(mofa_trained, factors = 1:4, color_by = "sex", shape_by = "timepoint")
```

Factor 2 appears to separate well by sex.

## Factor correlations

```{r}
plot_factor_cor(mofa_trained)
```

## Feature weight plots


```{r feature_weights, fig.width=12, fig.height=10}
# source(file.path(here(), "Other_Figures", "MOFA", "mofa_helpers.R"))

plot_feature_weights(mofa_trained, factors = 2, n_features = 12)
```

## Enrichment: Transcriptomics

```{r}
names = toupper(features_names(mofa_trained)[["Transcriptomics"]])
features_names(mofa_trained)[["Transcriptomics"]] = gsub("_TRANSCRIPTOMICS", "", names)

enrichment_trnscrpt = run_mofa_enrichment(mofa_trained,
                                          view = "Transcriptomics",
                                          feature_sets = trnscrpt_GN_gene_sets,
                                          factors = 1:10,
                                          csv_path = file.path(mofa_out, "enrichment_transcriptomics.csv")
)
```

```{r}
plot_enrichment_detailed(enrichment_trnscrpt$down,
                         factor = 1,
                         max.genes = 8,
                         max.pathways = 5
)
```

## Enrichment: Proteomics

```{r}
names = toupper(features_names(mofa_trained)[["Proteomics"]])
features_names(mofa_trained)[["Proteomics"]] = gsub("_PROTEOMICS", "", names)

enrichment_prot = run_mofa_enrichment(mofa_trained,
                                      view = "Proteomics",
                                      feature_sets = prot_GN_gene_sets,
                                      factors = 1:10,
                                      csv_path = file.path(mofa_out, "enrichment_proteomics.csv")
)
```

## Enrichment: Phosphoproteomics

```{r, eval = T}
features_names(mofa_trained)[["Phosphoproteomics"]] = toupper(features_names(mofa_trained)[["Phosphoproteomics"]])

enrichment_phospho = run_mofa_enrichment(mofa_trained,
                                         view = "Phosphoproteomics",
                                         feature_sets = phospho_GN_NORM_kinase_sets,
                                         factors = 1:10,
                                         csv_path = file.path(mofa_out, "enrichment_phosphoproteomics.csv")
)
```

## Enrichment: Redox Proteomics

Feature names are rat gene symbols (touppered to match gene sets).

```{r}
features_names(mofa_trained)[["Redox"]] = toupper(features_names(mofa_trained)[["Redox"]])

enrichment_redox = run_mofa_enrichment(mofa_trained,
                                       view = "Redox",
                                       feature_sets = redox_GN_NORM_gene_sets,
                                       factors = 1:10,
                                       csv_path = file.path(mofa_out, "enrichment_redox.csv")
)
```

## Enrichment: Metabolomics

Feature names are RefMet standardized names (touppered to match refmet sets).

```{r}
features_names(mofa_trained)[["Metabolomics"]] = toupper(features_names(mofa_trained)[["Metabolomics"]])

enrichment_metab = run_mofa_enrichment(mofa_trained,
                                       view = "Metabolomics",
                                       feature_sets = metab_GN_refmet_sets,
                                       factors = 1:10,
                                       csv_path = file.path(mofa_out, "enrichment_metabolomics.csv")
)
```

## Cross-ome enrichment: faceted per factor

Show the top enriched pathways/sets from every ome for a single factor in one panel.
Edit `factor_to_plot` to explore other factors.

```{r, fig.width=9, fig.height=14}
all_enrichments = list(
  Transcriptomics = enrichment_trnscrpt,
  Proteomics = enrichment_prot,
  # Phosphoproteomics = enrichment_phospho,
  Redox = enrichment_redox,
  Metabolomics = enrichment_metab
)
```


```{r}
sig_all = load_sig_enrichments(mofa_out)
```

```{r}
for (f in 1:10) {
  p = plot_enrichment_faceted(all_enrichments,
                              factor = f,
                              max_pathways = 8,
                              png_path = file.path(here(), "plots", "MOFA", paste0("factor", f, ".png")))
  print(p)
}
```

## Combined enrichment bubble heatmap

Bubble heatmap across all views and factors.  Each row is a top significant
gene set; columns are factors.  Circle size = -log10(padj) (capped at 5 mm),
circle color = signed -log10(padj) (red = up-weighted, blue = down-weighted),
grey background = padj < 0.05.

```{r mofa_enrichment_bubble, message=FALSE}
# Compute global signed -log10(padj) range for a shared colour scale
compute_signed_logp = function(enrich, n_top=8, padj_threshold=0.05) {
  factor_cols = intersect(colnames(enrich$up$pval.adj),
                          colnames(enrich$down$pval.adj))
  pu = enrich$up$pval.adj[, factor_cols, drop=FALSE]
  pd = enrich$down$pval.adj[, factor_cols, drop=FALSE]
  min_p = pmin(apply(pu, 1, min, na.rm=TRUE), apply(pd, 1, min, na.rm=TRUE))
  sig = names(sort(min_p[min_p < padj_threshold]))[seq_len(min(n_top, sum(min_p < padj_threshold)))]
  if (length(sig) == 0) return(numeric(0))
  as.vector(ifelse(pu[sig, , drop=FALSE] <= pd[sig, , drop=FALSE],
                   -log10(pu[sig, , drop=FALSE]),
                   log10(pd[sig, , drop=FALSE])))
}

all_vals = unlist(lapply(all_enrichments, compute_signed_logp))
max_abs_logp = min(max(abs(all_vals), na.rm=TRUE), 20)

col_fun_enrich = circlize::colorRamp2(
  c(-max_abs_logp, 0, max_abs_logp),
  c("#3366ff", "white", "darkred")
)

ome_cols_mofa = c(
  Transcriptomics="#4575b4",
  Proteomics="#d73027",
  Redox="#74add1",
  Metabolomics="#1a9850"
)

color_lg = ComplexHeatmap::Legend(
  title="-log10(padj)\n(+ up, \u2212 down)",
  col_fun=col_fun_enrich,
  at=c(-round(max_abs_logp), 0, round(max_abs_logp)),
  legend_height=grid::unit(4, "cm")
)
ome_lg = ComplexHeatmap::Legend(
  title="View",
  at=names(ome_cols_mofa),
  legend_gp=grid::gpar(fill=ome_cols_mofa)
)
sig_lg = ComplexHeatmap::Legend(
  title="Significance",
  at="padj < 0.05",
  legend_gp=grid::gpar(fill="grey90", col=NA)
)

enrich_heatmaps = list()
for (v in names(all_enrichments)) {
  ht = make_mofa_enrichment_heatmap(
    enrichment=all_enrichments[[v]],
    view_name=v,
    ome_cols=ome_cols_mofa,
    col_fun=col_fun_enrich,
    n_top=8
  )
  if (!is.null(ht)) enrich_heatmaps[[v]] = ht
}

if (length(enrich_heatmaps) > 0) {
  combined_enrich = Reduce(`%v%`, enrich_heatmaps)
  png_path = file.path(here(), "plots", "MOFA", "enrichment_bubble_heatmap.png")
  png(png_path, width=12, height=10, units="in", res=600)
  ComplexHeatmap::draw(
    combined_enrich,
    heatmap_legend_list=list(color_lg, ome_lg, sig_lg),
    merge_legend=TRUE,
    heatmap_legend_side="left"
  )
  dev.off()
  cat("Saved:", png_path, "\n")
}
```

## Feature heatmaps

Heatmaps of the top-weighted features per factor for three views whose feature
names map cleanly to the `gene_symbol` column of the differential-analysis
objects.  Transcriptomics features are human gene symbols in the MOFA object
and are reverse-mapped to rat gene symbols via `RAT_TO_HUMAN` before looking
up DA results.  Phosphoproteomics and Metabolomics are omitted here because
their feature identifiers (UniProt-site and RefMet names, respectively) do not
align directly with the `gene_symbol` column expected by `feature_heatmap`.

```{r feature_heatmaps, fig.width=10, fig.height=6, results='hide', message=FALSE}
library(ComplexHeatmap)

# Reverse-map MOFA transcriptomics human symbols → rat gene symbols
human_to_rat = function(human_features) {
  RAT_TO_HUMAN %>%
    dplyr::filter(toupper(human_symbol) %in% toupper(human_features)) %>%
    dplyr::pull(symbol)
}

view_da = list(
  Transcriptomics = TRNSCRPT_GN_DA[["trained_vs_SED"]],
  Proteomics = PROT_GN_DA[["trained_vs_SED"]],
  Phosphoproteomics = PHOSPHO_GN_DA[["trained_vs_SED"]],
  Redox = REDOX_GN_NORM_DA[["trained_vs_SED"]],
  Metabolomics = METAB_GN_DA[["trained_vs_SED"]]
)

phospho_mofa_to_da = function(mofa_features) {
  site_map %>%
    dplyr::filter(featureID %in% mofa_features) %>%
    dplyr::pull(feature_id)
}

view_gene_map = list(
  Transcriptomics = human_to_rat,
  Proteomics = NULL,
  Phosphoproteomics = phospho_mofa_to_da,
  Redox = NULL,
  Metabolomics = NULL
)

for (f in 2:3) {
  ht = mofa_feature_heatmap(mofa_trained, 
                            factor_num=f,
                            view_da=view_da, 
                            view_gene_map=view_gene_map,
                            n=15, 
                            scale=2,
                            view_feature_col=list(Phosphoproteomics="featureName",
                                                  Metabolomics="metabolite_refmet"))
  if (!is.null(ht)) {
    png_path = file.path(here(), "plots", "MOFA", paste0("factor", f, "_heatmap.png"))
    png(png_path, width=10, 
        height=12, 
        units="in", 
        res=600)
    ComplexHeatmap::draw(ht, merge_legend=TRUE,
                         column_title=paste0("Factor ", f),
                         column_title_gp=grid::gpar(fontsize=14, fontface="bold"))
    dev.off()
    cat("Saved:", png_path, "\n")
  }
}
```

