---
title: "Skeletal Muscle Differential Analysis of -Omics Datasets"
author: "Tyler Sagendorf"
date: "Last compiled `r format(Sys.time(), '%d %B, %Y')`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Skeletal Muscle Differential Analysis of -Omics Datasets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  fig.width = 6, 
  fig.height = 4,
  eval = FALSE
)
```

```{r setup, message=FALSE, warning=FALSE}
# Required packages
library(MotrpacRatTraining6moMuscleData)
library(limmaTesting) # limmaFit, limmaDEA
library(Biobase)
library(dplyr)
library(ggplot2)
```

We will test two types of contrasts: sex-matched trained vs. sedentary comparisons and male vs. female comparisons at SED and 8W.

```{r}
# Sex-matched trained vs. sedentary comparisons
timepoint <- rep(c("1W", "2W", "4W", "8W"), times = 2)
sex <- rep(c("F", "M"), each = 4)
contrasts <- sprintf("exp_group%s_%s - exp_group%s_SED", 
                     sex, timepoint, sex)

# Include sex comparisons
contrasts <- c(contrasts, 
               "exp_groupM_SED - exp_groupF_SED", 
               "exp_groupM_8W - exp_groupF_8W")

contrasts
```

```{r}
# Split DA results by contrast type (trained vs. SED, male vs. female) and
# adjust p-values.
split_by_type <- function(x) {
  x %>% 
    as.data.frame() %>% 
    filter(!is.na(logFC)) %>% 
    mutate(
      # Simplify contrast names
      contrast = gsub("exp_group", "", contrast),
      # Z-score
      z = zscoreT(t, df = df.total, method = "hill")
    ) %>%
    # Create indicator column
    mutate(group = ifelse(grepl("M.*F", contrast), 
                          "MvF", "trained_vs_SED"),
           # Convert to factor to preserve order when splitting
           group = factor(group, levels = c("trained_vs_SED", "MvF"))) %>% 
    relocate(z, .after = df.total) %>% 
    # Split results into a list of 2 tables
    split(f = .[["group"]]) %>%
    # For each table, adjust p-values across related contrasts
    lapply(function(xi) {
      xi %>%
        droplevels.data.frame() %>% 
        # Adjust p-values across sets of related contrasts
        mutate(adj.P.Val = p.adjust(P.Value, method = "BH"),
               contrast = factor(contrast, levels = unique(contrast))) %>%
        # Limit columns in output
        select(any_of(c("contrast", "featureName", 
                        "metabolite_refmet", "standardized_name",
                        "dataset", "is_targeted", "sub_class",
                        "description", "gene_symbol", 
                        "entrez_id", "uniprot_gn_id", "rgd_id",
                        "logFC", "AveExpr", "t", "df.total", 
                        "z", "P.Value", "adj.P.Val"))) %>% 
        # Reduce digits in numeric columns (excluding z)
        mutate(t = round(t, digits = 3),
               across(.cols = c(AveExpr, logFC, df.total),
                      .fns = ~ round(.x, digits = 2)),
               across(.cols = c(P.Value, adj.P.Val),
                      .fns = ~ signif(.x, digits = 2)))
    })
}
```


# Gastrocnemius

## Global Proteomics

```{r}
# Create MArrayLM fitted object
PROT_GN_FIT <- limmaFit(object = PROT_GN,
                        model.str = "~ 0 + exp_group",
                        contrasts = contrasts, 
                        # Robust mean-variance trend
                        trend = TRUE, robust = TRUE, 
                        # Sample-level quality weights
                        var.group = "viallabel",
                        plot = TRUE)
```

```{r}
# Differential analysis
PROT_GN_DA <- limmaDEA(fit = PROT_GN_FIT) %>%
  split_by_type()
```

```{r}
# Check p-value histograms
lapply(PROT_GN_DA, function(x) {
  ggplot(x, aes(x = P.Value)) +
    geom_histogram(breaks = seq(0, 1, 0.05)) +
    facet_wrap(~ contrast, ncol = 4) +
    theme_bw()
})
```

```{r}
# Count number of differential proteins
lapply(PROT_GN_DA, function(x) with(x, table(contrast, adj.P.Val < 0.05)))
```


## Phosphoproteomics

```{r}
# Create MArrayLM fitted object
PHOSPHO_GN_FIT <- limmaFit(object = PHOSPHO_GN,
                           model.str = "~ 0 + exp_group",
                           contrasts = contrasts, 
                           # Robust mean-variance trend
                           trend = TRUE, robust = TRUE, 
                           # Sample-level quality weights
                           var.group = "viallabel",
                           plot = TRUE)
```

```{r}
# Differential analysis
PHOSPHO_GN_DA <- limmaDEA(fit = PHOSPHO_GN_FIT) %>%
  split_by_type()
```

```{r}
# Check p-value histograms
lapply(PHOSPHO_GN_DA, function(x) {
  ggplot(x, aes(x = P.Value)) +
    geom_histogram(breaks = seq(0, 1, 0.05)) +
    facet_wrap(~ contrast, ncol = 4) +
    theme_bw()
})
```

```{r}
# Count number of differential phosphosites
lapply(PHOSPHO_GN_DA, function(x) with(x, table(contrast, adj.P.Val < 0.05)))
```


## Acetyl-proteomics

```{r}
# Create MArrayLM fitted object
ACETYL_GN_FIT <- limmaFit(object = ACETYL_GN,
                          model.str = "~ 0 + exp_group",
                          contrasts = contrasts, 
                          # Robust mean-variance trend
                          trend = TRUE, robust = TRUE, 
                          # Sample-level quality weights
                          var.group = "viallabel",
                          plot = TRUE)
```

```{r}
# Differential analysis
ACETYL_GN_DA <- limmaDEA(fit = ACETYL_GN_FIT) %>%
  split_by_type()
```

```{r}
# Check p-value histograms
lapply(ACETYL_GN_DA, function(x) {
  ggplot(x, aes(x = P.Value)) +
    geom_histogram(breaks = seq(0, 1, 0.05)) +
    facet_wrap(~ contrast, ncol = 4) +
    theme_bw()
})
```

```{r}
# Count number of differential phosphosites
lapply(ACETYL_GN_DA, function(x) with(x, table(contrast, adj.P.Val < 0.05)))
```


## Redox Proteomics

```{r}
# Create MArrayLM fitted object
REDOX_GN_FIT <- limmaFit(object = REDOX_GN,
                         model.str = "~ 0 + exp_group",
                         contrasts = contrasts, 
                         # Robust mean-variance trend
                         trend = TRUE, robust = TRUE, 
                         # Sample-level quality weights
                         var.group = "viallabel",
                         plot = TRUE)
```

```{r}
# Differential analysis
REDOX_GN_DA <- limmaDEA(fit = REDOX_GN_FIT) %>%
  split_by_type()
```

```{r}
# Check p-value histograms
lapply(REDOX_GN_DA, function(x) {
  ggplot(x, aes(x = P.Value)) +
    geom_histogram(breaks = seq(0, 1, 0.05)) +
    facet_wrap(~ contrast, ncol = 4) +
    theme_bw()
})
```

```{r}
# Count number of differential phosphosites
lapply(REDOX_GN_DA, function(x) with(x, table(contrast, adj.P.Val < 0.05)))
```

```{r}
# Include column for differential change in global protein abundance
f1 <- function(x) {
  mutate(x, protein_id = sub("([A-Z]P_[^_]+).*", "\\1", featureName)) %>%
    left_join(foo, by = c("contrast", "protein_id" = "featureName")) %>%
    select(-protein_id)
}

for (name_i in names(PROT_GN_DA)) {
  foo <- PROT_GN_DA[[name_i]] %>%
    mutate(signif_protein = adj.P.Val < 0.05) %>%
    select(contrast, featureName, signif_protein)

  PHOSPHO_GN_DA[[name_i]] <- f1(PHOSPHO_GN_DA[[name_i]])
  
  ACETYL_GN_DA[[name_i]] <- f1(ACETYL_GN_DA[[name_i]])
  
  REDOX_GN_DA[[name_i]] <- f1(REDOX_GN_DA[[name_i]])
}
```


## Transcriptomics

```{r}
# Model includes covariates
model.str <- paste("~ 0 + exp_group", "rin", "pct_globin",
                   "pct_umi_dup", "median_5_3_bias", sep = " + ")

TRNSCRPT_GN_FIT <- limmaFit(object = TRNSCRPT_GN, 
                            model.str = model.str,
                            contrasts = contrasts, 
                            # Robust prior variance
                            trend = FALSE, robust = TRUE,
                            # Sample-level quality weights
                            var.group = "viallabel",
                            plot = TRUE)
```

```{r}
# Differential analysis
TRNSCRPT_GN_DA <- limmaDEA(fit = TRNSCRPT_GN_FIT) %>%
  split_by_type()
```

```{r}
# Check p-value histograms
lapply(TRNSCRPT_GN_DA, function(x) {
  ggplot(x, aes(x = P.Value)) +
    geom_histogram(breaks = seq(0, 1, 0.05)) +
    facet_wrap(~ contrast, ncol = 4) +
    theme_bw()
})
```

```{r}
# Count number of differential transcripts
lapply(TRNSCRPT_GN_DA, function(x) with(x, table(contrast, adj.P.Val < 0.05)))
```


## Metabolomics

```{r}
METAB_GN_FIT <- limmaFit(object = METAB_GN, 
                         model.str = "~ 0 + exp_group",
                         contrasts = contrasts, 
                         # Robust mean-variance trend
                         trend = TRUE, robust = TRUE, 
                         # Sample-level quality weights
                         var.group = "pid",
                         plot = TRUE)
```

```{r}
# Differential analysis
METAB_GN_DA <- limmaDEA(fit = METAB_GN_FIT) %>%
  split_by_type()
```

```{r}
# Check p-value histograms
lapply(METAB_GN_DA, function(x) {
  ggplot(x, aes(x = P.Value)) +
    geom_histogram(breaks = seq(0, 1, 0.05)) +
    facet_wrap(~ contrast, ncol = 4) +
    theme_bw()
})
```

```{r}
# Count number of differential metabolites
lapply(METAB_GN_DA, function(x) with(x, table(contrast, adj.P.Val < 0.05)))
```


## ATAC-Seq

Differential analysis of ATAC-Seq data is performed in data-raw/ATAC_GN.R, so only a snippet is provided below. The analysis is identical to that of RNA-Seq, though the covariates are different. Only those features with adjusted p-values < 0.05 are provided, as there were 856,947 total features tested.

```{r eval=FALSE}
## Not run
model.str <- paste("~ 0 + exp_group", "Sample_batch", 
                   "peak_enrich.frac_reads_in_peaks.macs2.frip",
                   sep = " + ")

ATAC_GN_FIT <- limmaFit(object = ATAC_GN,
                        model.str = model.str,
                        contrasts = contrasts,
                        # Robust prior variance
                        trend = FALSE, robust = TRUE,
                        # Sample-level quality weights
                        var.group = "viallabel",
                        plot = TRUE)
```

```{r eval=FALSE}
lapply(ATAC_GN_DA_05FDR, 
       function(x) with(x, table(contrast, adj.P.Val < 0.05)))
```



# Vastus Lateralis

## Transcriptomics

```{r}
# Model includes covariates
model.str <- paste("~ 0 + exp_group", "rin", "pct_globin",
                   "pct_umi_dup", "median_5_3_bias", sep = " + ")

TRNSCRPT_VL_FIT <- limmaFit(object = TRNSCRPT_VL, 
                            model.str = model.str,
                            contrasts = contrasts, 
                            # Robust prior variance
                            trend = FALSE, robust = TRUE, 
                            # Sample-level quality weights
                            var.group = "viallabel",
                            plot = TRUE)
```

```{r}
# Differential analysis
TRNSCRPT_VL_DA <- limmaDEA(fit = TRNSCRPT_VL_FIT) %>%
  split_by_type()
```

```{r}
# Check p-value histograms
lapply(TRNSCRPT_VL_DA, function(x) {
  ggplot(x, aes(x = P.Value)) +
    geom_histogram(breaks = seq(0, 1, 0.05)) +
    facet_wrap(~ contrast, ncol = 4) +
    theme_bw()
})
```

```{r}
# Count number of differential transcripts
lapply(TRNSCRPT_VL_DA, function(x) with(x, table(contrast, adj.P.Val < 0.05)))
```


## Metabolomics

```{r}
METAB_VL_FIT <- limmaFit(object = METAB_VL, 
                         model.str = "~ 0 + exp_group",
                         contrasts = contrasts, 
                         # Robust mean-variance trend
                         trend = TRUE, robust = TRUE, 
                         # Sample-level quality weights
                         var.group = "pid",
                         plot = TRUE)
```

```{r}
# Differential analysis
METAB_VL_DA <- limmaDEA(fit = METAB_VL_FIT) %>%
  split_by_type()
```

```{r}
# Check p-value histograms
lapply(METAB_VL_DA, function(x) {
  ggplot(x, aes(x = P.Value)) +
    geom_histogram(breaks = seq(0, 1, 0.05)) +
    facet_wrap(~ contrast, ncol = 4) +
    theme_bw()
})
```

```{r}
# Count number of differential metabolites
lapply(METAB_VL_DA, function(x) with(x, table(contrast, adj.P.Val < 0.05)))
```


# Save Results

```{r eval=FALSE}
# Save
usethis::use_data(PROT_GN_DA, overwrite = TRUE, version = 3)
usethis::use_data(PHOSPHO_GN_DA, overwrite = TRUE, version = 3)
usethis::use_data(ACETYL_GN_DA, overwrite = TRUE, version = 3)
usethis::use_data(REDOX_GN_DA, overwrite = TRUE, version = 3)
usethis::use_data(TRNSCRPT_GN_DA, overwrite = TRUE, version = 3)
usethis::use_data(METAB_GN_DA, overwrite = TRUE, version = 3)

usethis::use_data(TRNSCRPT_VL_DA, overwrite = TRUE, version = 3)
usethis::use_data(METAB_VL_DA, overwrite = TRUE, version = 3)
```


# Session Info

```{r}
sessionInfo()
```
