---
title: "Fuzzy C-Means Clustering"
author: "Tyler Sagendorf and Christopher Jin"
date: "Last compiled `r format(Sys.time(), '%d %B, %Y')`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Fuzzy C-Means Clustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  fig.width = 6, 
  fig.height = 4
)

save_plots <- dir.exists(file.path("..", "plots"))
```
  
```{r setup, warning=FALSE, message=FALSE}
# Required packages
library(Biobase)
library(ComplexHeatmap) # max_text_width
library(dplyr)
library(ggplot2)
library(grid) # gpar, unit, convertUnit
library(latex2exp) # TeX
library(Mfuzz) # mfuzz, mestimate, mfuzz.plot2
library(MotrpacRatTraining6moMuscleData)
library(tibble)
library(tidyr)
library(TMSig)
```


# Prepare Input

```{r combine-DA-results}
combined_DA <- list(
  "PROT_GN" = PROT_GN_DA,
  "TRNSCRPT_GN" = TRNSCRPT_GN_DA,
  "METAB_GN" = METAB_GN_DA
) %>% 
  lapply(function(x) {
    x[["trained_vs_SED"]] %>% 
      select(contrast, featureName, z, P.Value, adj.P.Val,
             any_of(c("gene_symbol", "metabolite_refmet")))
  }) %>% 
  bind_rows(.id = "ome") %>% 
  mutate(
    contrast = sub(" .*", "", contrast),
    contrast = factor(contrast, levels = unique(contrast)),
    featureName2 = case_when(
      !is.na(gene_symbol) ~ gene_symbol,
      !is.na(metabolite_refmet) ~ metabolite_refmet, 
      TRUE ~ featureName
    )
  ) %>%
  select(
    -c(gene_symbol, metabolite_refmet)
  ) %>% 
  # Select the most extreme z-score by feature
  arrange(
    ome, contrast, desc(abs(z))
  ) %>%
  filter(
    .by = c(ome, contrast),
    !duplicated(featureName2)
  ) %>%
  mutate(
    sex = substr(x = as.character(contrast), 1L, 1L)
  ) %>%
  mutate(
    min_padj_M = ifelse(
      sex == "M", 
      adj.P.Val,
      NA_real_
    ),
    min_padj_F = ifelse(
      sex == "F",
      adj.P.Val,
      NA_real_
    )
  ) %>% 
  mutate(
    .by = c(ome, featureName),
    min_padj_M = ifelse(
      all(is.na(min_padj_M)),
      NA_real_,
      min(min_padj_M, na.rm = TRUE)
    ),
    min_padj_F = ifelse(
      all(is.na(min_padj_F)),
      NA_real_,
      min(min_padj_F, na.rm = TRUE)
    )
  ) %>%
  tidyr::pivot_wider(
    id_cols = c(ome, featureName, featureName2, contains("min_padj")),
    names_from = contrast,
    values_from = z
  ) %>% 
  # Complete data for clustering
  filter(
    complete.cases(
      across(.cols = ends_with("W", ignore.case = FALSE))
    )
  )

head(combined_DA)
```

```{r create-eset-list}
expr_mat <- combined_DA %>% 
  mutate(row_names = paste(ome, featureName)) %>% 
  tibble::column_to_rownames("row_names")

f_data <- expr_mat %>% 
  select(ome, featureName, featureName2, contains("min_padj"))

expr_mat <- select(expr_mat, -colnames(f_data)) %>%
  as.matrix()

eset <- ExpressionSet(
  assayData = expr_mat, 
  featureData = AnnotatedDataFrame(f_data)
)

eset # summary of ExpressionSet object
```

```{r reorder-clusters}
# Reorder clusters so those with similar trajectories are consecutive
reorder_clusters <- function(fclust) {
  d <- as.dist(1 - cor(t(fclust$centers)))
  hc <- hclust(d)
  neworder <- hc$order
  
  fclust$centers <- fclust$centers[neworder, ]
  rownames(fclust$centers) <- seq_len(nrow(fclust$centers))
  
  fclust$size <- fclust$size[neworder]
  
  fclust$cluster[] <- match(fclust$cluster, neworder)
  fclust$membership <- fclust$membership[, neworder]
  colnames(fclust$membership) <- seq_len(ncol(fclust$membership))
  
  return(fclust)
}
```

```{r eval=TRUE}
# Fuzzy C-Means (FCM) clustering ----

# Split by sex
eset_list <- lapply(c("F", "M"), function(sex_i) {
  eset[, grepl(sex_i, colnames(eset))]
}) %>% 
  setNames(c("Female", "Male"))

# Initial hierarchical clustering to define starting centroids for FCM
f1 <- function(eset, num_clusters) {
  d <- dist(exprs(eset))
  
  hc <- hclust(d = d)
  clusters <- cutree(tree = hc, k = num_clusters)
  clusters <- as.factor(as.numeric(clusters))
  
  # Compute cluster centroids (mean value by contrast)
  design_mat <- model.matrix(
    ~ 0 + clusters
  )
  colnames(design_mat) <- levels(clusters)
  
  cluster_size <- colSums(design_mat)
  
  centroid_mat <- t(design_mat) %*% exprs(eset) / cluster_size
  
  return(centroid_mat)
}


FCM <- lapply(names(eset_list), function(sex_i) {
  eset_i <- eset_list[[sex_i]]
  col <- paste0("min_padj_", substr(sex_i, 1L, 1L))
  eset_i_sub <- eset_i[fData(eset_i)[[col]] < 0.05, ]
  
  # Cluster significant features for a particular sex only. This creates higher
  # quality centroids, since features that do not change are not included (those
  # features would cause the centroids to flatten out). 
  # However, this generates clusters that don't directly input to tools like CAMERA-PR, and so this isn't actually used. 
  
  # fcm_1 <- mfuzz(
  #   eset = eset_i_sub,
  #   centers = 12L, # number of clusters
  #   m = 2,
  #   iter.max = 100L
  # )
  
  set.seed(0L)
  
  # Calculate membership probabilities for all features
  fcm_2 <- mfuzz(
    eset = eset_i,
    centers = f1(eset = eset_i_sub, num_clusters = 12L),
    m = 2,
    iter.max = 1L # more iterations will flatten clusters
  ) %>% 
    reorder_clusters()
  
  return(fcm_2)
}) %>% 
  setNames(names(eset_list))

```

```{r, eval = F}
ggplot_cmeans(
  cl = FCM[["Female"]],
  zmat = exprs(eset_list[["Female"]]),
  min.membership = 0,
  common_ylim = TRUE,
  ncol = 5L,
  filename = "../plots/female_clusters_new.pdf"
)

ggplot_cmeans(
  cl = FCM[["Male"]],
  zmat = exprs(eset_list[["Male"]]),
  min.membership = 0,
  common_ylim = TRUE,
  ncol = 5L,
  filename = "../plots/male_clusters_new.pdf"
)
```



# Localization of Gene Sets to Clusters

We can use the matrix of cluster membership scores as input for the non-parametric version of CAMERA-PR and perform upper-tail tests to test for localization of gene sets to each cluster. That is, for each gene set, what cluster trajector(y/ies) does it most closely follow?

## Global Proteomics

```{r FCM-CAMERA}
FCM_CAMERA <- lapply(unique(fData(eset)[["ome"]]), function(ome_i) {
  keep <- fData(eset)[["ome"]] == ome_i
  mem <- FCM_merged$membership[keep, , drop = FALSE]
  
  # Convert between identifiers
  rownames(mem) <- fData(eset)[["featureName2"]][keep]
  
  index_list <- list(
    "PROT_GN" = prot_GN_gene_sets,
    "TRNSCRPT_GN" = trnscrpt_GN_gene_sets,
    "METAB_GN" = metab_GN_refmet_sets
  )
  
  cameraPR(statistic = mem, 
           index = index_list[[ome_i]], 
           inter.gene.cor = 0.01,
           use.ranks = TRUE, # non-parametric
           adjust.globally = TRUE, 
           alternative = "greater", # upper-tail tests
           min.size = 5L) %>% 
    rename(Cluster = Contrast, PAdj = FDR) %>% 
    left_join(SET_TO_ID, by = "GeneSet") %>% 
    mutate(NRatio = round(NGenes / NGenesHuman, digits = 3)) %>% 
    # Reorder columns
    select(Cluster, GeneSetID, GeneSet, GeneSetShort,
           NGenes, NGenesHuman, NRatio, Direction, PValue, PAdj)
}) %>% 
  setNames(unique(fData(eset)[["ome"]])) %>% 
  bind_rows(.id = "ome") %>% 
  mutate(GeneSetShort = ifelse(is.na(GeneSetShort), GeneSet, GeneSetShort))

head(FCM_CAMERA)
```

```{r}
# Number of gene sets significantly localizing to each cluster
# filter(FCM_CAMERA, PAdj < 0.05) %>% 
#   count(ome, Cluster)
```

```{r prot-ORA}
FCM_ORA <- lapply(unique(fData(eset)[["ome"]]), function(ome_i) {
  cl <- FCM_merged
  
  keep <- fData(eset)[["ome"]] == ome_i
  cluster <- cl$cluster[keep]
  
  # Convert between identifiers
  names(cluster) <- fData(eset)[["featureName2"]][keep]
  cluster <- cluster[order(cluster, decreasing = FALSE)]
  cluster_list <- split(x = names(cluster),
                        f = factor(cluster,
                                   levels = as.character(unique(cluster))))
  
  index_list <- list(
    "PROT_GN" = prot_GN_gene_sets,
    "TRNSCRPT_GN" = trnscrpt_GN_gene_sets,
    "METAB_GN" = metab_GN_refmet_sets
  )
  
  out <- run_cluster_ORA(cluster_list = cluster_list,
                         index = index_list[[ome_i]],
                         min_size = 5L) %>%
    rename(GeneSet = set) %>% 
    left_join(SET_TO_ID, by = "GeneSet") %>% 
    relocate(all_of(colnames(SET_TO_ID)), .after = "GeneSet")
  
  return(out)
}) %>% 
  setNames(unique(fData(eset)[["ome"]])) %>% 
  bind_rows(.id = "ome") %>% 
  mutate(GeneSetShort = ifelse(is.na(GeneSetShort), GeneSet, GeneSetShort))

head(FCM_ORA)
```


```{r eval=FALSE}
# Save
usethis::use_data(FCM, overwrite = TRUE, version = 3)
usethis::use_data(FCM_CAMERA, overwrite = TRUE, version = 3)
usethis::use_data(FCM_ORA, overwrite = TRUE, version = 3)
```

## Plot Top Terms by Cluster

```{r camera-bubble-heatmaps, eval=save_plots}
## Bubble heatmaps of tests for gene set localization
for (ome_i in unique(FCM_CAMERA$ome)) {
  x <- FCM_CAMERA %>% 
    filter(ome == ome_i)
  
  # Select the top 6 terms by cluster
  top_terms <- x %>% 
    filter(PAdj < 0.05 + 0.05 * (ome_i == "PHOSPHO_GN")) %>% 
    slice_min(PValue, n = 6L, by = Cluster) %>%
    pull(GeneSetShort) %>% 
    unique()
  
  x <- filter(x, GeneSetShort %in% top_terms) %>% 
    mutate(logP = -log10(PValue))
  
  # Dynamic heatmap height. Each cell of the heatmap is 14 pt, by default
  height <- convertUnit(length(top_terms) * unit(15, "pt"), "in")
  height <- max(as.numeric(height) + 0.2, 4)
  
  # Dynamic heatmap width
  row_label_width <- ComplexHeatmap::max_text_width(x[["GeneSetShort"]])
  row_label_width <- convertUnit(row_label_width, "in")
  
  width <- convertUnit(nlevels(x$Cluster) * unit(15, "pt"), "in")
  width <- as.numeric(width + row_label_width) + 2.5
  
  enrichmap(x = x,
            plot_sig_only = FALSE,
            n_top = Inf, # plot all terms
            statistic_column = "logP",
            set_column = "GeneSetShort",
            padj_column = "PAdj",
            padj_legend_title = "BH Adjusted\nP-Value",
            contrast_column = "Cluster",
            heatmap_args = list(
              rect_gp = gpar(fill = "white", col = "grey85"),
              heatmap_legend_param = list(
                title = latex2exp::TeX(
                  "$\\bf{$-$log_{10}(P$-$Value)}$")
              ),
              column_names_side = "top"
            ),
            color = c("white", "#503080"),
            filename = file.path("..", "plots", "fuzzy_cmeans",
                                 sprintf("%s-cmeans_CAMERA.png",
                                         ome_i)),
            height = height, width = width)
}
```


```{r bubble-heatmaps, eval=save_plots}
## Bubble heatmaps of tests for gene set localization
for (ome_i in unique(FCM_CAMERA$ome)) {
  x <- FCM_ORA %>% 
    filter(ome == ome_i) %>% 
    rename(Cluster = cluster, PValue = p_value, PAdj = adj_p_value)
  
  # Select the top 6 terms by cluster
  top_terms <- x %>% 
    filter(PAdj < 0.05 + 0.05 * (ome_i == "PHOSPHO_GN")) %>% 
    slice_min(PValue, n = 6L, by = Cluster) %>%
    pull(GeneSetShort) %>% 
    unique()
  
  x <- filter(x, GeneSetShort %in% top_terms) %>% 
    mutate(logP = -log10(PValue))
  
  # Dynamic heatmap height. Each cell of the heatmap is 14 pt, by default
  height <- convertUnit(length(top_terms) * unit(15, "pt"), "in")
  height <- max(as.numeric(height) + 0.2, 4)
  
  # Dynamic heatmap width
  row_label_width <- ComplexHeatmap::max_text_width(x[["GeneSetShort"]])
  row_label_width <- convertUnit(row_label_width, "in")
  
  width <- convertUnit(nlevels(x$Cluster) * unit(15, "pt"), "in")
  width <- as.numeric(width + row_label_width) + 2.5
  
  enrichmap(x = x,
            plot_sig_only = FALSE,
            n_top = Inf, # plot all terms
            statistic_column = "logP",
            set_column = "GeneSetShort",
            padj_column = "PAdj",
            padj_legend_title = "BH Adjusted\nP-Value",
            contrast_column = "Cluster",
            heatmap_args = list(
              rect_gp = gpar(fill = "white", col = "grey85"),
              heatmap_legend_param = list(
                title = latex2exp::TeX(
                  "$\\bf{$-$log_{10}(P$-$Value)}$")
              ),
              column_names_side = "top"
            ),
            color = c("white", "#503080"),
            filename = file.path("..", "plots", "fuzzy_cmeans",
                                 sprintf("%s-cmeans_ORA.png",
                                         ome_i)),
            height = height, width = width)
}
```

```{r}
# layer_fun <- function(j, i, x, y, w, h, f) {
#   # Cell background
#   grid.rect(x = x, y = y, width = w, height = h,
#             gp = gpar(col = heatmap_args[["rect_gp"]][["col"]],
#                       fill = ifelse(
#                         is.na(pindex(padj_mat, i, j)),
#                         heatmap_args[["na_col"]], # black
#                         ifelse(pindex(padj_mat, i, j) < padj_cutoff,
#                                padj_fill, # grey
#                                heatmap_args[["rect_gp"]][["fill"]]) # white
#                       )
#             ))
# 
#   # Matrix of bubble diameters (optionally scaled to row or column max)
#   dmat <- -log10(padj_mat)
# 
#   if (scale_by == "max") {
#     # Scale bubbles relative to global max
#     for (method in c("CAMERA", "ORA")) {
#       idx <- grep(method, colnames(dmat))
#       dmat[, idx] <- dmat[, idx] / max(dmat[, idx], na.rm = TRUE)
#     }
# 
#   } else {
#     # Scale bubbles relative to row or column max
#     margin <- 1L + (scale_by == "column")
#     for (method in c("CAMERA-PR", "ORA")) {
#       idx <- grep(method, colnames(dmat))
#       dmat[, idx] <- sweep(dmat[, idx], MARGIN = margin,
#                            apply(dmat[, idx], MARGIN = margin,
#                                  max, na.rm = TRUE),
#                            FUN = "/")
#     }
#   }
# 
#   # Limits on bubble diameters
#   r_min <- 0.20 # only applies to significant adjusted p-values
#   r_max <- 0.95 # upper limit because a black border is added to the bubbles
#   dmat <- ifelse(padj_mat < padj_cutoff,
#                  dmat * (r_max - r_min) + r_min,
#                  dmat * r_max)
# 
#   # Draw bubbles
#   grid.circle(
#     x = x, y = y,
#     r = pindex(dmat, i, j) / 2 * cell_size,
#     # Significant bubbles get a black outline to separate from padj_fill
#     gp = gpar(col = ifelse(pindex(padj_mat, i, j) < padj_cutoff,
#                            "black", NA),
#               fill = f)
#   )
# }
```



```{r}
# ## Bubble heatmaps of tests for gene set localization
# cmeans_res <- list(
#   "CAMERA-PR" = list("PROT_GN" = PROT_GN_CMEANS_CAMERA,
#                   "TRNSCRPT_GN" = TRNSCRPT_GN_CMEANS_CAMERA,
#                   "METAB_GN" = METAB_GN_CMEANS_CAMERA),
#   "ORA" = list("PROT_GN" = PROT_GN_CMEANS_ORA,
#                "TRNSCRPT_GN" = TRNSCRPT_GN_CMEANS_ORA,
#                "METAB_GN" = METAB_GN_CMEANS_ORA)
# ) %>%
#   purrr::list_transpose() %>% 
#   lapply(function(xi) {
#     data.table::rbindlist(xi, fill = TRUE, idcol = "method") %>% 
#       mutate(method = factor(method, levels = c("CAMERA-PR", "ORA")),
#              Cluster = paste(method, Cluster),
#              Cluster = factor(Cluster, levels = unique(Cluster)))
#   })
# 
# for (ome_i in names(cmeans_res)) {
#   for (sex_i in c("Female", "Male")) {
#     message(paste(ome_i, sex_i))
#     
#     x <- cmeans_res[[ome_i]] %>% 
#       filter(Sex == sex_i) %>% 
#       rename(any_of(c("Set" = "GeneSet", "Set" = "Subclass"))) %>% 
#       mutate(Cluster2 = sub(".* ", "", Cluster),
#              Cluster2 = factor(Cluster2, levels = unique(Cluster2)))
#     
#     # Select the top 6 terms by cluster
#     top_terms <- x %>% 
#       filter(PAdj < 0.05) %>% 
#       slice_min(PValue, n = 6L, by = c(method, Cluster)) %>%
#       pull(Set) %>% 
#       unique()
#     
#     x <- filter(x, Set %in% top_terms) %>% 
#       mutate(logP = -log10(PValue))
#     
#     # Dynamic heatmap height. Each cell of the heatmap is 14 pt, by default
#     height <- convertUnit(length(top_terms) * unit(15, "pt"), "in")
#     height <- max(as.numeric(height) + 0.2, 4)
#     
#     set_column <- ifelse(grepl("^METAB", ome_i), "Set", "GeneSetShort")
#     
#     # Dynamic heatmap width
#     row_label_width <- ComplexHeatmap::max_text_width(x[[set_column]])
#     row_label_width <- convertUnit(row_label_width, "in")
#     
#     width <- convertUnit(nlevels(x$Cluster) * unit(15, "pt"), "in")
#     width <- as.numeric(width + row_label_width) + 2.5
#     
#     foo <- distinct(x, method, Cluster, Cluster2) %>% 
#       arrange(method, Cluster)
#     
#     enrichmap(x,
#               plot_sig_only = FALSE,
#               n_top = Inf, # plot all terms
#               statistic_column = "logP",
#               set_column = set_column,
#               padj_column = "PAdj",
#               padj_legend_title = "BH Adjusted\nP-Value",
#               contrast_column = "Cluster",
#               heatmap_args = list(
#                 layer_fun = layer_fun,
#                 column_split = foo$method,
#                 column_labels = foo$Cluster2,
#                 rect_gp = gpar(fill = "white", col = "grey85"),
#                 heatmap_legend_param = list(
#                   title = latex2exp::TeX(
#                     "$\\bf{$-$log_{10}(P$-$Value)}$")
#                 ),
#                 column_names_side = "top",
#                 column_names_rot = 0,
#                 column_names_centered = TRUE,
#                 column_title_gp = gpar(fontsize = unit(9, "pt"),
#                                        fontface = "bold")
#               ),
#               color = c("white", "#503080"),
#               filename = file.path("..", "plots", "fuzzy_cmeans",
#                                    sprintf("%s-%s-cmeans_combined.png",
#                                            ome_i, tolower(sex_i))),
#               height = height, width = width)
#   }
# }
```



# Session Info

```{r}
print(sessionInfo(), locale = FALSE)
```

