---
title: "Gastrocnemius Metabolomics RefMet Subclass Testing"
author: "Tyler Sagendorf"
date: "Last compiled `r format(Sys.time(), '%d %B, %Y')`"
output: 
  rmarkdown::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Gastrocnemius Metabolomics RefMet Subclass Testing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  fig.width = 6, 
  fig.height = 4
)
```

```{r setup, message=FALSE, warning=FALSE}
# Required packages
library(Biobase)
library(dplyr)
library(fgsea) # fgsea  
library(MotrpacRatTraining6moMuscleData)
library(TMSig) # filter_sets
library(stats)
```

# Prepare Sets

```{r}
background <- unique(fData(METAB_GN)$metabolite_refmet)
length(background) # 1046 metabolites/lipids
```

```{r}
metab_GN_refmet_sets <- with(fData(METAB_GN), 
                             split(metabolite_refmet, sub_class)) %>% 
  filter_sets(min_size = 5L)

length(metab_GN_refmet_sets) # 27
```

We will plot the sizes of the RefMet chemical subclass sets.

```{r, fig.height=3, fig.width=5}
set_sizes <- lengths(metab_GN_refmet_sets)
hist(set_sizes, breaks = 30)
```


```{r}
(coverage <- length(unique(unlist(metab_GN_refmet_sets))) /
   length(background)) # 0.5908222
```

About `r round(coverage * 100, 1)`% of the metabolites are captured by the RefMet chemical subclass sets.



# RefMet Chemical Subclass Testing

For each contrast and metabolite/lipid pairing, we will calculate the mean of the z-scores. This will be our input for FGSEA and CAMERA.

```{r}
# Collapse redundant metabolites
metab_GN_stats <- lapply(METAB_GN_DA, function(x) {
  x %>% 
    select(contrast, metabolite_refmet, z) %>% 
    # Faster `slice_max(order_by = abs(z))`
    arrange(contrast, metabolite_refmet, -abs(z)) %>% 
    filter(!duplicated(metabolite_refmet), .by = contrast) %>% 
    tidyr::pivot_wider(id_cols = metabolite_refmet, 
                       names_from = contrast, 
                       values_from = z) %>% 
    as.data.frame() %>% 
    `rownames<-`(.$metabolite_refmet) %>% 
    select(-metabolite_refmet) %>% 
    as.matrix()
})

# How much is the data collapsed? To about 86% of original
sapply(metab_GN_stats, nrow) / nrow(METAB_GN)

# Proportion of duplicates (affects FGSEA)
vapply(metab_GN_stats, function(x) mean(duplicated(x)), numeric(1L))
```


## FGSEA

We will perform fast gene set enrichment analysis (FGSEA) for each contrast.

```{r eval=FALSE, message=FALSE, warning=FALSE}
METAB_GN_FGSEA <- lapply(metab_GN_stats, function(x) {
  lapply(colnames(x), function(contrast_i) {
    
    stats_i <- na.omit(x[, contrast_i])

    set.seed(9001)
    
    fgseaMultilevel(
      pathways = metab_GN_refmet_sets, 
      stats = stats_i, 
      # precision of p-values
      eps = 0, 
      nproc = 4,
      # Number of permutations. Affects precision of NES and p-values.
      nPermSimple = 1e4
    ) %>% 
      mutate(Contrast = contrast_i)
  }) %>% 
    bind_rows() %>%
    rename(Subclass = pathway, NMetabolites = size,
           PValue = pval, PAdj = padj,
           LeadingEdge = leadingEdge) %>%
    mutate(
      Contrast = factor(Contrast, levels = unique(Contrast)),
      Direction = ifelse(NES >= 0, "Up", "Down"),
      # Adjust p-values across related contrasts
      PAdj = p.adjust(PValue, method = "BH"),
      Direction = factor(Direction, levels = c("Up", "Down"))
    ) %>%
    # Reorder columns
    select(Contrast, Subclass, NMetabolites,
           Direction, NES,
           PValue, PAdj, LeadingEdge) %>%
    arrange(Contrast, PValue, NES)
})
```

```{r}
# Count number of significant gene sets
lapply(METAB_GN_FGSEA, function(x) with(x, table(Contrast, PAdj < 0.05)))
```

```{r}
# Format of results
head(METAB_GN_FGSEA$trained_vs_SED)
```


## CAMERA

We will also perform Correlation Adjusted MEan RAnk gene set testing (CAMERA) with `TMSig::cameraPR.matrix`, a fast matrix method for `limma::cameraPR`.

```{r eval=FALSE, message=FALSE}
# CAMERA (pre-ranked)
METAB_GN_CAMERA <- lapply(metab_GN_stats, function(x) {
  cameraPR(statistic = x, 
           index = metab_GN_refmet_sets, 
           adjust.globally = TRUE) %>% 
    rename(Subclass = GeneSet, NMetabolites = NGenes, PAdj = FDR) %>% 
    mutate(
      ZScore = zscoreT(TwoSampleT, df = df, method = "hill"),
      Direction = factor(Direction, levels = c("Up", "Down"))
    ) %>% 
    select(Contrast, Subclass, NMetabolites, 
           Direction, TwoSampleT, df, ZScore, PValue, PAdj)
})
```

```{r}
# Count number of significant subclasses
lapply(METAB_GN_CAMERA, function(x) with(x, table(Contrast, PAdj < 0.05)))
```

```{r}
# Format of results
head(METAB_GN_CAMERA$trained_vs_SED)
```


```{r eval=FALSE}
# Save results
usethis::use_data(metab_GN_refmet_sets, overwrite = TRUE, version = 3)
usethis::use_data(METAB_GN_FGSEA, overwrite = TRUE, version = 3)
usethis::use_data(METAB_GN_CAMERA, overwrite = TRUE, version = 3)
```


# Session Info

```{r}
sI <- sessionInfo()
print(sI, locale = FALSE)
```
