---
title: "Normalized Phosphoproteomics Kinase-Substrate Enrichment Analysis"
author: "Tyler Sagendorf"
date: "Last compiled `r format(Sys.time(), '%d %B, %Y')`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Normalized Phosphoproteomics Kinase-Substrate Enrichment Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
# Required packages
library(MotrpacRatTraining6moMuscleData)
library(dplyr)
library(fgsea)
library(TMSig)
```

# Prepare Phosphoproteomics DA Results

```{r}
y <- lapply(PHOSPHO_GN_NORM_DA, function(yi) {
  yi %>% 
    select(contrast, featureName, z) %>% 
    left_join(RAT_TO_HUMAN_SITE, 
              by = c("featureName" = "ptm_id_rat_refseq")) %>% 
    rename(human_feature = ptm_id_human_uniprot) %>% 
    mutate(
      is_human = !is.na(human_feature),
      # If the human phosphosite is missing, use the rat site to avoid removing
      # or collapsing those rows.
      human_feature = ifelse(is_human,
                             human_feature,
                             featureName),
      # Separate UniProt ID from sites to split sites
      human_uniprot = ifelse(is_human, 
                             sub("_.*", "", human_feature),
                             sub("([A-Z]P_[^_]+).*", "\\1", human_feature)),
      human_site = sub(".*_", "", human_feature),
      human_site = strsplit(human_site, split = ";")
    ) %>% 
    tidyr::unnest(cols = human_site) %>% 
    # Select row with largest absolute z-score for each site
    arrange(contrast, human_uniprot, human_site, -abs(z)) %>% 
    # Total number of z-scores per site. Only the most extreme is kept
    mutate(n = n(), .by = c(contrast, human_uniprot, human_site)) %>% 
    filter(!duplicated(human_site), .by = c(contrast, human_uniprot)) %>% 
    mutate(featureID = paste0(human_uniprot, "-", human_site))
})
```


```{r}
# What proportion of sites are on human proteins? About 0.683
prop_human_sites <- bind_rows(y) %>% 
  distinct(featureID, is_human) %>% 
  pull(is_human) %>% 
  mean()
prop_human_sites # 0.6579831
```

About `r round(100 * prop_human_sites, digits = 1)`% of the phosphorylation sites are on human proteins. Only these sites have the potential to be included in the analysis.

```{r}
# Histogram of number of rows per site
n_rows_per_site <- bind_rows(y) %>% 
  arrange(featureID, -n) %>% 
  filter(!duplicated(featureID)) %>% 
  pull(n)

hist(n_rows_per_site, breaks = 1:20, 
     xlab = NULL, main = "Histogram of z-scores per site")
```

The majority of sites (`r round(100 * mean(n_rows_per_site == 1), digits = 1)`)% were singly phosphorylated.

```{r}
# We will reformat the results to a list of matrices
phospho_GN_stats <- lapply(y, function(yi) {
  yi %>% 
    tidyr::pivot_wider(id_cols = featureID, 
                       names_from = contrast, 
                       values_from = z) %>% 
    # Avoids tibble import for column_to_rownames
    as.data.frame() %>% 
    `rownames<-`(.$featureID) %>% 
    select(-featureID) %>% 
    as.matrix()
})
```



# Prepare Kinase Sets

```{r}
background <- lapply(phospho_GN_stats, rownames) %>% 
  unlist() %>% 
  unique()

table(unique(unlist(human_kinase_sets)) %in% background)
```

Only `r sum(unique(unlist(human_kinase_sets)) %in% background)` human sites in the background (~`r round(100 * mean(unique(unlist(human_kinase_sets)) %in% background), digits = 1)`%) are known to be phosphorylated by at least one kinase.

```{r}
phospho_GN_NORM_kinase_sets <- filter_sets(human_kinase_sets,
                                      background = background,
                                      min_size = 3L,
                                      max_size = length(background) - 1L)

length(human_kinase_sets) # 319
length(phospho_GN_NORM_kinase_sets) # 103
```

Only `r length(phospho_GN_NORM_kinase_sets)` kinase sets remain for testing.

```{r}
hist(lengths(phospho_GN_NORM_kinase_sets), breaks = 20)
```

Most kinases have few sites.



# KSEA

## FGSEA

We will perform fast gene set enrichment analysis (FGSEA) for each contrast.

```{r eval=FALSE, message=FALSE, warning=FALSE}
PHOSPHO_GN_NORM_FGSEA <- lapply(phospho_GN_stats, function(x) {
  lapply(colnames(x), function(contrast_i) {
    
    stats_i <- na.omit(x[, contrast_i])

    set.seed(9001)
    
    fgseaMultilevel(
      pathways = phospho_GN_NORM_kinase_sets, 
      stats = stats_i,
      # precision of p-values
      eps = 0, 
      nproc = 4,
      # Number of permutations. Affects precision of NES and p-values.
      nPermSimple = 1e4
    ) %>% 
      mutate(Contrast = contrast_i)
  }) %>% 
    bind_rows() %>% 
    rename(Kinase = pathway, NSites = size, 
           PValue = pval, PAdj = padj, 
           LeadingEdge = leadingEdge) %>% 
    mutate(
      Contrast = factor(Contrast, levels = unique(Contrast)),
      Direction = ifelse(NES >= 0, "Up", "Down"),
      PAdj = p.adjust(PValue, method = "BH"),
      NSitesPSP = lengths(human_kinase_sets)[Kinase],
      Direction = factor(Direction, levels = c("Up", "Down"))
    ) %>% 
    # Reorder columns
    select(Contrast, Kinase, NSites, NSitesPSP, 
           Direction, NES, 
           PValue, PAdj, LeadingEdge) %>% 
    arrange(Contrast, PValue, NES)
})
```

```{r}
# Count number of significant gene sets
lapply(PHOSPHO_GN_NORM_FGSEA, 
       function(x) with(x, table(Contrast, PAdj < 0.05)))
```

```{r}
# Format of results
head(PHOSPHO_GN_NORM_FGSEA$trained_vs_SED)
```


## CAMERA

We will also perform Correlation Adjusted MEan RAnk gene set testing (CAMERA) with `TMSig::cameraPR.matrix`, a fast matrix method for `limma::cameraPR`.

```{r eval=FALSE, message=FALSE}
# CAMERA (pre-ranked)
PHOSPHO_GN_NORM_CAMERA <- lapply(phospho_GN_stats, function(x) {
  cameraPR(statistic = x,
           index = phospho_GN_NORM_kinase_sets, 
           adjust.globally = TRUE, 
           min.size = 3L) %>% 
    rename(Kinase = GeneSet, NSites = NGenes, PAdj = FDR) %>% 
    mutate(
      NSitesPSP = lengths(human_kinase_sets)[Kinase],
      # Modifications to reduce object size
      ZScore = zscoreT(TwoSampleT, df = df, method = "hill"),
      Direction = factor(Direction, levels = c("Up", "Down"))
    ) %>%
    # Reorder columns
    select(Contrast, Kinase, NSites, NSitesPSP,
           Direction, TwoSampleT, df, ZScore, PValue, PAdj)
})
```

```{r}
# Count number of significant gene sets
lapply(PHOSPHO_GN_NORM_CAMERA, 
       function(x) with(x, table(Contrast, PAdj < 0.05)))
```

```{r}
# Format of results
head(PHOSPHO_GN_NORM_CAMERA$trained_vs_SED)
```


```{r eval=FALSE}
# Save results
usethis::use_data(phospho_GN_NORM_kinase_sets, overwrite = TRUE, version = 3)
usethis::use_data(PHOSPHO_GN_NORM_FGSEA, overwrite = TRUE, version = 3)
usethis::use_data(PHOSPHO_GN_NORM_CAMERA, overwrite = TRUE, version = 3)
```


# Session Info

```{r}
sI <- sessionInfo()
print(sI, locale = FALSE)
```
