---
title: "SKMGN Proteomics Gene Set Testing"
author: "Tyler Sagendorf"
date: "Last compiled `r format(Sys.time(), '%d %B, %Y')`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{SKMGN Proteomics Gene Set Testing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>", 
fig.width = 6, 
fig.height = 4
)
```

```{r setup, message=FALSE, warning=FALSE}
# Required packages
library(Biobase)
library(data.table)
library(dplyr)
library(fgsea)
library(MotrpacRatTraining6moMuscleData)
library(TMSig)
library(stats)
```

# Prepare Gene Sets

## Restrict Sets to Background

We begin by restricting the rat gene sets to only those 5932 genes from the proteomics expression data (the `background`).

```{r}
# Filter to genes only present in PROT_GN
background <- unique(fData(PROT_GN)[["gene_symbol"]],
                     incomparables = NA)

mean(!is.na(background)) # no genes are missing

length(background) # 5932
```

```{r}
prot_GN_gene_sets <- filterSets(rat_gene_sets, 
                                background = background, 
                                min_size = 5L, 
                                max_size = length(background) - 1L)
length(rat_gene_sets) # 9293
length(prot_GN_gene_sets) # 6317
```

Only `r length(prot_GN_gene_sets)` gene sets have at least 5 genes in the proteomics data. We will plot the sizes of the gene sets.

```{r}
rat_set_sizes <- lengths(prot_GN_gene_sets)
hist(rat_set_sizes, breaks = 30)
```

For each gene set, we calculate the ratio of the number of rat genes in the proteomics data to the original number of human genes. This gives us a measure of confidence for the set descriptions. That is, if the ratio is low, it is unlikely that what is being tested is correctly described by the set name.

```{r}
r <- rat_set_sizes / lengths(human_gene_sets)[names(rat_set_sizes)]
hist(r, breaks = seq(0, 1.4, 0.05))
```

Notice that some gene sets increase in size. This happens because there is a small proportion of human genes that map to multiple rat genes in our data. As shown below, this only affects smaller gene sets, as larger gene sets tend to lose more genes than what they may gain from a one-to-many mapping.

```{r}
plot(rat_set_sizes, r, log = "x")
abline(h = 1, col = "red")
```

```{r}
# Minimum ratio for top 90% of gene sets
(q <- quantile(r, probs = 1 - 0.9)) # 0.2077922
```

The top 90% of sets retain at least `r round(100 * q, 1)`% of their original genes.

```{r}
# How much of the background do the gene sets cover? 0.934592
(coverage <- length(unique(unlist(prot_GN_gene_sets))) / length(background))
```

The gene sets cover about `r round(coverage * 100, 1)`% of the unique genes in the proteomics data.


## Cluster Gene Sets

We will now try to reduce the redundancy of the sets. First, we will identify clusters of highly similar sets based on their Jaccard coefficients, $J$. Within each cluster, we will keep the largest set. Ties will be broken by selecting the set that was most similar to its human counterpart (ratio closest to 1), then the set with the shortest description, and finally according to alphanumeric order. See `?TMSig::cluster_sets` for details on the clustering algorithm.

For the GO gene sets, sets may be clustered together if $J \geq 0.85$.

```{r}
prot_cluster_df <- cluster_sets(prot_GN_gene_sets) %>% 
  mutate(diff = abs(1 - r[set]),
         n_char = nchar(set)) %>% 
  arrange(cluster, -set_size, diff, n_char, set)

# Show first few clusters with multiple GO gene sets
filter(prot_cluster_df, cluster <= 5L)
```

Now, we will take all the clusters with at least 2 sets and provide them as a separate table. In this table, we will have columns for the primary set kept for the analysis (`Set`), sets that appeared in the same cluster and were discarded (`SimilarSets`), the pairwise Jaccard coefficients (`JaccardCoef`), the pairwise overlap coefficients (`OverlapCoef`), and a unique ID for each set in `Sets` (`SetID`) that can be used to quickly cross-reference the analysis results. It is also useful for shortening the set descriptions for plotting while ensuring they are unique.

```{r}
# Sets that were clustered together
jaccard_sets <- prot_cluster_df %>% 
  as.data.frame() %>% 
  filter(n() > 1, .by = cluster) %>% 
  pull(set)

# Jaccard similarity matrix
jaccard <- similarity(prot_GN_gene_sets[jaccard_sets]) %>% 
  as.matrix()

# Overlap similarity matrix
overlap <- similarity(prot_GN_gene_sets[jaccard_sets], 
                      type = "overlap") %>% 
  as.matrix()

PROT_GN_REDUNDANT_SETS <- prot_cluster_df %>% 
  mutate(type = ifelse(duplicated(cluster), 
                       "SimilarGeneSets", "GeneSet")) %>% 
  tidyr::pivot_wider(id_cols = cluster,
                     names_from = type, 
                     values_from = set,
                     values_fn = list,
                     values_fill = list(character(0L))) %>% 
  select(-cluster) %>%
  tidyr::unnest(cols = c(GeneSet, SimilarGeneSets)) %>% 
  mutate(
    JaccardCoef = sapply(seq_len(n()), function(i) {
      jaccard[GeneSet[i], SimilarGeneSets[i]]
    }),
    OverlapCoef = sapply(seq_len(n()), function(i) {
      overlap[GeneSet[i], SimilarGeneSets[i]]
    }),
    across(.cols = ends_with("Coef"), 
           .fns = ~ round(.x, digits = 3))
  ) %>%
  left_join(select(SET_TO_ID, GeneSetID, GeneSet), by = "GeneSet") %>% 
  arrange(GeneSetID, -JaccardCoef, -OverlapCoef, SimilarGeneSets) %>% 
  relocate(GeneSetID, .before = GeneSet) %>% 
  as.data.frame()

head(PROT_GN_REDUNDANT_SETS)
```

Now, we can discard all but the first set from each cluster.

```{r}
# Select first set in each cluster (clusters were previously sorted)
prot_filtered_sets <- 
  prot_cluster_df[!duplicated(prot_cluster_df$cluster), "set"]

length(prot_filtered_sets) # 5984 sets to test

# Filter gene sets
prot_GN_gene_sets <- prot_GN_gene_sets[prot_filtered_sets]
```

Only `r length(prot_GN_gene_sets)` sets remain for testing. We will check if clustering impacted the coverage.

```{r}
# How much of the background do the gene sets cover after reducing redundancy?
# 0.934592
(coverage <- length(unique(unlist(prot_GN_gene_sets))) / length(background))
```

The gene sets cover about `r round(coverage * 100, 1)`% of the unique genes in the proteomics data. The redundancy of the gene sets was reduced without impacting the coverage.



# Gene Set Testing

For each contrast and gene symbol pairing, we will keep the most extreme z-score. This will be our input for FGSEA and CAMERA.

```{r}
# Collapse DA results to gene level
prot_GN_stats <- lapply(PROT_GN_DA, function(x) {
  x %>% 
    # If genes are missing, use the protein ID
    mutate(featureID = ifelse(!is.na(gene_symbol),
                              gene_symbol, 
                              featureName)) %>% 
    select(contrast, featureID, z) %>% 
    arrange(contrast, featureID, -abs(z)) %>% 
    filter(!duplicated(featureID), .by = contrast) %>% 
    tidyr::pivot_wider(id_cols = featureID, 
                       names_from = contrast, 
                       values_from = z) %>% 
    as.data.frame() %>% 
    `rownames<-`(.$featureID) %>% 
    select(-featureID) %>% 
    as.matrix()
})

# How much is the data collapsed? To about 96% of original
sapply(prot_GN_stats, nrow) / nrow(PROT_GN)

# Proportion of duplicates (affects FGSEA)
vapply(prot_GN_stats, function(x) mean(duplicated(x)), numeric(1L))
```


## FGSEA

We will perform fast gene set enrichment analysis (FGSEA) for each contrast.

```{r eval=FALSE, message=FALSE, warning=FALSE}
PROT_GN_FGSEA <- lapply(prot_GN_stats, function(x) {
  lapply(colnames(x), function(contrast_i) {
    
    # Extract the named vector of z-scores
    stats_i <- na.omit(x[, contrast_i])
    
    set.seed(9001)
    
    fgsea(
      pathways = prot_GN_gene_sets, 
      stats = stats_i,
      # precision of p-values
      eps = 0, 
      nproc = 4,
      # Number of permutations. Affects precision of NES and p-values.
      nPermSimple = 1e4
    ) %>% 
      mutate(Contrast = contrast_i)
  }) %>%
    bind_rows() %>% 
    rename(GeneSet = pathway, NGenesRat = size,
           PValue = pval, PAdj = padj, 
           LeadingEdge = leadingEdge) %>%
    # Include GeneSetID, NGenesHuman columns
    left_join(SET_TO_ID, by = "GeneSet") %>% 
    mutate(
      Contrast = factor(Contrast, levels = unique(Contrast)),
      Direction = ifelse(NES >= 0, "Up", "Down"),
      PAdj = p.adjust(PValue, method = "BH"),
      # NRatio is a measure of confidence that the set being tested is what is
      # described by GeneSet
      NRatio = NGenesRat / NGenesHuman,
      # Modifications to reduce object size
      NRatio = round(NRatio, digits = 3),
      Direction = factor(Direction, levels = c("Up", "Down"))
    ) %>% 
    # Reorder columns
    select(Contrast, GeneSetID, GeneSet, GeneSetShort, 
           NGenesRat, NGenesHuman, NRatio, 
           Direction, NES, 
           PValue, PAdj, LeadingEdge) %>% 
    arrange(Contrast, PValue, NES)
})
```

```{r}
# Count number of significant gene sets
lapply(PROT_GN_FGSEA, function(x) with(x, table(Contrast, PAdj < 0.05)))
```

```{r}
# Format of results
head(PROT_GN_FGSEA$trained_vs_SED)
```


## CAMERA

We will also perform Correlation Adjusted MEan RAnk gene set testing (CAMERA) with `TMSig::cameraPR.matrix`, a fast matrix method for `limma::cameraPR`.

```{r eval=FALSE, message=FALSE}
# CAMERA (pre-ranked)
PROT_GN_CAMERA <- lapply(prot_GN_stats, function(x) {
  cameraPR(statistic = x, 
           index = prot_GN_gene_sets, 
           adjust.globally = TRUE) %>% 
    rename(NGenesRat = NGenes, PAdj = FDR) %>%
    # Include GeneSetID, NGenesHuman columns
    left_join(SET_TO_ID, by = "GeneSet") %>%
    mutate(
      ZScore = zscoreT(TwoSampleT, df = df, method = "hill"),
      # NRatio is a measure of confidence that the set being tested is what is
      # described by GeneSet
      NRatio = NGenesRat / NGenesHuman,
      # Modifications to reduce object size
      NRatio = round(NRatio, digits = 3),
      Direction = factor(Direction, levels = c("Up", "Down"))
    ) %>%
    select(Contrast, GeneSetID, GeneSet, GeneSetShort, 
           NGenesRat, NGenesHuman, NRatio, 
           Direction, TwoSampleT, df, ZScore, PValue, PAdj)
})
```

```{r}
# Count number of significant gene sets
lapply(PROT_GN_CAMERA, function(x) with(x, table(Contrast, PAdj < 0.05)))
```

```{r}
# Format of results
head(PROT_GN_CAMERA$trained_vs_SED)
```

```{r eval=FALSE}
# Save results
usethis::use_data(prot_GN_gene_sets, overwrite = TRUE, version = 3)
usethis::use_data(PROT_GN_REDUNDANT_SETS, overwrite = TRUE, version = 3)
usethis::use_data(PROT_GN_FGSEA, overwrite = TRUE, version = 3)
usethis::use_data(PROT_GN_CAMERA, overwrite = TRUE, version = 3)
```


# Session Info

```{r}
sI <- sessionInfo()
print(sI, locale = FALSE)
```
