---
title: "SKMGN Redox Proteomics (Protein-Normalized) Overview"
author: "Christopher Jin"
date: "Last compiled `r format(Sys.time(), '%d %B, %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{SKMGN Redox Proteomics (Protein-Normalized) Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup, message=FALSE, warning=FALSE}
library(Biobase)
library(dplyr)
library(ggplot2)
library(MotrpacRatTraining6moMuscleData)
```

`REDOX_GN_NORM` contains gastrocnemius redox proteomics data where each cysteine site's abundance has been adjusted by the corresponding protein's abundance from global proteomics. This normalization removes protein-level changes to isolate regulation at the redox (cysteine oxidation) level.

# Data Summary

```{r}
eset = REDOX_GN_NORM

cat("Features:", nrow(eset), "\n")
cat("Samples: ", ncol(eset), "\n")
```

```{r}
# Sample breakdown by sex and timepoint
table(pData(eset)$sex, pData(eset)$timepoint)
```

# PCA

PCA is computed on the subset of features with no missing values across all samples.

```{r}
mat = exprs(eset)
mat_complete = mat[complete.cases(mat), ]
cat("Features used for PCA:", nrow(mat_complete), "\n")

pca_res = prcomp(t(mat_complete), center = TRUE, scale. = FALSE)
var_exp = summary(pca_res)$importance[2, 1:2] * 100

pd = as.data.frame(pData(eset))

pca_df = as.data.frame(pca_res$x[, 1:2])
pca_df$viallabel = rownames(pca_df)

pca_df = pca_df %>%
  left_join(pd[, c("viallabel", "sex", "timepoint")], by = "viallabel") %>%
  mutate(
    timepoint = recode(timepoint, "SED" = "control"),
    timepoint = factor(timepoint, levels = c("control", "1W", "2W", "4W", "8W"))
  )
```

```{r, fig.width=10, fig.height=8}
ggplot(pca_df, aes(x = PC1, y = PC2, color = sex, fill = timepoint, shape = sex)) +
  geom_point(size = 6, stroke = 1) +
  scale_color_manual(values = MotrpacRatTraining6moData::SEX_COLORS) +
  scale_fill_manual(values = c("control" = "white", "1W" = "#F7FCB9", "2W" = "#ADDD8E", "4W" = "#238443", "8W" = "#002612")) +
  scale_shape_manual(values = c("Female" = 21, "Male" = 24)) +
  labs(
    x = sprintf("PC1 (%.1f%%)", var_exp[1]),
    y = sprintf("PC2 (%.1f%%)", var_exp[2]),
    color = "Sex",
    fill = "Group",
    shape = "Sex",
    title = "REDOX_GN_NORM PCA"
  ) +
  guides(
    fill = guide_legend(override.aes = list(color = "grey30", shape = 21, size = 4)),
    color = guide_legend(override.aes = list(fill = NA, size = 4))
  ) +
  theme_bw()

# ggsave(filename = file.path(here(), "plots", "pcas", "redox_gn.pdf"))
```

# Session Info

```{r}
sessionInfo()
```
