---
title: "SKMGN RNA Splicing Gene Set Testing"
author: "Christopher Jin"
date: "Last compiled `r format(Sys.time(), '%d %B, %Y')`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
%\VignetteIndexEntry{SKMGN RNA Splicing Gene Set Testing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  fig.width = 6, 
  fig.height = 4
)
```

```{r setup, message=FALSE, warning=FALSE}
# Required packages
library(Biobase)
library(dplyr)
library(fgsea)
library(MotrpacRatTraining6moMuscleData)
library(TMSig) # set filtering, similarity, clustering
library(stats)
library(here)

source(file.path(here(), "Figures", "Splicing", "splicing_helper_functions.R"))
```

```{r}
splicing_transcripts = load_differential_splicing(selected_tissue = "gastroc")

background = tolower(unique(splicing_transcripts$gene_symbol))
```


We prepare the testable gene sets. I use tolower because of the way that I have
the gene names set up compared to what I previously had. 
```{r}
lower_rat_gene_sets = lapply(MotrpacRatTraining6moMuscleData::rat_gene_sets, tolower)
trnscrpt_GN_gene_sets <- TMSig::filterSets(lower_rat_gene_sets,
                                           background = background, 
                                           min_size = 5L,
                                           max_size = length(background) - 1L)
length(rat_gene_sets) # 9293
length(trnscrpt_GN_gene_sets) # 5382

```

So the number of remaining qualifying sets is lower, which is expected given the
number of transcripts that have multiple isoforms is quite a bit lower. So 
our background is effectively filtered a bit already. 

Only `r length(trnscrpt_GN_gene_sets)` gene sets have at least 5 genes in the transcriptomics data. We will plot the sizes of the gene sets.

```{r}
rat_set_sizes <- lengths(trnscrpt_GN_gene_sets)
hist(rat_set_sizes, breaks = 30)
```

## Cluster Gene Sets

We will now try to reduce the redundancy of the sets. First, we will identify clusters of highly similar sets based on their Jaccard coefficients. Within each cluster, we will keep the largest set. Ties will be broken by selecting the set that was most similar to its human counterpart (ratio closest to 1), then the set with the shortest description, and finally according to alphanumeric order.

```{r}
trnscrpt_cluster_df <- TMSig::clusterSets(trnscrpt_GN_gene_sets) %>% 
  mutate(ratio = r[set],
         diff = abs(1 - ratio),
         n_char = nchar(set)) %>% 
  arrange(cluster, desc(set_size), diff, n_char, set)

# Show first few clusters with multiple gene sets
filter(trnscrpt_cluster_df, cluster %in% 1:5)
```

Now, we will take all the clusters with at least 2 sets and provide them as a separate table. In this table, we will have columns for the primary set kept for the analysis (`GeneSet`), sets that appeared in the same cluster and were discarded (`SimilarGeneSets`), the pairwise Jaccard coefficients (`JaccardCoef`), the pairwise overlap coefficients (`OverlapCoef`), and a unique ID for each set in `GeneSet` (`GeneSetID`) that can be used to quickly cross-reference the FGSEA results. It is also useful for shortening the set descriptions for plotting while ensuring they are unique.

```{r}
# Sets that were clustered together
jaccard_sets <- trnscrpt_cluster_df %>% 
  filter(n() > 1L, .by = cluster) %>% 
  pull(set)

# Jaccard similarity matrix
jaccard <- similarity(trnscrpt_GN_gene_sets[jaccard_sets]) %>% 
  as.matrix()

# Overlap similarity matrix
overlap <- similarity(trnscrpt_GN_gene_sets[jaccard_sets], 
                      type = "overlap") %>% 
  as.matrix()

TRNSCRPT_GN_SPLICING_REDUNDANT_SETS <- trnscrpt_cluster_df %>% 
  mutate(type = ifelse(duplicated(cluster), 
                       "SimilarGeneSets", "GeneSet")) %>% 
  tidyr::pivot_wider(id_cols = cluster,
                     names_from = type, 
                     values_from = set,
                     values_fn = list,
                     values_fill = list(character(0L))) %>% 
  select(-cluster) %>%
  tidyr::unnest(cols = c(GeneSet, SimilarGeneSets)) %>% 
  mutate(
    JaccardCoef = sapply(seq_len(n()), function(i) {
      jaccard[GeneSet[i], SimilarGeneSets[i]]
    }),
    OverlapCoef = sapply(seq_len(n()), function(i) {
      overlap[GeneSet[i], SimilarGeneSets[i]]
    }),
    across(.cols = ends_with("Coef"), 
           .fns = ~ round(.x, digits = 3))
  ) 

# %>%
#   left_join(select(SET_TO_ID, GeneSetID, GeneSet), by = "GeneSet") %>% 
#   arrange(GeneSetID, -JaccardCoef, -OverlapCoef, SimilarGeneSets) %>% 
#   relocate(GeneSetID, .before = GeneSet) %>% 
#   as.data.frame()

head(TRNSCRPT_GN_SPLICING_REDUNDANT_SETS)
```

Now, we can discard all but the first set from each cluster.

```{r}
# Select first set in each cluster (clusters were previously sorted)
trnscrpt_filtered_sets <-
  trnscrpt_cluster_df[!duplicated(trnscrpt_cluster_df$cluster), "set"]
length(trnscrpt_filtered_sets) # 5067

# Filter gene sets
trnscrpt_GN_gene_sets <- trnscrpt_GN_gene_sets[trnscrpt_filtered_sets]
```

```{r}
# How much of the background do the gene sets cover after reducing redundancy?
coverage <- length(unique(unlist(trnscrpt_GN_gene_sets))) / length(background)
coverage # 0.9346893
```

The gene sets cover about `r round(coverage * 100, 1)`% of the genes in the transcriptomics data. The redundancy of the gene sets was reduced without impacting the coverage.



# Gene Set Testing

For each contrast and gene symbol pairing, we will keep the most extreme z-score. This will be our input for CAMERA.

```{r}
# For each combination of gene isoform, select the most extreme z-score per gene
trnscrpt_GN_splicing_stats = splicing_transcripts %>%
  mutate(featureID = ifelse(!is.na(gene_symbol),
                            gene_symbol, 
                            ensembl_gene)) %>% 
  group_by(featureID) %>%
  slice_max(order_by = F_statistic, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(featureID, F_statistic) %>% # we use F statistic bc the only test done is the LRT
  arrange(featureID, -abs(F_statistic)) %>%
  as.data.frame() %>% 
  `rownames<-`(tolower(.$featureID)) %>% 
  select(-featureID) %>% 
  filter(!is.na(F_statistic)) %>%
  as.matrix() 
```


## CAMERA-PR

We will perform fast gene set enrichment analysis (FGSEA) for each contrast.

```{r eval=T, message=FALSE, warning=FALSE}
set.seed(9001)

TRNSCRPT_SPLICING_GN_CAMERA = cameraPR(statistic = trnscrpt_GN_splicing_stats[,1],
                index = trnscrpt_GN_gene_sets) %>% 
    rename(NGenesRat = NGenes, PAdj = FDR) %>%
  tibble::rownames_to_column("Set") %>%
  select("Set", everything())

write.csv(TRNSCRPT_SPLICING_GN_CAMERA,
          file = file.path(here(),
                           "Figures",
                           "Splicing",
                           "data",
                           "GN_CAMERA.csv"),
          row.names = FALSE)
```

