---
title: "Fuzzy C-Means Clustering"
author: "Christopher Jin and Tyler Sagendorf"
date: "Last compiled `r format(Sys.time(), '%d %B, %Y')`"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  fig.width = 6, 
  fig.height = 4
)

save_plots <- dir.exists(file.path("..", "..", "plots"))
```

Documentation:

We present the CAMERA-PR results as the primary results in the manuscript, primarily
because the membership probability is an easy, continuous marker that can be used for CAMERA. 

One large reason I don't present the ORA (e.g. fischer exact for features in cluster) as the primary result is mainly due to concerns with the arbitrary
number of clusters, and the potential relatively small number of features in each cluster, which ORA can be very sensitive to. 

However, I realize just doing the fischer exact is the more common tool in the literature, and so those results are included for comparison purposes.


```{r setup, warning=FALSE, message=FALSE}
# Required packages
library(Biobase)
library(ComplexHeatmap) # max_text_width
library(dplyr)
library(ggplot2)
library(grid) # gpar, unit, convertUnit
library(latex2exp) # TeX
library(Mfuzz) # mfuzz, mestimate, mfuzz.plot2
library(MotrpacRatTraining6moMuscleData)
library(tibble)
library(tidyr)
library(TMSig)
library(here)
library(data.table)

num_clusters = 12
source(here::here("Other_Figures", "cmeans", "cmeans_helpers.R"))
```

# Prepare Input

```{r combine-DA-results}
combined_DA <- list(
  "PROT_GN" = PROT_GN_DA,
  "TRNSCRPT_GN" = TRNSCRPT_GN_DA,
  "METAB_GN" = METAB_GN_DA
) %>% 
  lapply(function(x) {
    x[["trained_vs_SED"]] %>% 
      select(contrast, featureName, z, P.Value, adj.P.Val,
             any_of(c("gene_symbol", "metabolite_refmet")))
  }) %>% 
  bind_rows(.id = "ome") %>% 
  mutate(
    contrast = sub(" .*", "", contrast),
    contrast = factor(contrast, levels = unique(contrast)),
    featureName2 = case_when(
      !is.na(gene_symbol) ~ gene_symbol,
      !is.na(metabolite_refmet) ~ metabolite_refmet, 
      TRUE ~ featureName
    )
  ) %>%
  select(
    -c(gene_symbol, metabolite_refmet)
  ) %>% 
  # Select the most extreme z-score by feature
  arrange(
    ome, contrast, desc(abs(z))
  ) %>%
  filter(
    .by = c(ome, contrast),
    !duplicated(featureName2)
  ) %>%
  mutate(
    sex = substr(x = as.character(contrast), 1L, 1L)
  ) %>%
  mutate(
    min_padj_M = ifelse(
      sex == "M", 
      adj.P.Val,
      NA_real_
    ),
    min_padj_F = ifelse(
      sex == "F",
      adj.P.Val,
      NA_real_
    )
  ) %>% 
  mutate(
    .by = c(ome, featureName),
    min_padj_M = ifelse(
      all(is.na(min_padj_M)),
      NA_real_,
      min(min_padj_M, na.rm = TRUE)
    ),
    min_padj_F = ifelse(
      all(is.na(min_padj_F)),
      NA_real_,
      min(min_padj_F, na.rm = TRUE)
    )
  ) %>%
  tidyr::pivot_wider(
    id_cols = c(ome, featureName, featureName2, contains("min_padj")),
    names_from = contrast,
    values_from = z
  ) %>% 
  # Complete data for clustering
  filter(
    complete.cases(
      across(.cols = ends_with("W", ignore.case = FALSE))
    )
  )

head(combined_DA)
```

```{r create-eset-list}
expr_mat <- combined_DA %>% 
  mutate(row_names = paste(ome, featureName)) %>% 
  tibble::column_to_rownames("row_names")

f_data <- expr_mat %>% 
  select(ome, featureName, featureName2, contains("min_padj"))

expr_mat <- select(expr_mat, -colnames(f_data)) %>%
  as.matrix()

eset <- ExpressionSet(
  assayData = expr_mat, 
  featureData = AnnotatedDataFrame(f_data)
)

eset # summary of ExpressionSet object
```


```{r eval=TRUE}
# Fuzzy C-Means (FCM) clustering ----
# Split by sex
eset_list <- lapply(c("F", "M"), function(sex_i) {
  eset[, grepl(sex_i, colnames(eset))]
}) %>% 
  setNames(c("Female", "Male"))

FCM = list()
for(sex_i in c("Female", "Male")){
  eset_i <- eset_list[[sex_i]]
  
  col <- paste0("min_padj_", substr(sex_i, 1L, 1L))
  #we plan on just simplifying and using only the significant 
  #features for this analysis.
  
  #if we are using 12 clusters -> we don't do a p-value threshold.
  #if we are using a smaller number of clusters -> we do filter. 
  eset_i_sub <- eset_i #[fData(eset_i)[[col]] < 0.05, ]
  
  #add the 0 columns representing the SED groups
  non_scaled = exprs(eset_i_sub)
  # scaled = t(scale(t(exprs(eset_i))))
  esets_zeros = cbind(rep(0, nrow(non_scaled)),
                      non_scaled)
  # message(paste0("Sig Feat: ",sex_i,  nrow(esets_zeros)))
  colnames(esets_zeros) = c("SED", colnames(eset_i))
  
  eset_cmeans <- ExpressionSet(
    assayData = esets_zeros, 
    featureData = AnnotatedDataFrame(fData(eset_i_sub))
  )
  
  set.seed(0L)
  # Calculate membership probabilities for all features
  fcm_2 <- mfuzz(
    eset = eset_cmeans,
    centers = num_clusters,
    m = 2,
    iter.max = 1L # more iterations will flatten clusters
  ) %>% 
    reorder_clusters()
  
  FCM[[sex_i]] = fcm_2
  
  file_name = file.path(here(), "plots/fuzzy_cmeans", paste0(sex_i, ".png"))
  png(file = file_name,
      width = 14, height = 12, units = "in", res = 1200)
  Mfuzz::mfuzz.plot2(eset_cmeans,
                     mfrow = c(3,4),
                     ylim.set = c(-4,4),
                     fcm_2,
                     min.mem = 0.3,
                     xlab = paste(sex_i, "Clusters"),
                     ylab = "",
                     centre = TRUE,
                     time.labels = colnames(esets_zeros),
                     col.axis = "black",
                     cex.lab = 1.5,
                     cex.axis = 1.5, #axis text size,
                     cex.main = 1.5, #main title text size
                     # single = i,
                     x11 = FALSE)
  dev.off()
}

# saveRDS(FCM, file = "~/Downloads/muscle_6mo_FCM.RDS")
```

# Localization of Gene Sets to Clusters

## Overrepresentation analysis

That is, for each gene set, what cluster trajector(y/ies) does it most closely follow?

```{r ORA}
index_list <- list(
  "PROT_GN" = prot_GN_gene_sets,
  "TRNSCRPT_GN" = trnscrpt_GN_gene_sets,
  "METAB_GN" = metab_GN_refmet_sets
)

FCM_ORA_SEX_STRATIFIED = data.frame()

for(sex_i in c("Female", "Male")){
  FCM_ORA <- lapply(unique(fData(eset)[["ome"]]), function(ome_i) {
    cl <- FCM[[sex_i]]
    eset_i <- eset_list[[sex_i]]
    
    col <- paste0("min_padj_", substr(sex_i, 1L, 1L))
    #we plan on just simplifying and using only the significant features for this analysis. 
    #no filtering if we include all the values. 
    eset_i_sub <- eset_i #[fData(eset_i)[[col]] < 0.05, ]
    keep <- fData(eset_i_sub)[["ome"]] == ome_i
    cluster <- cl$cluster[keep]
    #automatically assigns the cluster with highest membership. 
    
    # Convert between identifiers
    names(cluster) <- fData(eset_i_sub)[["featureName2"]][keep]
    cluster <- cluster[order(cluster, decreasing = FALSE)]
    cluster_list <- split(x = names(cluster),
                          f = factor(cluster,
                                     levels = as.character(unique(cluster))))
    
    background = fData(eset_list[[sex_i]]) %>%
      filter(ome == ome_i) %>% 
      pull(featureName2)
    
    out = data.frame()
    
    for(cluster in seq(1:num_clusters)){
      cluster_res = muscle_ora(input = cluster_list[[cluster]],
                               background = background,
                               database = index_list[[ome_i]])
      cluster_res$cluster_num = cluster
      cluster_res$adj_p_value = p.adjust(cluster_res$p_value, method = "BH")
      out = rbind(out, cluster_res)
    }
    out$ome = ome_i
    return(out)
  }) %>%
    bind_rows() 
  FCM_ORA$sex = sex_i
  FCM_ORA_SEX_STRATIFIED = rbind(FCM_ORA_SEX_STRATIFIED, FCM_ORA)
}

FCM_ORA_SEX_STRATIFIED = FCM_ORA_SEX_STRATIFIED %>% 
  arrange(p_value) %>%
  select(sex, ome, cluster_num, everything()) %>%
  rename(GeneSet = set) %>%
  left_join(., SET_TO_ID, by = "GeneSet")
head(FCM_ORA_SEX_STRATIFIED)
```


```{r camerapr}
index_list <- list(
  "PROT_GN" = prot_GN_gene_sets,
  "TRNSCRPT_GN" = trnscrpt_GN_gene_sets,
  "METAB_GN" = metab_GN_refmet_sets
)

FCM_CAMERA_SEX_STRATIFIED = data.frame()

for(sex_i in c("Female", "Male")){
  FCM_CAMERA <- lapply(unique(fData(eset)[["ome"]]), function(ome_i) {
    cl <- FCM[[sex_i]]
    eset_i <- eset_list[[sex_i]]
    
    col <- paste0("min_padj_", substr(sex_i, 1L, 1L))
    #we plan on just simplifying and using only the significant features for this analysis. 
    #no filtering if we include all the values. 
    eset_i_sub <- eset_i 
    keep <- fData(eset_i_sub)[["ome"]] == ome_i
    
    cluster_membership = cl$membership[keep,]
    # Convert between identifiers
    rownames(cluster_membership) <- fData(eset_i_sub)[["featureName2"]][keep]
    
    camera_res = cameraPR(cluster_membership,
                          index = index_list[[ome_i]]) %>%
      rename(cluster_num = Contrast,
             p_value = PValue,
             adj_p_value = FDR) %>%
      mutate(ome = ome_i)
    return(camera_res)
  }) %>%
    bind_rows() 
  FCM_CAMERA$sex = sex_i
  FCM_CAMERA_SEX_STRATIFIED = rbind(FCM_CAMERA_SEX_STRATIFIED, FCM_CAMERA)
}

FCM_CAMERA_SEX_STRATIFIED = FCM_CAMERA_SEX_STRATIFIED %>% 
  arrange(p_value) %>%
  select(sex, ome, cluster_num, everything()) %>%
  left_join(., SET_TO_ID, by = "GeneSet")
head(FCM_CAMERA_SEX_STRATIFIED)
```


## Plot Top Terms by Cluster

```{r ORA enrichment-bubble-heatmaps, eval=save_plots}
## Bubble heatmaps of tests for gene set localization
for (ome_i in unique(FCM_ORA_SEX_STRATIFIED$ome)) {
  for(sex_i in c("Female", "Male")){
    
    if(sex_i == "Female") desired_clusters = c(2, 3, 8, 11)
    if(sex_i == "Male") desired_clusters = c(1, 4, 7, 12)
    
    x <- FCM_ORA_SEX_STRATIFIED %>%
      filter(ome == ome_i,
             sex == sex_i,
             cluster_num %in% as.factor(desired_clusters))
    if(ome_i == "METAB_GN") x$GeneSetShort = x$GeneSet
    
    # Select the top 6 terms by cluster
    top_terms <- x %>% 
      filter(adj_p_value < 0.05 + 0.05 * (ome_i == "PHOSPHO_GN")) %>% 
      slice_min(p_value, n = 6L, by = cluster_num) %>%
      pull(GeneSetShort) %>% 
      unique()
    
    x <- filter(x, GeneSetShort %in% top_terms) %>% 
      mutate(logP = -log10(p_value))
    
    # Dynamic heatmap height. Each cell of the heatmap is 14 pt, by default
    height <- convertUnit(length(top_terms) * unit(15, "pt"), "in")
    height <- max(as.numeric(height) + 0.2, 4)
    
    # Dynamic heatmap width
    row_label_width <- ComplexHeatmap::max_text_width(x[["GeneSetShort"]])
    row_label_width <- convertUnit(row_label_width, "in")
    
    width <- convertUnit(nlevels(x$cluster_num) * unit(15, "pt"), "in")
    width <- as.numeric(width + row_label_width) + 5
    
    enrichmap(x = x,
              plot_sig_only = FALSE,
              n_top = Inf, # plot all terms
              statistic_column = "logP",
              set_column = "GeneSetShort",
              padj_column = "adj_p_value",
              padj_legend_title = "BH Adjusted\nP-Value",
              contrast_column = "cluster_num",
              heatmap_args = list(
                na_col = "grey95",
                rect_gp = gpar(fill = "white", col = "grey85"),
                cluster_rows = FALSE,
                heatmap_legend_param = list(
                  title = latex2exp::TeX(
                    "$\\bf{$-$log_{10}(P$-$Value)}$")
                ),
                column_names_side = "top"
              ),
              color = c("white", "#503080"),
              filename = file.path("..", "..", "plots", "fuzzy_cmeans", "ORA",
                                   sprintf("%s-%s_select-clusters_cmeans_ORA.png",
                                           ome_i, sex_i)),
              height = height, width = width)
  }
}
```

```{r camera select enrichment-bubble-heatmaps, eval=save_plots}
## Bubble heatmaps of tests for gene set localization
for (ome_i in unique(FCM_CAMERA_SEX_STRATIFIED$ome)) {
  for(sex_i in c("Female", "Male")){
    
    if(sex_i == "Female") desired_clusters = c(2, 3, 8, 11)
    if(sex_i == "Male") desired_clusters = c(1, 4, 7, 12)
    
    x <- FCM_CAMERA_SEX_STRATIFIED %>% 
      filter(ome == ome_i,
             sex == sex_i,
             cluster_num %in% as.factor(desired_clusters))
    if(ome_i == "METAB_GN") x$GeneSetShort = x$GeneSet
    
    # Select the top n terms by cluster
    top_terms <- x %>% 
      filter(adj_p_value < 0.05 + 0.05 * (ome_i == "PHOSPHO_GN")) %>% 
      slice_min(p_value, n = 10L, by = cluster_num) %>%
      pull(GeneSetShort) %>% 
      unique()
    
    x <- filter(x, GeneSetShort %in% top_terms) %>% 
      mutate(logP = -log10(p_value))
    
    # Dynamic heatmap height. Each cell of the heatmap is 14 pt, by default
    height <- convertUnit(length(top_terms) * unit(15, "pt"), "in")
    height <- max(as.numeric(height) + 0.2, 4) + 1
    
    # Dynamic heatmap width
    row_label_width <- ComplexHeatmap::max_text_width(x[["GeneSet"]])
    row_label_width <- convertUnit(row_label_width, "in")
    
    width <- convertUnit(nlevels(x$cluster_num) * unit(15, "pt"), "in")
    width <- as.numeric(width + row_label_width) + 6
    
    enrichmap(x = x,
              plot_sig_only = FALSE,
              n_top = Inf, # plot all terms
              statistic_column = "logP",
              set_column = "GeneSet",
              padj_column = "adj_p_value",
              padj_legend_title = "BH Adjusted\nP-Value",
              contrast_column = "cluster_num",
              heatmap_args = list(
                na_col = "grey95",
                rect_gp = gpar(fill = "white", col = "grey85"),
                cluster_rows = TRUE,
                cluster_columns = FALSE,
                heatmap_legend_param = list(
                  title = latex2exp::TeX(
                    "$\\bf{$-$log_{10}(P$-$Value)}$"),
                  at = c(0, 20),
                  labels = c(0, 20)
                ),
                column_names_side = "top"
              ),
              color = c("white", "#503080"),
              filename = file.path("..", "..", "plots", "fuzzy_cmeans", "cameraPR",
                                   sprintf("%s-%s_select-clusters_cmeans_CAMERA.png",
                                           ome_i, sex_i)),
              height = height, width = width)
  }
}
```

We also want a version without select clusters for CAMERA

```{r camera all enrichment-bubble-heatmaps, eval=save_plots}
## Bubble heatmaps of tests for gene set localization
for (ome_i in unique(FCM_CAMERA_SEX_STRATIFIED$ome)) {
  for(sex_i in c("Female", "Male")){
    x <- FCM_CAMERA_SEX_STRATIFIED %>% 
      filter(ome == ome_i,
             sex == sex_i)
    if(ome_i == "METAB_GN") x$GeneSetShort = x$GeneSet
    
    # Select the top n terms by cluster
    top_terms <- x %>% 
      filter(adj_p_value < 0.05 + 0.05 * (ome_i == "PHOSPHO_GN")) %>% 
      slice_min(p_value, n = 10L, by = cluster_num) %>%
      pull(GeneSetShort) %>% 
      unique()
    
    x <- filter(x, GeneSetShort %in% top_terms) %>% 
      mutate(logP = -log10(p_value))
    
    # Dynamic heatmap height. Each cell of the heatmap is 14 pt, by default
    height <- convertUnit(length(top_terms) * unit(15, "pt"), "in")
    height <- max(as.numeric(height) + 0.2, 4) + 1
    
    # Dynamic heatmap width
    row_label_width <- ComplexHeatmap::max_text_width(x[["GeneSet"]])
    row_label_width <- convertUnit(row_label_width, "in")
    
    width <- convertUnit(nlevels(x$cluster_num) * unit(15, "pt"), "in")
    width <- as.numeric(width + row_label_width) + 6
    
    enrichmap(x = x,
              plot_sig_only = FALSE,
              n_top = Inf, # plot all terms
              statistic_column = "logP",
              set_column = "GeneSet",
              padj_column = "adj_p_value",
              padj_legend_title = "BH Adjusted\nP-Value",
              contrast_column = "cluster_num",
              heatmap_args = list(
                na_col = "grey95",
                rect_gp = gpar(fill = "white", col = "grey85"),
                cluster_rows = TRUE,
                cluster_columns = FALSE,
                heatmap_legend_param = list(
                  title = latex2exp::TeX(
                    "$\\bf{$-$log_{10}(P$-$Value)}$"),
                  at = c(0, 20),
                  labels = c(0, 20)
                ),
                column_names_side = "top"
              ),
              color = c("white", "#503080"),
              filename = file.path("..", "..", "plots", "fuzzy_cmeans", "cameraPR",
                                   sprintf("%s-%s_cmeans_CAMERA.png",
                                           ome_i, sex_i)),
              height = height, width = width)
  }
}
```

The task now is to highlight specific clusters and omes in a way that doesn't have infinite legends.

Current plan is to have M + F up clusters together as columns, then the rows be split by ome.

So there will be 2 heatmaps (one for up, one for down), each with all omes + M/F.

This corresponds to figures 3B, C. We will have a different version below with all the clusters displayed also.
```{r}
FCM_CAMERA_SEX_STRATIFIED =  FCM_CAMERA_SEX_STRATIFIED %>%
  mutate(
    GeneSetShort =
      dplyr::case_when(
        ome == "METAB_GN" ~ GeneSet,
        TRUE ~ GeneSetShort
      )
  )

#up clusters
CAMERA_filtered_up =
  FCM_CAMERA_SEX_STRATIFIED %>%
  filter(
    (sex == "Female" & cluster_num %in% c(2, 3)) |
      (sex == "Male"   & cluster_num %in% c(4, 7))
  ) %>%
  mutate(
    sex_cluster_pattern =
      dplyr::case_when(
        sex == "Female" & cluster_num == 3 ~ "F Continuous Up",
        sex == "Female" & cluster_num == 2 ~ "F Transient Up",
        sex == "Male"   & cluster_num == 7 ~ "M Continuous Up",
        sex == "Male"   & cluster_num == 4 ~ "M Transient Up",
        TRUE ~ NA_character_
      )
  )

CAMERA_filtered_down = FCM_CAMERA_SEX_STRATIFIED %>%
  filter(
    (sex == "Female" & cluster_num %in% c(8, 11)) |
      (sex == "Male"   & cluster_num %in% c(1, 12))
  ) %>%
  mutate(
    sex_cluster_pattern =
      dplyr::case_when(
        sex == "Female" & cluster_num == 11 ~ "F Continuous Down",
        sex == "Female" & cluster_num == 8 ~ "F Transient Down",
        sex == "Male"   & cluster_num == 1 ~ "M Continuous Down",
        sex == "Male"   & cluster_num == 12 ~ "M Transient Down",
        TRUE ~ NA_character_
      )
  )
camera_res = list()
camera_res[["upregulation"]] = CAMERA_filtered_up
camera_res[["downregulation"]] = CAMERA_filtered_down
```

```{r}
cell_size = unit(6, "mm")
ome_cols = c(METAB_GN = "darkgreen", PROT_GN = "steelblue", TRNSCRPT_GN = "orange")
ome_lg = ComplexHeatmap::Legend(
  title = "Omics enrichments",
  at = names(ome_cols),
  legend_gp = grid::gpar(fill = ome_cols)
)

sig_lg = ComplexHeatmap::Legend(
  title = "Significance",
  at = "FDR < 0.05",
  legend_gp = grid::gpar(fill = "grey90", col = NA)
)

sex_cols = sex_cols = c(Female = MotrpacBicQC::sex_cols[["Female"]],
                        Male = MotrpacBicQC::sex_cols[["Male"]]) 
sex_lg = ComplexHeatmap::Legend(
  title = "Sex",
  at = names(sex_cols),
  legend_gp = grid::gpar(fill = sex_cols)
)


for(direction in names(camera_res)){
  camera_select = camera_res[[direction]]
  camera_heatmaps = list()
  for (ome_i in unique(FCM_CAMERA_SEX_STRATIFIED$ome)) {
    x = camera_select %>%
      filter(ome == ome_i)
    # Select the top n terms by cluster
    top_terms <- x %>% 
      filter(adj_p_value < 0.05) %>% 
      slice_min(p_value, n = 3L, by = sex_cluster_pattern) %>%
      pull(GeneSetShort) %>% 
      unique()
    
    x <- filter(x, GeneSetShort %in% top_terms) %>% 
      mutate(logP = -log10(p_value)) %>%
      arrange(sex_cluster_pattern)
    
    statistic_mat = pivot_to_matrix(x, "GeneSetShort", "sex_cluster_pattern", "logP")
    padj_mat = pivot_to_matrix(x, "GeneSetShort", "sex_cluster_pattern", "adj_p_value")

    col_fun = make_col_fun(statistic_mat)
    color_lg = make_color_legend(col_fun)

    is_first_ome = ome_i == unique(FCM_CAMERA_SEX_STRATIFIED$ome)[1]
    camera_heatmaps[[ome_i]] = make_camera_heatmap(
      statistic_mat, padj_mat, col_fun,
      cell_size, sex_cols, ome_cols, ome_i, is_first_ome
    )
  }
  combined_heatmap =  Reduce(`%v%`, camera_heatmaps)
  
  pdf(
    file.path("..", "..", "plots", "fuzzy_cmeans", "cameraPR",
              sprintf("%s_select-clusters_combined_cmeans_CAMERA.pdf", direction)),
    width = 12, height = 10
  )
  
  ComplexHeatmap::draw(
    combined_heatmap,
    heatmap_legend_list = list(color_lg, ome_lg, sig_lg, sex_lg),
    merge_legend = TRUE,
    heatmap_legend_side = "left"
  )
  dev.off()
}
# }
```


Now we repeat but with all clusters instead of just select ones, for a supplement
```{r}
camera_heatmaps_supplement = list()
for (ome_i in unique(FCM_CAMERA_SEX_STRATIFIED$ome)) {
  x = FCM_CAMERA_SEX_STRATIFIED %>%
    filter(ome == ome_i) %>%
    mutate(sex_cluster_pattern = interaction(sex, cluster_num))
  # Select the top n terms by cluster
  top_terms <- x %>% 
    filter(adj_p_value < 0.05) %>% 
    slice_min(p_value, n = 2L, by = cluster_num) %>%
    pull(GeneSetShort) %>% 
    unique()
  
  x <- filter(x, GeneSetShort %in% top_terms) %>% 
    mutate(logP = -log10(p_value)) %>%
    arrange(sex, cluster_num)
  
  statistic_mat = pivot_to_matrix(x, "GeneSetShort", "sex_cluster_pattern", "logP")
  padj_mat = pivot_to_matrix(x, "GeneSetShort", "sex_cluster_pattern", "adj_p_value")

  col_fun = make_col_fun(statistic_mat)
  color_lg = make_color_legend(col_fun)

  is_first_ome = ome_i == unique(FCM_CAMERA_SEX_STRATIFIED$ome)[1]
  camera_heatmaps_supplement[[ome_i]] = make_camera_heatmap(
    statistic_mat, padj_mat, col_fun,
    cell_size, sex_cols, ome_cols, ome_i, is_first_ome
  )
}
combined_heatmap =  Reduce(`%v%`, camera_heatmaps_supplement)

pdf(
  file.path("..", "..", "plots", "fuzzy_cmeans", "cameraPR",
            "select-clusters_combined_cmeans_supplement_CAMERA.pdf"),
  width = 16, height = 30
)

ComplexHeatmap::draw(
  combined_heatmap,
  heatmap_legend_list = list(color_lg, ome_lg, sig_lg, sex_lg),
  merge_legend = TRUE,
  heatmap_legend_side = "left"
)
dev.off()


```
Now based on some of the more interesting pathways that we'd like to expand on, we're going to make some feature heatmaps

```{r}
pathway_ome = "PROT_GN"; da_object = PROT_GN_DA$trained_vs_SED
interested_pathways = c(7550, #respirome
                        4609 #purine deoxyribonucleoside... 
)
pathway_names = SET_TO_ID %>% filter(GeneSetID %in% interested_pathways) 

relevant_pathways = index_list[[pathway_ome]]
filt_index = relevant_pathways[names(relevant_pathways) %in% pathway_names$GeneSet]


lapply(names(filt_index), FUN = function(pathway_name){
  interested_features = filt_index[[pathway_name]]
  ht = feature_heatmap(interested_genes = interested_features,
                       da_results = da_object,
                       filter_nonsig = TRUE,
                       scale = 2,
                       title = "")
  safe_name = gsub("[^A-Za-z0-9_-]", "_", pathway_name)
  pdf_path = file.path(here(), "plots", "fuzzy_cmeans", "heatmaps",
                       paste0("PROT_", safe_name, ".pdf"))
  pdf(pdf_path, height = 9)
  print(ht)
  dev.off()
})

```



# Session Info

```{r}
print(sessionInfo(), locale = FALSE)
```

