---
title: "Fuzzy C-Means Clustering"
author: "Christopher Jin and Tyler Sagendorf"
date: "Last compiled `r format(Sys.time(), '%d %B, %Y')`"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  fig.width = 6, 
  fig.height = 4
)

save_plots <- dir.exists(file.path(here(), "plots"))
```

Documentation:

We present the CAMERA-PR results as the primary results in the manuscript, primarily
because the membership probability is an easy, continuous marker that can be used for CAMERA. 

One large reason I don't present the ORA (e.g. fischer exact for features in cluster) as the primary result is mainly due to concerns with the arbitrary
number of clusters, and the potential relatively small number of features in each cluster, which ORA can be very sensitive to. 

However, I realize just doing the fischer exact is the more common tool in the literature, and so those results are included for comparison purposes.


```{r setup, warning=FALSE, message=FALSE}
# Required packages
library(Biobase)
library(ComplexHeatmap) # max_text_width
library(dplyr)
library(ggplot2)
library(grid) # gpar, unit, convertUnit
library(latex2exp) # TeX
library(Mfuzz) # mfuzz, mestimate, mfuzz.plot2
library(MotrpacRatTraining6moMuscleData)
library(tibble)
library(tidyr)
library(fgsea)
library(TMSig)
library(here)
library(data.table)

num_clusters = 12
source(here::here("Other_Figures", "cmeans", "cmeans_helpers.R"))
```

# Prepare Input

```{r combine-DA-results}
combined_DA <- list(
  "TRNSCRPT_GN" = TRNSCRPT_GN_DA,
  "PROT_GN" = PROT_GN_DA,
  "METAB_GN" = METAB_GN_DA
) %>% 
  lapply(function(x) {
    x[["trained_vs_SED"]] %>% 
      select(contrast, featureName, z, P.Value, adj.P.Val,
             any_of(c("gene_symbol", "metabolite_refmet")))
  }) %>% 
  bind_rows(.id = "ome") %>% 
  mutate(
    contrast = sub(" .*", "", contrast),
    contrast = factor(contrast, levels = unique(contrast)),
    featureName2 = case_when(
      !is.na(gene_symbol) ~ gene_symbol,
      !is.na(metabolite_refmet) ~ metabolite_refmet, 
      TRUE ~ featureName
    )
  ) %>%
  select(
    -c(gene_symbol, metabolite_refmet)
  ) %>% 
  # Select the most extreme z-score by feature
  arrange(
    ome, contrast, desc(abs(z))
  ) %>%
  filter(
    .by = c(ome, contrast),
    !duplicated(featureName2)
  ) %>%
  mutate(
    sex = substr(x = as.character(contrast), 1L, 1L)
  ) %>%
  mutate(
    min_padj_M = ifelse(
      sex == "M", 
      adj.P.Val,
      NA_real_
    ),
    min_padj_F = ifelse(
      sex == "F",
      adj.P.Val,
      NA_real_
    )
  ) %>% 
  mutate(
    .by = c(ome, featureName),
    min_padj_M = ifelse(
      all(is.na(min_padj_M)),
      NA_real_,
      min(min_padj_M, na.rm = TRUE)
    ),
    min_padj_F = ifelse(
      all(is.na(min_padj_F)),
      NA_real_,
      min(min_padj_F, na.rm = TRUE)
    )
  ) %>%
  tidyr::pivot_wider(
    id_cols = c(ome, featureName, featureName2, contains("min_padj")),
    names_from = contrast,
    values_from = z
  ) %>% 
  # Complete data for clustering
  filter(
    complete.cases(
      across(.cols = ends_with("W", ignore.case = FALSE))
    )
  )

head(combined_DA)
```

```{r create-eset-list}
expr_mat <- combined_DA %>% 
  mutate(row_names = paste(ome, featureName)) %>% 
  tibble::column_to_rownames("row_names")

f_data <- expr_mat %>% 
  select(ome, featureName, featureName2, contains("min_padj"))

expr_mat <- select(expr_mat, -colnames(f_data)) %>%
  as.matrix()

eset <- ExpressionSet(
  assayData = expr_mat, 
  featureData = AnnotatedDataFrame(f_data)
)

eset # summary of ExpressionSet object
```

```{r eval=TRUE}
# Fuzzy C-Means (FCM) clustering ----
# Split by sex
eset_list <- lapply(c("F", "M"), function(sex_i) {
  eset[, grepl(sex_i, colnames(eset))]
}) %>% 
  setNames(c("Female", "Male"))

FCM = list()
for(sex_i in c("Female", "Male")){
  eset_i <- eset_list[[sex_i]]
  
  col <- paste0("min_padj_", substr(sex_i, 1L, 1L))
  #we plan on just simplifying and using only the significant 
  #features for this analysis.
  
  #if we are using 12 clusters -> we don't do a p-value threshold.
  #if we are using a smaller number of clusters -> we do filter. 
  eset_i_sub <- eset_i #[fData(eset_i)[[col]] < 0.05, ]
  
  #add the 0 columns representing the SED groups
  non_scaled = exprs(eset_i_sub)
  # scaled = t(scale(t(exprs(eset_i))))
  esets_zeros = cbind(rep(0, nrow(non_scaled)),
                      non_scaled)
  # message(paste0("Sig Feat: ",sex_i,  nrow(esets_zeros)))
  colnames(esets_zeros) = c("SED", colnames(eset_i))
  
  eset_cmeans <- ExpressionSet(
    assayData = esets_zeros, 
    featureData = AnnotatedDataFrame(fData(eset_i_sub))
  )
  
  set.seed(0L)
  # Calculate membership probabilities for all features
  fcm_2 <- mfuzz(
    eset = eset_cmeans,
    centers = num_clusters,
    m = 2,
    iter.max = 1L # more iterations will flatten clusters
  ) %>% 
    reorder_clusters()
  
  FCM[[sex_i]] = fcm_2
  
  file_name = file.path(here(), "plots/fuzzy_cmeans", paste0(sex_i, ".png"))
  png(file = file_name,
      width = 14, height = 12, units = "in", res = 1200)
  Mfuzz::mfuzz.plot2(eset_cmeans,
                     mfrow = c(3,4),
                     ylim.set = c(-4,4),
                     fcm_2,
                     min.mem = 0.3,
                     xlab = paste(sex_i, "Clusters"),
                     ylab = "",
                     centre = TRUE,
                     time.labels = colnames(esets_zeros),
                     col.axis = "black",
                     cex.lab = 1.5,
                     cex.axis = 1.5, #axis text size,
                     cex.main = 1.5, #main title text size
                     # single = i,
                     x11 = FALSE)
  dev.off()
}

```

```{r}
index_list <- list(
  "PROT_GN" = prot_GN_gene_sets,
  "TRNSCRPT_GN" = trnscrpt_GN_gene_sets,
  "METAB_GN" = metab_GN_refmet_sets
)
```

## Overrepresentation Analysis (ORA)

For each cluster, tests whether gene set members are over-represented among features assigned to that cluster via Fisher's exact test on hard cluster assignments.

```{r ORA}
FCM_ORA_SEX_STRATIFIED <- run_fcm_ora(eset, eset_list, FCM, index_list, num_clusters) %>%
  arrange(p_value) %>%
  select(sex, ome, cluster_num, everything()) %>%
  rename(GeneSet=set) %>%
  left_join(SET_TO_ID, by="GeneSet")
head(FCM_ORA_SEX_STRATIFIED)
```

## CAMERA-PR

Tests gene set enrichment using continuous membership probabilities as pre-ranked statistics, detecting whether set members tend to have higher membership to each cluster than background.

```{r camerapr}
FCM_CAMERA_SEX_STRATIFIED <- run_fcm_gst(eset, eset_list, FCM, index_list,
  function(membership_mat, gene_sets) {
    cameraPR(membership_mat, index=gene_sets) %>%
      rename(cluster_num=Contrast, p_value=PValue, adj_p_value=FDR)
  }
) %>%
  arrange(p_value) %>%
  select(sex, ome, cluster_num, everything()) %>%
  left_join(SET_TO_ID, by="GeneSet")
head(FCM_CAMERA_SEX_STRATIFIED)
```

## FGSEA

Ranks features within each cluster by membership probability and tests whether gene set members are enriched at the top of the ranking.

```{r fgsea, eval = FALSE}
FCM_FGSEA_SEX_STRATIFIED <- run_fcm_gst(eset, eset_list, FCM, index_list,
  function(membership_mat, gene_sets) {
    lapply(colnames(membership_mat), function(cluster_i) {
      stats_i <- na.omit(membership_mat[, cluster_i])
      set.seed(9001)
      fgsea(
        pathways=gene_sets,
        stats=stats_i,
        scoreType = "pos",
        # precision of p-values
        eps=0,
        nproc=4,
        # Number of permutations. Affects precision of NES and p-values.
        nPermSimple=1e4
      ) %>%
        mutate(cluster_num=cluster_i)
    }) %>%
      bind_rows() %>%
      rename(GeneSet=pathway, NGenesRat=size,
             PValue=pval, PAdj=padj,
             LeadingEdge=leadingEdge) %>%
      left_join(SET_TO_ID, by="GeneSet") %>%
      mutate(
        Direction=ifelse(NES >= 0, "Up", "Down"),
        PAdj=p.adjust(PValue, method="BH"),
        # NRatio is a measure of confidence that the set being tested is what is
        # described by GeneSet
        NRatio=round(NGenesRat / NGenesHuman, digits=3),
        Direction=factor(Direction, levels=c("Up", "Down"))
      ) %>%
      select(cluster_num, GeneSetID, GeneSet, GeneSetShort,
             NGenesRat, NGenesHuman, NRatio,
             Direction, NES,
             PValue, PAdj, LeadingEdge) %>%
      arrange(cluster_num, PValue, NES)
  }
) %>%
  arrange(PValue) %>%
  select(sex, ome, cluster_num, everything())

FCM_FGSEA_SEX_STRATIFIED = FCM_FGSEA_SEX_STRATIFIED %>%
  rename(
    adj_p_value = PAdj,
    p_value = PValue
  )
head(FCM_FGSEA_SEX_STRATIFIED)
```


## Plot Top Terms by Cluster

All clusters (supplement version):

```{r camera all enrichment-bubble-heatmaps, eval=save_plots}
for (ome_i in names(ome_cols)) {
  for (sex_i in c("Female", "Male")) {
    x <- FCM_CAMERA_SEX_STRATIFIED %>%
      filter(ome==ome_i, sex==sex_i)
    plot_enrichmap_clusters(
      x,
      set_col="GeneSet",
      n_top=10L,
      filename=file.path(here(), "plots", "fuzzy_cmeans", "cameraPR",
                         sprintf("%s-%s_cmeans_CAMERA.png", ome_i, sex_i))
    )
  }
}
```

The task now is to highlight specific clusters and omes in a way that doesn't have infinite legends.

Current plan is to have M + F up clusters together as columns, then the rows be split by ome.

So there will be 2 heatmaps (one for up, one for down), each with all omes + M/F.

This corresponds to figures 3B, C. We will have a different version below with all the clusters displayed also.
```{r}
FCM_CAMERA_SEX_STRATIFIED =  FCM_CAMERA_SEX_STRATIFIED %>%
  mutate(
    GeneSetShort =
      dplyr::case_when(
        ome == "METAB_GN" ~ GeneSet,
        TRUE ~ GeneSetShort
      )
  )

#up clusters
CAMERA_filtered_up =
  FCM_CAMERA_SEX_STRATIFIED %>%
  filter(
    (sex == "Female" & cluster_num %in% c(2, 3)) |
      (sex == "Male"   & cluster_num %in% c(4, 7))
  ) %>%
  mutate(
    sex_cluster_pattern =
      dplyr::case_when(
        sex == "Female" & cluster_num == 3 ~ "F Continuous Up",
        sex == "Female" & cluster_num == 2 ~ "F Transient Up",
        sex == "Male"   & cluster_num == 7 ~ "M Continuous Up",
        sex == "Male"   & cluster_num == 4 ~ "M Transient Up",
        TRUE ~ NA_character_
      )
  )

CAMERA_filtered_down = FCM_CAMERA_SEX_STRATIFIED %>%
  filter(
    (sex == "Female" & cluster_num %in% c(8, 11)) |
      (sex == "Male"   & cluster_num %in% c(1, 12))
  ) %>%
  mutate(
    sex_cluster_pattern =
      dplyr::case_when(
        sex == "Female" & cluster_num == 11 ~ "F Continuous Down",
        sex == "Female" & cluster_num == 8 ~ "F Transient Down",
        sex == "Male"   & cluster_num == 1 ~ "M Continuous Down",
        sex == "Male"   & cluster_num == 12 ~ "M Transient Down",
        TRUE ~ NA_character_
      )
  )
camera_res = list()
camera_res[["upregulation"]] = CAMERA_filtered_up
camera_res[["downregulation"]] = CAMERA_filtered_down
```

```{r}
cell_size = unit(6, "mm")
ome_cols = c(TRNSCRPT_GN = "#377EB8", PROT_GN = "#4DAF4A", METAB_GN = "#E41A1C")
ome_lg = ComplexHeatmap::Legend(
  title = "Omics enrichments",
  at = names(ome_cols),
  legend_gp = grid::gpar(fill = ome_cols)
)

sig_lg = ComplexHeatmap::Legend(
  title = "Significance",
  at = "FDR < 0.05",
  legend_gp = grid::gpar(fill = "grey90", col = NA)
)

sex_cols = sex_cols = c(Female = MotrpacBicQC::sex_cols[["Female"]],
                        Male = MotrpacBicQC::sex_cols[["Male"]]) 
sex_lg = ComplexHeatmap::Legend(
  title = "Sex",
  at = names(sex_cols),
  legend_gp = grid::gpar(fill = sex_cols)
)


for(direction in names(camera_res)){
  camera_select = camera_res[[direction]]
  camera_heatmaps = list()
  for (ome_i in names(ome_cols)) {
    x = camera_select %>%
      filter(ome == ome_i)
    # Select the top n terms by cluster
    top_terms <- x %>% 
      filter(adj_p_value < 0.05) %>% 
      slice_min(p_value, n = 3L, by = sex_cluster_pattern) %>%
      pull(GeneSetShort) %>% 
      unique()
    
    x <- filter(x, GeneSetShort %in% top_terms) %>% 
      mutate(logP = -log10(p_value)) %>%
      arrange(sex_cluster_pattern)
    
    statistic_mat = pivot_to_matrix(x, "GeneSetShort", "sex_cluster_pattern", "logP")
    padj_mat = pivot_to_matrix(x, "GeneSetShort", "sex_cluster_pattern", "adj_p_value")

    col_fun = make_col_fun(statistic_mat)
    color_lg = make_color_legend(col_fun)

    is_first_ome = ome_i == names(ome_cols)[1]
    camera_heatmaps[[ome_i]] = make_camera_heatmap(
      statistic_mat, padj_mat, col_fun,
      cell_size, sex_cols, ome_cols, ome_i, is_first_ome
    )
  }
  combined_heatmap =  Reduce(`%v%`, camera_heatmaps)
  
  pdf(
    file.path(here(), "plots", "fuzzy_cmeans", "cameraPR",
              sprintf("%s_select-clusters_combined_cmeans_CAMERA.pdf", direction)),
    width = 12, height = 10
  )
  
  ComplexHeatmap::draw(
    combined_heatmap,
    heatmap_legend_list = list(color_lg, ome_lg, sig_lg, sex_lg),
    merge_legend = TRUE,
    heatmap_legend_side = "left"
  )
  dev.off()
}
# }
```


Now we repeat but with all clusters instead of just select ones, for a supplement
```{r}
camera_heatmaps_supplement = list()
for (ome_i in names(ome_cols)) {
  x = FCM_CAMERA_SEX_STRATIFIED %>%
    filter(ome == ome_i) %>%
    mutate(sex_cluster_pattern = interaction(sex, cluster_num))
  # Select the top n terms by cluster
  top_terms <- x %>% 
    filter(adj_p_value < 0.05) %>% 
    slice_min(p_value, n = 2L, by = cluster_num) %>%
    pull(GeneSetShort) %>% 
    unique()
  
  x <- filter(x, GeneSetShort %in% top_terms) %>% 
    mutate(logP = -log10(p_value)) %>%
    arrange(sex, cluster_num)
  
  statistic_mat = pivot_to_matrix(x, "GeneSetShort", "sex_cluster_pattern", "logP")
  padj_mat = pivot_to_matrix(x, "GeneSetShort", "sex_cluster_pattern", "adj_p_value")

  col_fun = make_col_fun(statistic_mat)
  color_lg = make_color_legend(col_fun)

  is_first_ome = ome_i == names(ome_cols)[1]
  camera_heatmaps_supplement[[ome_i]] = make_camera_heatmap(
    statistic_mat, padj_mat, col_fun,
    cell_size, sex_cols, ome_cols, ome_i, is_first_ome
  )
}
combined_heatmap =  Reduce(`%v%`, camera_heatmaps_supplement)

pdf(
  file.path("..", "..", "plots", "fuzzy_cmeans", "cameraPR",
            "select-clusters_combined_cmeans_supplement_CAMERA.pdf"),
  width = 16, height = 30
)

ComplexHeatmap::draw(
  combined_heatmap,
  heatmap_legend_list = list(color_lg, ome_lg, sig_lg, sex_lg),
  merge_legend = TRUE,
  heatmap_legend_side = "left"
)
dev.off()


```
Now based on some of the more interesting pathways that we'd like to expand on, we're going to make some feature heatmaps

```{r}
pathway_ome = "PROT_GN"; da_object = PROT_GN_DA$trained_vs_SED
interested_pathways = c(7550, #respirome
                        4609 #purine deoxyribonucleoside... 
)
pathway_names = SET_TO_ID %>% filter(GeneSetID %in% interested_pathways) 

relevant_pathways = index_list[[pathway_ome]]
filt_index = relevant_pathways[names(relevant_pathways) %in% pathway_names$GeneSet]


lapply(names(filt_index), FUN = function(pathway_name){
  interested_features = filt_index[[pathway_name]]
  ht = feature_heatmap(interested_genes = interested_features,
                       da_results = da_object,
                       filter_nonsig = TRUE,
                       scale = 2,
                       title = "")
  safe_name = gsub("[^A-Za-z0-9_-]", "_", pathway_name)
  pdf_path = file.path(here(), "plots", "fuzzy_cmeans", "heatmaps",
                       paste0("PROT_", safe_name, ".pdf"))
  pdf(pdf_path, height = 9)
  print(ht)
  dev.off()
})

```


# Session Info

```{r}
print(sessionInfo(), locale = FALSE)
```

