---
title: "Epi vs Motrpac RNAseq (FGSEA)"
author: "Akshay Bareja, Christopher Jin"
date: Mar 12, 2026
toc: true
number-sections: true
format:
  html:
    df-print: paged
    code-fold: true
    code-tools: true
    embed-resources: true
execute:
  cache: true
editor: source
---

```{r}
#| label: load-packages
#| include: false

library(tidyverse)
library(readxl)
library(limma)
library(edgeR)
library(fgsea)
library(DT)
library(pheatmap)
library(MotrpacRatTraining6moData)
library(TMSig)
library(GEOquery)
library(plotly)
library(patchwork)
library(ggrepel)
library(conflicted)

devtools::load_all()
source(file.path(here::here(), "Other_Figures", "human_comparison", "human_comparison_helpers.R"))
conflicts_prefer(base::setdiff)
```

## Background

This document contains the results for gene-level differential expression analysis done for the [EpiTrain cohort](https://www.tandfonline.com/doi/10.4161/15592294.2014.982445?url_ver=Z39.88-2003&rfr_id=ori:rid:crossref.org&rfr_dat=cr_pub%20%200pubmed#d1e738) and Motrpac. EpiTrain gene level counts were downloaded from [GEO](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE60590).

For EpiTrain, post-versus-pre exercise intervention gene expression changes were evaluated by fitting mixed-effects models that used Timepoint, Dominant leg, and BMI as fixed effect covariates (with Timepoint being the factor interest), and subject treated as a random effect.

For Motrpac, the comparison used was 8 week versus Sedentary.

In both cases, analyses were split by gender/sex.

Enrichment analysis was performed using `fgsea`.

## Differential Expression Analysis

### EpiTrain

```{r}
#| label: epitrain-de
#| echo: false
#| message: false
#| warning: false

epitrain <- load_epitrain_de()
efit_vfit <- epitrain$efit_vfit
annot <- epitrain$annot
akshay_gse60590_f_results <- epitrain$female
akshay_gse60590_m_results <- epitrain$male

akshay_gse60590_f_results %>%
  dplyr::mutate(across(where(is.numeric), ~round(.x, 3))) %>%
  DT::datatable(caption = "EpiTrain DE results (Female)")

akshay_gse60590_m_results %>%
  dplyr::mutate(across(where(is.numeric), ~round(.x, 3))) %>%
  DT::datatable(caption = "EpiTrain DE results (Male)")
```

### MotrPac

```{r}
#| label: motrpac-de
#| echo: false
#| message: false
#| warning: false

motrpac <- load_motrpac_de()
motr_female <- motrpac$female
motr_male <- motrpac$male

motr_female %>%
  data.frame() %>%
  dplyr::mutate(across(where(is.numeric), ~round(.x, 3))) %>%
  dplyr::arrange(adj.P.Val) %>%
  DT::datatable(caption = "MotrPac DE results (Female)")

motr_male %>%
  data.frame() %>%
  dplyr::mutate(across(where(is.numeric), ~round(.x, 3))) %>%
  dplyr::arrange(adj.P.Val) %>%
  DT::datatable(caption = "MotrPac DE results (Male)")
```

## Enrichment Analysis

### EpiTrain

```{r}
#| label: epitrain-ea
#| echo: false
#| message: false
#| warning: false

# Use the "human_gene_sets" object to be consistent with paper.
# see "data-raw/gene_sets.R" to see how these were created.

# process gene sets for the epi project
background_epi <- unique(akshay_gse60590_f_results$Symbol)

epi_gene_sets <- TMSig::filterSets(x = human_gene_sets,
                                   background = background_epi,
                                   min_size = 5L,
                                   max_size = length(background_epi) - 1L)
# length(epi_gene_sets) #8787
clusterDF_epi <- clusterSets(x = epi_gene_sets,
                             type = "jaccard",
                             cutoff = 0.85,
                             method = "complete",
                             h = 0.9)
overlapProp_epi <- nrow(clusterDF_epi) / length(background_epi)

clusterDF_epi$overlap_prop <- overlapProp_epi[clusterDF_epi$set]
clusterDF_epi$n_char <- nchar(clusterDF_epi$set)

o_epi <- with(clusterDF_epi,
              order(cluster, set_size, overlap_prop, n_char, set,
                    decreasing = c(FALSE, TRUE, TRUE, FALSE, TRUE),
                    method = "radix"))
clusterDF_epi <- clusterDF_epi[o_epi, ]

# Subset filtered gene sets
human_gene_sets_epi_filt <- epi_gene_sets[with(clusterDF_epi, set[!duplicated(cluster)])]
# length(human_gene_sets_epi_filt) - 8496

# Build z-score matrix from moderated t-statistics (same mapping used for cameraPR)
modz_epi <- with(efit_vfit, limma::zscoreT(x = t, df = df.total)) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("GeneID") %>%
  dplyr::mutate(GeneID = as.integer(GeneID)) %>%
  dplyr::left_join(annot[, 1:2], dplyr::join_by(GeneID)) %>%
  dplyr::select(-GeneID) %>%
  tibble::column_to_rownames("Symbol") %>%
  dplyr::mutate(across(everything(), as.numeric)) %>%
  as.matrix()

# Run FGSEA per sex contrast
set.seed(9001)
epi_fgsea_f <- fgsea(
  pathways = human_gene_sets_epi_filt,
  stats = na.omit(modz_epi[, "women_comp"]),
  eps = 0,
  nproc = 4,
  nPermSimple = 1e4
)

epi_fgsea_m <- fgsea(
  pathways = human_gene_sets_epi_filt,
  stats = na.omit(modz_epi[, "men_comp"]),
  eps = 0,
  nproc = 4,
  nPermSimple = 1e4
)

# present results
epi_fgsea_f %>%
  dplyr::filter(str_detect(pathway, "^GO")) %>%
  dplyr::arrange(pval) %>%
  dplyr::select(-leadingEdge) %>%
  dplyr::mutate(across(where(is.numeric), ~round(.x, 3))) %>%
  DT::datatable(caption = "EpiTrain Female (FGSEA results)")

epi_fgsea_m %>%
  dplyr::filter(str_detect(pathway, "^GO")) %>%
  dplyr::arrange(pval) %>%
  dplyr::select(-leadingEdge) %>%
  dplyr::mutate(across(where(is.numeric), ~round(.x, 3))) %>%
  DT::datatable(caption = "EpiTrain Male (FGSEA results)")
```

### MotrPac

```{r}
#| label: motrpac-ea
#| echo: false
#| message: false
#| warning: false

# see: vignettes/TRNSCRPT GST for more information on how
# gene set testing was done.
res_motr_f <- TRNSCRPT_GN_FGSEA$trained_vs_SED %>%
  dplyr::filter(Contrast == "F_8W - F_SED")

# male
res_motr_m <- TRNSCRPT_GN_FGSEA$trained_vs_SED %>%
  dplyr::filter(Contrast == "M_8W - M_SED")

# present results
res_motr_f %>%
  dplyr::filter(str_detect(GeneSet, "^GO")) %>%
  dplyr::arrange(PValue) %>%
  dplyr::select(-LeadingEdge) %>%
  dplyr::mutate(across(where(is.numeric), ~round(.x, 3))) %>%
  DT::datatable(caption = "MotrPac Female (FGSEA results)")

res_motr_m %>%
  dplyr::filter(str_detect(GeneSet, "^GO")) %>%
  dplyr::arrange(PValue) %>%
  dplyr::select(-LeadingEdge) %>%
  dplyr::mutate(across(where(is.numeric), ~round(.x, 3))) %>%
  DT::datatable(caption = "MotrPac Male (FGSEA results)")
```

## EpiTrain versus MotrPac

### Differential Expression Analysis

Comparing logFC between EpiTrain and MoTrPAC for the top 1000 MoTrPAC features by adjusted p-value, restricted to sex-matched comparisons (female–female and male–male). The female comparison shows a stronger correlation than the male comparison, suggesting greater conservation of the female transcriptomic response to endurance training across species.

```{r}
#| label: epi-vs-motr-de
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "EpiTrain versus MotrPac (DE results, sex-matched comparisons)"
#| column: page
#| fig-width: 14
#| fig-height: 10

# Join EpiTrain and MoTrPAC DE results for one sex.
# motr_top: MoTrPAC results pre-filtered to the top N features.
join_de <- function(epi, motr_top) {
  dplyr::inner_join(epi, motr_top, dplyr::join_by(Symbol), suffix=c("_epi", "_motr"))
}

# Scatter plot of logFC values with the top n_label MoTrPAC features labeled.
scatter_panel <- function(df, sex_label, cor) {
  df %>%
    ggplot(aes(logFC_motr, logFC_epi, label=Symbol)) +
    geom_point(alpha=0.6) +
    geom_abline(linetype="dashed") +
    ggrepel::geom_label_repel(
      data=~dplyr::slice_min(.x, adj.P.Val_motr, n=10),
      size=3.5, color="red", segment.color="red",
      fill=alpha("white", 0.85), label.size=0.2,
      min.segment.length=0, max.overlaps=Inf,
      point.padding=2, box.padding=0.5, force=8, force_pull=0.3, seed=42
    ) +
    theme_bw() +
    labs(title=sprintf("%s: EpiTrain vs MoTrPAC", sex_label),
         subtitle=sprintf("Pearson r = %.3f, p = %.2e", cor$estimate, cor$p.value),
         x="MoTrPAC logFC", y="EpiTrain logFC")
}

# Dot plot of top concordant (smallest |z_diff|, from top 100 MoTrPAC features)
# and top discordant (largest |z_diff|, from all significant features) genes.
# z-scores are computed within the significant feature set to put both datasets
# on a common scale before computing |z_motr - z_epi|.
feature_panel <- function(df, sex_label, n=6) {
  sig <- df %>%
    dplyr::filter(adj.P.Val_motr < 0.05) %>%
    dplyr::mutate(
      z_motr = scale(logFC_motr)[, 1],
      z_epi  = scale(logFC_epi)[, 1],
      z_diff = abs(z_motr - z_epi)
    )
  concordant <- sig %>%
    dplyr::slice_min(adj.P.Val_motr, n=100, with_ties=FALSE) %>%
    dplyr::slice_min(z_diff, n=n, with_ties=FALSE) %>%
    dplyr::mutate(type="Concordant")
  discordant <- sig %>%
    dplyr::slice_max(z_diff, n=n, with_ties=FALSE) %>%
    dplyr::mutate(type="Discordant")
  dplyr::bind_rows(concordant, discordant) %>%
    dplyr::select(Symbol, logFC_motr, logFC_epi, type) %>%
    tidyr::pivot_longer(cols=c(logFC_motr, logFC_epi),
                        names_to="Dataset", values_to="logFC") %>%
    dplyr::mutate(Dataset=dplyr::recode(Dataset, logFC_motr="MoTrPAC", logFC_epi="EpiTrain")) %>%
    ggplot(aes(logFC, reorder(Symbol, logFC), col=Dataset, shape=Dataset)) +
    geom_point(size=3) +
    geom_vline(xintercept=0, linetype="dashed", color="grey50") +
    facet_wrap(~type, scales="free_y", ncol=1) +
    scale_color_manual(values=c(MoTrPAC="#E41A1C", EpiTrain="#377EB8")) +
    theme_bw() +
    theme(strip.text=element_text(face="bold")) +
    labs(title=sprintf("%s concordant & discordant genes", sex_label),
         x="logFC", y=NULL, col=NULL, shape=NULL)
}

# ── build panels ────────────────────────────────────────────────────────────
epi_vs_motr_de_f <- join_de(akshay_gse60590_f_results,
                             dplyr::slice_min(motr_female, adj.P.Val, n=1000))
epi_vs_motr_de_m <- join_de(akshay_gse60590_m_results,
                             dplyr::slice_min(motr_male, adj.P.Val, n=1000))

cor_f <- cor.test(epi_vs_motr_de_f$logFC_motr, epi_vs_motr_de_f$logFC_epi, method="pearson")
cor_m <- cor.test(epi_vs_motr_de_m$logFC_motr, epi_vs_motr_de_m$logFC_epi, method="pearson")

panel_a <- scatter_panel(epi_vs_motr_de_f, "Female", cor_f)
panel_b <- feature_panel(epi_vs_motr_de_f, "Female")
panel_c <- scatter_panel(epi_vs_motr_de_m, "Male", cor_m)
panel_d <- feature_panel(epi_vs_motr_de_m, "Male")

# Combine: (A scatter | B features) / (C scatter | D features)
de_combined_plot <- (panel_a + panel_b) / (panel_c + panel_d) +
  patchwork::plot_layout(widths=c(2, 1)) +
  patchwork::plot_annotation(tag_levels="A")

de_combined_plot

ggsave(filename="figures/feature_level_epitrain_plot_for_manuscript.png",
       plot=de_combined_plot, width=14, height=10)
```



## Figures for manuscript (old)

```{r, eval=FALSE}
#| label: manuscript-figures (old)
#| echo: false
#| message: false
#| warning: false

# Spearman correlation coefficients
cor(epi_v_motr_fgsea$NES_f_epi, epi_v_motr_fgsea$NES_f_motr, method = "spearman") # 0.370
cor(epi_v_motr_fgsea$NES_f_epi, epi_v_motr_fgsea$NES_m_motr, method = "spearman") # 0.404
cor(epi_v_motr_fgsea$NES_m_epi, epi_v_motr_fgsea$NES_f_motr, method = "spearman") # 0.366
cor(epi_v_motr_fgsea$NES_m_epi, epi_v_motr_fgsea$NES_m_motr, method = "spearman") # -0.173

# p-values
cor.test(epi_v_motr_fgsea$NES_f_epi, epi_v_motr_fgsea$NES_f_motr, method = "spearman") #p-value < 2.2e-16
cor.test(epi_v_motr_fgsea$NES_f_epi, epi_v_motr_fgsea$NES_m_motr, method = "spearman") #p-value 0.04456
cor.test(epi_v_motr_fgsea$NES_m_epi, epi_v_motr_fgsea$NES_f_motr, method = "spearman") #p-value < 2.2e-16
cor.test(epi_v_motr_fgsea$NES_m_epi, epi_v_motr_fgsea$NES_m_motr, method = "spearman") #p-value < 2.2e-16

# What are the pathways with highest average NES across all comparisons?
top_pathways <- epi_v_motr_fgsea %>%
  dplyr::rowwise() %>%
  dplyr::mutate(nes_avg = mean(c_across(starts_with("NES")), na.rm = TRUE)) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(desc(nes_avg)) %>%
  dplyr::slice_max(nes_avg, n = 50) %>%
  dplyr::pull(GeneSet)

# Epi F vs Motr F
f_v_f <- make_zscore_scatter(epi_v_motr_fgsea,
  x_col = "NES_f_epi", y_col = "NES_f_motr",
  top_pathways = top_pathways,
  title = "EpiTrain Female vs MoTrPAC Female",
  subtitle = "Spearman's rho: 0.370\np-value < 2.2e-16",
  x_lab = "EpiTrain Female (NES)",
  y_lab = "MoTrPAC Female (NES)")

# Epi F vs Motr M
f_v_m <- make_zscore_scatter(epi_v_motr_fgsea,
  x_col = "NES_f_epi", y_col = "NES_m_motr",
  top_pathways = top_pathways,
  title = "EpiTrain Female vs MoTrPAC Male",
  subtitle = "Spearman's rho: 0.404\np-value = 0.045",
  x_lab = "EpiTrain Female (NES)",
  y_lab = "MoTrPAC Male (NES)")

# Epi M vs Motr F
m_v_f <- make_zscore_scatter(epi_v_motr_fgsea,
  x_col = "NES_m_epi", y_col = "NES_f_motr",
  top_pathways = top_pathways,
  title = "EpiTrain Male vs MoTrPAC Female",
  subtitle = "Spearman's rho: 0.366\np-value < 2.2e-16",
  x_lab = "EpiTrain Male (NES)",
  y_lab = "MoTrPAC Female (NES)")

# Epi M vs Motr M
m_v_m <- make_zscore_scatter(epi_v_motr_fgsea,
  x_col = "NES_m_epi", y_col = "NES_m_motr",
  top_pathways = top_pathways,
  title = "EpiTrain Male vs MoTrPAC Male",
  subtitle = "Spearman's rho: -0.173\np-value < 2.2e-16",
  x_lab = "EpiTrain Male (NES)",
  y_lab = "MoTrPAC Male (NES)")

manuscript_plot <- (f_v_f + f_v_m) /
  (m_v_f + m_v_m) +
  patchwork::plot_annotation(tag_levels = "A")

manuscript_plot

ggsave(filename = "figures/fgsea_plot_for_manuscript.png",
       plot = manuscript_plot,
       width = 15,
       height = 15)
```

Figure caption

Figure 6. Comparison of transcriptomic responses to exercise in humans and rats at the pathway level. Comparison of FGSEA results for (A) EpiTrain Female versus MoTrPAC Female; (B) EpiTrain Female versus MoTrPAC Male; (C) EpiTrain Male versus MoTrPAC Female; (D) EpiTrain Male versus MoTrPAC Male. Points highlighted in red indicate gene sets with the highest average NES across all comparisons (see text for details).



