---
title: "Epi vs Motrpac RNAseq"
author: "Akshay Bareja"
date: 11/07/2025
toc: true
number-sections: true
format: 
  html:
    df-print: paged
    code-fold: true
    code-tools: true
    embed-resources: true
execute:
  cache: true
editor: visual
---

```{r}
#| label: load-packages
#| include: false

library(tidyverse)
library(readxl)
library(limma)
library(edgeR)
library(DT)
library(pheatmap)
library(MotrpacRatTraining6moData)
library(TMSig)
library(GEOquery)
library(plotly)
library(patchwork)
library(ggside)
library(ggrepel)
```

## Background

This document contains the results for gene-level differential expression analysis done for the [EpiTrain cohort](https://www.tandfonline.com/doi/10.4161/15592294.2014.982445?url_ver=Z39.88-2003&rfr_id=ori:rid:crossref.org&rfr_dat=cr_pub%20%200pubmed#d1e738) and Motrpac. EpiTrain gene level counts were downloaded from [GEO](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE60590).

For EpiTrain, post-versus-pre exercise intervention gene expression changes were evaluated by fitting mixed-effects models that used Timepoint, Dominant leg, and BMI as fixed effect covariates (with Timepoint being the factor interest), and subject treated as a random effect.

For Motrpac, the comparison used was 8 week versus Sedentary.

In both cases, analyses were split by gender/sex.

Enrichment analysis was performed using `cameraPR`.

## Differential Expression Analysis

### EpiTrain

```{r}
#| label: epitrain-de
#| echo: false
#| message: false
#| warning: false
## differential expression analysis

# get metadata
gds <- getGEO("GSE60590")
GSE60590_meta <- pData(phenoData(gds[[1]]))

# get count data
# load counts table from GEO
urld <- "https://www.ncbi.nlm.nih.gov/geo/download/?format=file&type=rnaseq_counts"
path <- paste(urld, "acc=GSE60590", "file=GSE60590_raw_counts_GRCh38.p13_NCBI.tsv.gz", sep="&");
tbl <- as.matrix(data.table::fread(path, header=T, colClasses="integer"), rownames="GeneID")

# load gene annotations 
apath <- paste(urld, "type=rnaseq_counts", "file=Human.GRCh38.p13.annot.tsv.gz", sep="&")
annot <- data.table::fread(apath, header=T, quote="", stringsAsFactors=F, data.table=F)
rownames(annot) <- annot$GeneID


# import additional metadata from Malene
EpiTrain_summary_clinical_data <- read_excel("EpiTrain_summary_clinical_data.xlsx", 
    sheet = "Period 1")
# Make sure to compare with `sample_metadata`

add_meta <- EpiTrain_summary_clinical_data %>% 
              dplyr::select(Subject, BMI, `Trained leg`) %>% 
              dplyr::mutate(Subject = as.character(Subject)) %>%
              dplyr::mutate(dominant_leg = if_else(str_detect(`Trained leg`,
                                                          "\\(D\\)$"),
                                               TRUE, FALSE)) %>% 
              dplyr::rename(sample_id = Subject)

sample_metadata <- GSE60590_meta %>% 
  tibble::rownames_to_column("subject") %>% 
  dplyr::filter(subject %in% colnames(tbl)) %>% # only keep subjects who have count data
  dplyr::select(subject, characteristics_ch1.1:characteristics_ch1.5) %>% 
  dplyr::rename(gender = characteristics_ch1.1,
                leg = characteristics_ch1.2,
                exercise_status = characteristics_ch1.3,
                timepoint = characteristics_ch1.4,
                sample_id = characteristics_ch1.5) %>% 
  dplyr::mutate(gender = str_remove(gender, "^gender: "),
                leg = str_remove(leg, "^biopsy location: "),
                exercise_status = str_remove(exercise_status, "^exercise status: "),
                timepoint = if_else(str_detect(timepoint, "pre"),
                                    "pre", "post"),
                sample_id = str_remove(sample_id, "^original sample id: ")) %>% 
  dplyr::mutate(sample_id = str_remove(sample_id, "T\\d{1}")) %>% # note that there are some subjects with just a pre or a post biopsy
  dplyr::arrange(sample_id) %>% 
  dplyr::mutate(replicate_id = consecutive_id(sample_id)) %>% 
  dplyr::group_by(sample_id) %>% 
  dplyr::filter(n() > 1) %>% 
  dplyr::ungroup() %>% 
  dplyr::mutate(exercise_status = factor(exercise_status,
                                         levels = c("untrained", "trained"))) %>% 
  dplyr::inner_join(add_meta,
                    join_by(sample_id))

# only keep corresponding samples
counts <- tbl[ , colnames(tbl) %in% sample_metadata$subject]

# check order
#identical(colnames(counts), sample_metadata$subject) # FALSE

# fix order
counts <- counts[ , match(sample_metadata$subject, colnames(counts))]

# check order
#identical(colnames(counts), sample_metadata$subject) # TRUE


# make DGEList object
x <- edgeR::DGEList(counts = counts,
                    samples = sample_metadata)
#dim(x)
# 39376    20

# update samples table
# concatenate gender and exercise status
x$samples <- x$samples %>% 
              data.frame() %>% 
              tidyr::unite(col = "gender_ex",
                           c(gender, exercise_status),
                          remove = FALSE)

# filter out lowly expressed genes
keep.exprs <- edgeR::filterByExpr(x, group = x$samples$exercise_status) 
x <- x[keep.exprs,, keep.lib.sizes=FALSE]
#dim(x)
# 13926    20

x <- edgeR::calcNormFactors(x, method = "TMM")

# normalize count data
lcpm_x <- edgeR::cpm(x$counts, log = TRUE)

# fit
# make design matrix
design <- model.matrix(~0 + gender_ex + dominant_leg + BMI, x$samples)
#design

# fit model with voomLmFit
fit_vfit <- edgeR::voomLmFit(
  x,
  design,
  block = x$samples$sample_id,  
  sample.weights = TRUE,        
  plot = FALSE                   
)

# make contrast matrix
cont.mat <- limma::makeContrasts(
  women_comp = gender_exfemale_trained - gender_exfemale_untrained,
  men_comp = gender_exmale_trained - gender_exmale_untrained,
  interaction_effect = (gender_exfemale_trained - gender_exfemale_untrained) - (gender_exmale_trained - gender_exmale_untrained),
  levels = colnames(design)
)

fit_vfit.cont <- limma::contrasts.fit(fit_vfit, cont.mat)

efit_vfit <- limma::eBayes(fit_vfit.cont)

# Female
akshay_gse60590_f_results <- topTable(efit_vfit, 
                                      coef = "women_comp", 
                                      number = nrow(x)) %>% 
                                tibble::rownames_to_column("GeneID") %>% 
                                dplyr::mutate(GeneID = as.integer(GeneID)) %>% 
                                dplyr::left_join(annot[ , 1:2], join_by(GeneID)) %>% 
                                dplyr::relocate(Symbol, .after = GeneID)

akshay_gse60590_f_results %>% 
  dplyr::mutate(across(where(is.numeric), ~round(.x, 3))) %>% 
  DT::datatable(
    caption = "EpiTrain DE results (Female)"
  )

# Male
akshay_gse60590_m_results <- topTable(efit_vfit, 
                                      coef = "men_comp", 
                                      number = nrow(x)) %>% 
                                tibble::rownames_to_column("GeneID") %>% 
                                dplyr::mutate(GeneID = as.integer(GeneID)) %>% 
                                dplyr::left_join(annot[ , 1:2], join_by(GeneID)) %>% 
                                dplyr::relocate(Symbol, .after = GeneID)

akshay_gse60590_m_results %>% 
  dplyr::mutate(across(where(is.numeric), ~round(.x, 3))) %>% 
  DT::datatable(
    caption = "EpiTrain DE results (Male)"
  )
```

### MotrPac

```{r}
#| label: motrpac-de
#| echo: false
#| message: false
#| warning: false
# load data
data("TRNSCRPT_GN_DA.rda")
data("RAT_TO_HUMAN_GENE")

rat_to_hum_df <- RAT_TO_HUMAN_GENE %>% 
                      dplyr::select(RAT_ENSEMBL_ID, HUMAN_ORTHOLOG_NCBI_GENE_ID, HUMAN_ORTHOLOG_SYMBOL) %>% 
                      tidyr::drop_na() %>% 
                      dplyr::distinct()

motr_male <- TRNSCRPT_GN_DA[["trained_vs_SED"]] %>% 
              dplyr::filter(contrast == "M_8W - M_SED") %>% 
              dplyr::inner_join(
                rat_to_hum_df,
                join_by(featureName == RAT_ENSEMBL_ID)
              ) %>% 
              dplyr::select(HUMAN_ORTHOLOG_NCBI_GENE_ID, logFC, t, P.Value, adj.P.Val, z, HUMAN_ORTHOLOG_SYMBOL) %>% 
              tidyr::drop_na() %>% 
              dplyr::group_by(HUMAN_ORTHOLOG_NCBI_GENE_ID) %>% 
              dplyr::slice_max(abs(z)) %>% # pick highest abosolute z value per group     
              dplyr::ungroup() %>% 
              tibble::column_to_rownames("HUMAN_ORTHOLOG_NCBI_GENE_ID") %>% 
              dplyr::relocate(z)

motr_female <- TRNSCRPT_GN_DA[["trained_vs_SED"]] %>% 
                dplyr::filter(contrast == "F_8W - F_SED") %>% 
                dplyr::inner_join(
                rat_to_hum_df,
                join_by(featureName == RAT_ENSEMBL_ID)
              ) %>% 
              dplyr::select(HUMAN_ORTHOLOG_NCBI_GENE_ID, logFC, t, P.Value, adj.P.Val, z, HUMAN_ORTHOLOG_SYMBOL) %>% 
              tidyr::drop_na() %>% 
              dplyr::group_by(HUMAN_ORTHOLOG_NCBI_GENE_ID) %>% 
              dplyr::slice_max(abs(z)) %>% # pick highest abosolute z value per group     
              dplyr::ungroup() %>% 
              tibble::column_to_rownames("HUMAN_ORTHOLOG_NCBI_GENE_ID") %>% 
              dplyr::relocate(z)

# present results
motr_female %>% 
  data.frame() %>% 
  tibble::rownames_to_column("GeneID") %>% 
  dplyr::mutate(across(where(is.numeric), ~round(.x, 3))) %>% 
  dplyr::arrange(adj.P.Val) %>% 
  DT::datatable(
    caption = "MotrPac DE results (Female)"
  )

motr_male %>% 
  data.frame() %>% 
  tibble::rownames_to_column("GeneID") %>% 
  dplyr::mutate(across(where(is.numeric), ~round(.x, 3))) %>% 
  dplyr::arrange(adj.P.Val) %>% 
  DT::datatable(
    caption = "MotrPac DE results (Male)"
  )
```

## Enrichment Analysis

### EpiTrain

```{r}
#| label: epitrain-ea
#| echo: false
#| message: false
#| warning: false

# download and prep gene sets
# Use MSigDB v2023.2.Hs to be consistent with paper
c5.all.v2023.2.Hs.entrez <- fgsea::gmtPathways("https://data.broadinstitute.org/gsea-msigdb/msigdb/release/2023.2.Hs/c5.all.v2023.2.Hs.entrez.gmt")
c8.all.v2023.2.Hs.entrez <- fgsea::gmtPathways("https://data.broadinstitute.org/gsea-msigdb/msigdb/release/2023.2.Hs/c8.all.v2023.2.Hs.entrez.gmt")

# combine c5 and c8 gene sets
comb.c5.c8.Hs.entrez <- c(c5.all.v2023.2.Hs.entrez, c8.all.v2023.2.Hs.entrez)
length(comb.c5.c8.Hs.entrez)
# 16838

# process gene sets
background_epi <- unique(rownames(x))

# Restrict genes sets to background of genes
comb.c5.c8.Hs.entrez_filt <- filterSets(
    x = comb.c5.c8.Hs.entrez, 
    background = background_epi, 
    min_size = 10L # min. overlap of each set with background
)

#length(comb.c5.c8.Hs.entrez_filt) # 10418 gene sets pass

# Calculate proportion of overlap with background
setSizes_epi <- lengths(comb.c5.c8.Hs.entrez_filt)
setSizesOld_epi <- lengths(comb.c5.c8.Hs.entrez)[names(comb.c5.c8.Hs.entrez_filt)]
overlapProp_epi <- setSizes_epi / setSizesOld_epi

#plot(setSizesOld_epi, overlapProp_epi, main = "Set Size vs. Overlap")

# clusterSets with default arguments
clusterDF_epi <- clusterSets(x = comb.c5.c8.Hs.entrez_filt, 
                             type = "jaccard", 
                             cutoff = 0.85, 
                             method = "complete",
                             h = 0.9)
## Use clusterSets output to select sets to retain for analysis
clusterDF_epi$overlap_prop <- overlapProp_epi[clusterDF_epi$set] # overlap proportion
clusterDF_epi$n_char <- nchar(clusterDF_epi$set) # length of set description

# Reorder rows so that the first set in each cluster is the one to keep
o_epi <- with(clusterDF_epi, 
          order(cluster, set_size, overlap_prop, n_char, set,
                decreasing = c(FALSE, TRUE, TRUE, FALSE, TRUE),
                method = "radix"))
clusterDF_epi <- clusterDF_epi[o_epi, ]

# Subset filtered gene sets
comb.c5.c8.Hs.entrez_filt <- comb.c5.c8.Hs.entrez_filt[with(clusterDF_epi, set[!duplicated(cluster)])]

#length(comb.c5.c8.Hs.entrez_filt) # 10215 gene sets

## camera (see Notes below)

## cameraPR
## For limma + voom
# Matrix of z-score equivalents of moderated t-statistics
# female
modz_epi <- with(efit_vfit, zscoreT(x = t, df = df.total))

res_epi <- cameraPR(statistic = modz_epi, 
                    index = comb.c5.c8.Hs.entrez_filt)

epi_camerapr_f <- res_epi %>% 
                    dplyr::filter(Contrast == "women_comp")
# male
epi_camerapr_m <- res_epi %>% 
                    dplyr::filter(Contrast == "men_comp")

# present results
epi_camerapr_f %>% 
  dplyr::filter(
    str_detect(GeneSet, "^GO")
  ) %>% 
  dplyr::mutate(across(where(is.numeric), ~round(.x, 3))) %>% 
  DT::datatable(
    caption = "EpiTrain Female (cameraPR results)"
  )

epi_camerapr_m %>% 
  dplyr::filter(
    str_detect(GeneSet, "^GO")
  ) %>% 
  dplyr::mutate(across(where(is.numeric), ~round(.x, 3))) %>% 
  DT::datatable(
    caption = "EpiTrain Male (cameraPR results)"
  )
```

### MotrPac

```{r}
#| label: motrpac-ea
#| echo: false
#| message: false
#| warning: false

# process gene sets
background_motr <- unique(rownames(motr_male))

# Restrict genes sets to background of genes
comb.c5.c8.Hs.entrez_filt_motr <- filterSets(
    x = comb.c5.c8.Hs.entrez, 
    background = background_motr, 
    min_size = 10L 
)

#length(comb.c5.c8.Hs.entrez_filt_motr) # 10836 gene sets pass

# Calculate proportion of overlap with background
setSizes_motr <- lengths(comb.c5.c8.Hs.entrez_filt_motr)
setSizesOld_motr <- lengths(comb.c5.c8.Hs.entrez)[names(comb.c5.c8.Hs.entrez_filt_motr)]
overlapProp_motr <- setSizes_motr / setSizesOld_motr

#plot(setSizesOld_motr, overlapProp_motr, main = "Set Size vs. Overlap")

# clusterSets with default arguments
clusterDF_motr <- clusterSets(x = comb.c5.c8.Hs.entrez_filt_motr, 
                             type = "jaccard", 
                             cutoff = 0.85, 
                             method = "complete",
                             h = 0.9)
## Use clusterSets output to select sets to retain for analysis
clusterDF_motr$overlap_prop <- overlapProp_motr[clusterDF_motr$set] # overlap proportion
clusterDF_motr$n_char <- nchar(clusterDF_motr$set) # length of set description

# Reorder rows so that the first set in each cluster is the one to keep
o_motr <- with(clusterDF_motr, 
          order(cluster, set_size, overlap_prop, n_char, set,
                decreasing = c(FALSE, TRUE, TRUE, FALSE, TRUE),
                method = "radix"))
clusterDF_motr <- clusterDF_motr[o_motr, ]

comb.c5.c8.Hs.entrez_filt_motr <- comb.c5.c8.Hs.entrez_filt_motr[with(clusterDF_motr, set[!duplicated(cluster)])]

#length(comb.c5.c8.Hs.entrez_filt_motr) # 10640 gene sets

## cameraPR
# female
res_motr_f <- cameraPR(statistic = as.matrix(motr_female[1]), 
                      index = comb.c5.c8.Hs.entrez_filt_motr)
# male
res_motr_m <- cameraPR(statistic = as.matrix(motr_male[1]), 
                      index = comb.c5.c8.Hs.entrez_filt_motr)

# present results
res_motr_f %>% 
  dplyr::filter(str_detect(GeneSet, "^GO")) %>% 
  dplyr::mutate(across(where(is.numeric), ~round(.x, 3))) %>% 
  DT::datatable(
    caption = "MotrPac Female (cameraPR results)"
  )

res_motr_m %>% 
  dplyr::filter(str_detect(GeneSet, "^GO")) %>% 
  dplyr::mutate(across(where(is.numeric), ~round(.x, 3))) %>% 
  DT::datatable(
    caption = "MotrPac Male (cameraPR results)"
  )
```

## EpiTrain versus MotrPac

### Differential Expression Analysis

Comparing t-statistics.

```{r}
#| label: epi-vs-motr-de
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "EpiTrain versus MotrPac (DE results)"
#| fig-subcap: 
#|   - "Female Comparison"
#|   - "Male Comparison"
#|   - "Epi Female vs Motr Male"
#|   - "Epi Male vs Motr Female"
#| layout-ncol: 1
#| column: page
#| fig-width: 10
#| fig-height: 7

# female
epi_vs_motr_de_f <- akshay_gse60590_f_results %>% 
  dplyr::inner_join(
    motr_female,
    join_by(Symbol == HUMAN_ORTHOLOG_SYMBOL),
    suffix = c("_epi", "_motr")
  ) 

cor(epi_vs_motr_de_f$t_epi, epi_vs_motr_de_f$t_motr,
    method = "spearman") # 0.15
cor.test(epi_vs_motr_de_f$t_epi, epi_vs_motr_de_f$t_motr,
    method = "spearman") # p-value < 2.2e-16

epi_vs_motr_de_f_plot <- epi_vs_motr_de_f %>% 
  ggplot(aes(t_epi, t_motr, label = Symbol)) +
  geom_point() +
  geom_abline(linetype = "dashed") +
  theme_bw() +
  labs(title = "Epi vs Motrpac - Female (t-stats); Spearman's cor coeff: 0.15")
ggplotly(epi_vs_motr_de_f_plot)

# male
epi_vs_motr_de_m <- akshay_gse60590_m_results %>% 
  dplyr::inner_join(
    motr_male,
    join_by(Symbol == HUMAN_ORTHOLOG_SYMBOL),
    suffix = c("_epi", "_motr")
  ) 

cor(epi_vs_motr_de_m$t_epi, epi_vs_motr_de_m$t_motr,
    method = "spearman") # 0.026
cor.test(epi_vs_motr_de_m$t_epi, epi_vs_motr_de_m$t_motr,
    method = "spearman") # p-value = 0.007003

epi_vs_motr_de_m_plot <- epi_vs_motr_de_m %>% 
  ggplot(aes(t_epi, t_motr, label = Symbol)) +
  geom_point() +
  geom_abline(linetype = "dashed") +
  theme_bw() +
  labs(title = "Epi vs Motrpac - Male (t-stats); Spearman's cor coeff: 0.026")
ggplotly(epi_vs_motr_de_m_plot)

# Epi Female vs Motrpac Male
epi_vs_motr_de_combo <- epi_vs_motr_de_f %>% 
  dplyr::inner_join(
  epi_vs_motr_de_m,
  join_by(Symbol),
  suffix = c("_f", "_m")
  )

cor(epi_vs_motr_de_combo$t_epi_f, epi_vs_motr_de_combo$t_motr_m,
    method = "spearman") # 0.12
cor.test(epi_vs_motr_de_combo$t_epi_f, epi_vs_motr_de_combo$t_motr_m,
    method = "spearman") # p-value < 2.2e-16

epi_vs_motr_combo_fvm_plot <- epi_vs_motr_de_combo %>% 
  ggplot(aes(t_epi_f, t_motr_m, label = Symbol)) +
  geom_point() +
  geom_abline(linetype = "dashed") +
  theme_bw() +
  labs(title = "Epi Female vs Motrpac Male (t-stats); Spearman's cor coeff: 0.12")
ggplotly(epi_vs_motr_combo_fvm_plot)

# Epi Male vs Motrpac Female
cor(epi_vs_motr_de_combo$t_epi_m, epi_vs_motr_de_combo$t_motr_f,
    method = "spearman") # 0.20
cor.test(epi_vs_motr_de_combo$t_epi_m, epi_vs_motr_de_combo$t_motr_f,
    method = "spearman") # p-value < 2.2e-16

epi_vs_motr_combo_mvf_plot <- epi_vs_motr_de_combo %>% 
  ggplot(aes(t_epi_m, t_motr_f, label = Symbol)) +
  geom_point() +
  geom_abline(linetype = "dashed") +
  theme_bw() +
  labs(title = "Epi Male vs Motrpac Female (t-stats); Spearman's cor coeff: 0.20")
ggplotly(epi_vs_motr_combo_mvf_plot)
```

<!-- ### Enrichment Analysis -->

```{r}
#| label: epi-vs-motr-ea
#| echo: false
#| include: false
#| message: false
#| warning: false
#| fig-cap: "EpiTrain versus MotrPac (cameraPR results)"
#| fig-subcap: 
#|   - "Female Comparison"
#|   - "Male Comparison"
#|   - "Epi Female vs Motr Male"
#|   - "Epi Male vs Motr Female"
#|   - "Epi Female vs Male"
#|   - "Motr Female vs Male"
#| layout-ncol: 1
#| column: page
#| fig-width: 10
#| fig-height: 7

# Epi women vs men
epi_fvm_camera <- epi_camerapr_f %>% 
  dplyr::select(GeneSet, ZScore) %>% 
  dplyr::inner_join(
    epi_camerapr_m %>% 
      dplyr::select(GeneSet, ZScore),
    join_by(GeneSet),
    suffix = c("_f", "_m")
  )

# Motrpac Female vs Male
motr_fvm_camera <- res_motr_f %>% 
  dplyr::select(GeneSet, ZScore) %>% 
  dplyr::inner_join(
    res_motr_m %>% 
      dplyr::select(GeneSet, ZScore),
    join_by(GeneSet),
    suffix = c("_f", "_m")
  )

# Epi vs Motrpac - female
epi_v_motr_camera <- epi_fvm_camera %>% 
                      dplyr::inner_join(
                        motr_fvm_camera,
                        join_by(GeneSet),
                        suffix = c("_epi", "_motr")
                      ) %>% 
                      dplyr::filter(str_detect(GeneSet, "^GO"))

#cor(epi_v_motr_camera$ZScore_f_epi, epi_v_motr_camera$ZScore_f_motr) 
# 0.64 overall
# 0.65 filtering on GO terms
# cor.test(epi_v_motr_camera$ZScore_f_epi, epi_v_motr_camera$ZScore_f_motr)$p.value
# 0 filtering on GO terms

epi_v_motr_camera_f_plot <-epi_v_motr_camera %>% 
  ggplot(aes(ZScore_f_epi, ZScore_f_motr, label = GeneSet)) +
  geom_point() +
  geom_abline(linetype = "dashed") +
  theme_bw() +
  labs(title = "Epi vs Motrpac - Female (camera); Pearson's cor coeff: 0.65")
ggplotly(epi_v_motr_camera_f_plot)

# concordantly upregulated - OXPHOS related terms
# concordantly downregulated - splicing terms?

# Epi vs Motrpac - male
#cor(epi_v_motr_camera$ZScore_m_epi, epi_v_motr_camera$ZScore_m_motr) 
# -0.09 overall
# -0.10 filtering on GO terms
# cor.test(epi_v_motr_camera$ZScore_m_epi, epi_v_motr_camera$ZScore_m_motr)$p.value
# 0 filtering on GO terms

epi_v_motr_camera_m_plot <- epi_v_motr_camera %>% 
  ggplot(aes(ZScore_m_epi, ZScore_m_motr, label = GeneSet)) +
  geom_point() +
  geom_abline(linetype = "dashed") +
  theme_bw() +
  labs(title = "Epi vs Motrpac - Male (camera); Pearson's cor coeff: -0.10")
ggplotly(epi_v_motr_camera_m_plot)
# OXPHOS-related terms seem to be more potently induced in motrpac
# Endothelial cell terms seem to be more potently induced in epi

# Epi Female vs Motrpac Male
#cor(epi_v_motr_camera$ZScore_f_epi, epi_v_motr_camera$ZScore_m_motr) 
# 0.3 overall
# 0.3 filtering on GO terms
# cor.test(epi_v_motr_camera$ZScore_f_epi, epi_v_motr_camera$ZScore_m_motr)$p.value
# 0 filtering on GO terms

epi_v_motr_camera_fvm_plot <-epi_v_motr_camera %>% 
  ggplot(aes(ZScore_f_epi, ZScore_m_motr, label = GeneSet)) +
  geom_point() +
  geom_abline(linetype = "dashed") +
  theme_bw() +
  labs(title = "Epi Female vs Motrpac Male (camera); Pearson's cor coeff: 0.3")
ggplotly(epi_v_motr_camera_fvm_plot)

# Epi Female vs Motrpac Male
#cor(epi_v_motr_camera$ZScore_m_epi, epi_v_motr_camera$ZScore_f_motr) 
#  overall
# 0.48 filtering on GO terms
#cor.test(epi_v_motr_camera$ZScore_m_epi, epi_v_motr_camera$ZScore_f_motr)$p.value 
# 0 filtering on GO terms

epi_v_motr_camera_mvf_plot <-epi_v_motr_camera %>% 
  ggplot(aes(ZScore_m_epi, ZScore_f_motr, label = GeneSet)) +
  geom_point() +
  geom_abline(linetype = "dashed") +
  theme_bw() +
  labs(title = "Epi Male vs Motrpac Female (camera); Pearson's cor coeff: 0.48")
ggplotly(epi_v_motr_camera_mvf_plot)

# Epi Female vs Male
#cor(epi_v_motr_camera$ZScore_f_epi, epi_v_motr_camera$ZScore_m_epi) 
#  overall
# 0.67 filtering on GO terms

epi_v_motr_camera_epi_fvm_plot <- epi_v_motr_camera %>% 
  ggplot(aes(ZScore_f_epi, ZScore_m_epi, label = GeneSet)) +
  geom_point() +
  geom_abline(linetype = "dashed") +
  theme_bw() +
  labs(title = "Epi Male vs Female (camera); Pearson's cor coeff: 0.67")
ggplotly(epi_v_motr_camera_epi_fvm_plot)

# Motr Female vs Male
#cor(epi_v_motr_camera$ZScore_f_motr, epi_v_motr_camera$ZScore_m_motr) 
#  overall
# 0.58 filtering on GO terms

epi_v_motr_camera_motr_fvm_plot <-epi_v_motr_camera %>% 
  ggplot(aes(ZScore_f_motr, ZScore_m_motr, label = GeneSet)) +
  geom_point() +
  geom_abline(linetype = "dashed") +
  theme_bw() +
  labs(title = "Motrpac Male vs Female (camera); Pearson's cor coeff: 0.58")
ggplotly(epi_v_motr_camera_motr_fvm_plot)
```

<!-- ## Additional plots -->

```{r}
#| label: additional-plots
#| echo: false
#| message: false
#| warning: false
#| include: false

epi_v_motr_camera %>% 
  ggplot(aes(ZScore_f_epi, ZScore_m_epi, label = GeneSet)) +
  geom_point() +
  geom_xsidedensity(aes(y = after_stat(density)), alpha = 0.5) +
  geom_ysidedensity(aes(x = after_stat(density)), alpha = 0.5) + 
  geom_abline(linetype = "dashed") +
  theme_bw() +
  labs(title = "Epi Male vs Female (camera); Pearson's cor coeff: 0.65")

epi_v_motr_camera %>% 
  ggplot(aes(ZScore_f_motr, ZScore_m_motr, label = GeneSet)) +
  geom_point() +
  geom_xsidedensity(aes(y = after_stat(density)), alpha = 0.5) +
  geom_ysidedensity(aes(x = after_stat(density)), alpha = 0.5) + 
  geom_abline(linetype = "dashed") +
  theme_bw() +
  labs(title = "Motrpac Male vs Female (camera); Pearson's cor coeff: 0.58")
```

## Figures for manuscript

```{r}
#| label: manuscript-figures
#| echo: false
#| message: false
#| warning: false

# Spearman correlation coefficients
cor(epi_v_motr_camera$ZScore_f_epi, epi_v_motr_camera$ZScore_f_motr, method = "spearman") # 0.49
cor(epi_v_motr_camera$ZScore_f_epi, epi_v_motr_camera$ZScore_m_motr, method = "spearman") # 0.015
cor(epi_v_motr_camera$ZScore_m_epi, epi_v_motr_camera$ZScore_f_motr, method = "spearman") # 0.48
cor(epi_v_motr_camera$ZScore_m_epi, epi_v_motr_camera$ZScore_m_motr, method = "spearman") # -0.20

# p-values
cor.test(epi_v_motr_camera$ZScore_f_epi, epi_v_motr_camera$ZScore_f_motr, method = "spearman") # p-value < 2.2e-16
cor.test(epi_v_motr_camera$ZScore_f_epi, epi_v_motr_camera$ZScore_m_motr, method = "spearman") # p-value = 0.263
cor.test(epi_v_motr_camera$ZScore_m_epi, epi_v_motr_camera$ZScore_f_motr, method = "spearman") # p-value < 2.2e-16
cor.test(epi_v_motr_camera$ZScore_m_epi, epi_v_motr_camera$ZScore_m_motr, method = "spearman") # p-value < 2.2e-16

# What are the pathways with highest average z-score across all comparisons?
# pick top N
top_pathways <- epi_v_motr_camera %>% 
  dplyr::rowwise() %>% 
  dplyr::mutate(z_avg = mean(c_across(starts_with("Z")), na.rm = TRUE)) %>% 
  dplyr::ungroup() %>% 
  dplyr::arrange(desc(z_avg)) %>% 
  dplyr::slice_max(z_avg, n = 50) %>% 
  dplyr::pull(GeneSet)

# Epi F vs Motr F
f_v_f <- epi_v_motr_camera %>% 
  ggplot(aes(ZScore_f_epi, ZScore_f_motr, label = GeneSet)) +
  geom_point() +
  geom_point(data = . %>% filter(GeneSet %in% top_pathways),
             col = "red") +
  # geom_text_repel(data = . %>% filter(GeneSet %in% top_pathways),
  #                 aes(label = GeneSet),
  #                 size = 3) + # not showing because the plot get too cluttered!
  geom_xsidedensity(aes(y = after_stat(density)), alpha = 0.5) +
  geom_ysidedensity(aes(x = after_stat(density)), alpha = 0.5) + 
  geom_abline(linetype = "dashed") +
  theme_bw() +
  labs(title = "EpiTrain Female vs MoTrPAC Female",
       subtitle = "Spearman's cor coeff: 0.49\np-value < 2.2e-16",
       x = "EpiTrain Female (Z-score)",
       y = "MoTrPAC Female (Z-score)") +
  theme(ggside.panel.scale.x = .2,
        ggside.panel.scale.y = .2,
        axis.title = element_text(face = "bold",
                                  size = 15),
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(face = "bold"),
        axis.text = element_text(face = "bold",
                                 size = 10))

# Epi F vs Motr M
f_v_m <- epi_v_motr_camera %>% 
  ggplot(aes(ZScore_f_epi, ZScore_m_motr, label = GeneSet)) +
  geom_point() +
  geom_point(data = . %>% filter(GeneSet %in% top_pathways),
             col = "red") +
  # geom_text_repel(data = . %>% filter(GeneSet %in% top_pathways),
  #                 aes(label = GeneSet),
  #                 size = 3) +
  geom_xsidedensity(aes(y = after_stat(density)), alpha = 0.5) +
  geom_ysidedensity(aes(x = after_stat(density)), alpha = 0.5) + 
  geom_abline(linetype = "dashed") +
  theme_bw() +
  labs(title = "EpiTrain Female vs MoTrPAC Male",
       subtitle = "Spearman's cor coeff: 0.015\np-value = 0.263",
       x = "EpiTrain Female (Z-score)",
       y = "MoTrPAC Male (Z-score)") +
  theme(ggside.panel.scale.x = .2,
        ggside.panel.scale.y = .2,
        axis.title = element_text(face = "bold",
                                  size = 15),
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(face = "bold"),
        axis.text = element_text(face = "bold",
                                 size = 10))

# Epi M vs Motr F
m_v_f <- epi_v_motr_camera %>% 
  ggplot(aes(ZScore_m_epi, ZScore_f_motr, label = GeneSet)) +
  geom_point() +
  geom_point(data = . %>% filter(GeneSet %in% top_pathways),
             col = "red") +
  # geom_text_repel(data = . %>% filter(GeneSet %in% top_pathways),
  #                 aes(label = GeneSet),
  #                 size = 3) +
  geom_xsidedensity(aes(y = after_stat(density)), alpha = 0.5) +
  geom_ysidedensity(aes(x = after_stat(density)), alpha = 0.5) + 
  geom_abline(linetype = "dashed") +
  theme_bw() +
  labs(title = "EpiTrain Male vs MoTrPAC Female",
       subtitle = "Spearman's cor coeff: 0.48\np-value < 2.2e-16",
       x = "EpiTrain Male (Z-score)",
       y = "MoTrPAC Female (Z-score)") +
  theme(ggside.panel.scale.x = .2,
        ggside.panel.scale.y = .2,
        axis.title = element_text(face = "bold",
                                  size = 15),
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(face = "bold"),
        axis.text = element_text(face = "bold",
                                 size = 10))

# Epi M vs Motr M
m_v_m <- epi_v_motr_camera %>% 
  ggplot(aes(ZScore_m_epi, ZScore_m_motr, label = GeneSet)) +
  geom_point() +
  geom_point(data = . %>% filter(GeneSet %in% top_pathways),
             col = "red") +
  # geom_text_repel(data = . %>% filter(GeneSet %in% top_pathways),
  #                 aes(label = GeneSet),
  #                 size = 3) +
  geom_xsidedensity(aes(y = after_stat(density)), alpha = 0.5) +
  geom_ysidedensity(aes(x = after_stat(density)), alpha = 0.5) + 
  geom_abline(linetype = "dashed") +
  theme_bw() +
  labs(title = "EpiTrain Male vs MoTrPAC Male",
       subtitle = "Spearman's cor coeff: -0.20\np-value < 2.2e-16",
       x = "EpiTrain Male (Z-score)",
       y = "MoTrPAC Male (Z-score)") +
  theme(ggside.panel.scale.x = .2,
        ggside.panel.scale.y = .2,
        axis.title = element_text(face = "bold",
                                  size = 15),
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(face = "bold"),
        axis.text = element_text(face = "bold",
                                 size = 10))

manuscript_plot <- (f_v_f + f_v_m) /
                      (m_v_f + m_v_m) +
                         patchwork::plot_annotation(tag_levels = "A")

manuscript_plot
# ggsave(filename = "figures/camera_plot_for_manuscript_AB_082825.png",
#        plot = manuscript_plot,
#        width = 15,
#        height = 15)

ggsave(filename = "figures/camera_plot_for_manuscript_AB_110725.png",
       plot = manuscript_plot,
       width = 15,
       height = 15)
```

Figure caption

Figure 6. Comparison of transcriptomic responses to exercise in humans and rats at the pathway level. Comparison of cameraPR results for (A) EpiTrain Female versus MoTrPAC Female; (B) EpiTrain Female versus MoTrPAC Male; (C) EpiTrain Male versus MoTrPAC Female; (D) EpiTrain Male versus MoTrPAC Male. Points highlighted in red indicate gene sets with the highest average Z-score across all comparisons (see text for details).

## Motrpac (Proteomics)

```{r}
#| label: motr-prot
#| echo: false
#| message: false
#| warning: false
# load data
load("data/PROT_GN_DA.rda")

rat_to_hum_df_prot <- RAT_TO_HUMAN_GENE %>% 
                      dplyr::select(RAT_SYMBOL, RAT_ENSEMBL_ID, HUMAN_ORTHOLOG_NCBI_GENE_ID, HUMAN_ORTHOLOG_SYMBOL) %>% 
                      tidyr::drop_na() %>% 
                      dplyr::distinct()

motr_male_prot <- PROT_GN_DA[["trained_vs_SED"]] %>% 
              dplyr::filter(contrast == "M_8W - M_SED") %>% 
              dplyr::inner_join(
                rat_to_hum_df_prot,
                join_by(gene_symbol == RAT_SYMBOL)
              ) %>% 
              dplyr::select(HUMAN_ORTHOLOG_NCBI_GENE_ID, logFC, t, P.Value, adj.P.Val, z, HUMAN_ORTHOLOG_SYMBOL) %>% 
              tidyr::drop_na() %>% 
              dplyr::group_by(HUMAN_ORTHOLOG_NCBI_GENE_ID) %>% 
              dplyr::slice_max(abs(z)) %>% # pick highest abosolute z value per group     
              dplyr::ungroup() %>% 
              dplyr::distinct() %>% 
              tibble::column_to_rownames("HUMAN_ORTHOLOG_NCBI_GENE_ID") %>% 
              dplyr::relocate(z)

motr_female_prot <- PROT_GN_DA[["trained_vs_SED"]] %>% 
                dplyr::filter(contrast == "F_8W - F_SED") %>% 
                dplyr::inner_join(
                rat_to_hum_df_prot,
                join_by(gene_symbol == RAT_SYMBOL)
              ) %>% 
              dplyr::select(HUMAN_ORTHOLOG_NCBI_GENE_ID, logFC, t, P.Value, adj.P.Val, z, HUMAN_ORTHOLOG_SYMBOL) %>% 
              tidyr::drop_na() %>% 
              dplyr::group_by(HUMAN_ORTHOLOG_NCBI_GENE_ID) %>% 
              dplyr::slice_max(abs(z)) %>% # pick highest abosolute z value per group     
              dplyr::ungroup() %>% 
              dplyr::distinct() %>% 
              tibble::column_to_rownames("HUMAN_ORTHOLOG_NCBI_GENE_ID") %>% 
              dplyr::relocate(z)

# process gene sets
background_motr_prot <- unique(c(rownames(motr_male_prot), rownames(motr_male_prot)))

# Restrict genes sets to background of genes
comb.c5.c8.Hs.entrez_filt_motr_prot <- filterSets(
    x = comb.c5.c8.Hs.entrez, 
    background = background_motr_prot, 
    min_size = 10L 
)

#length(comb.c5.c8.Hs.entrez_filt_motr_prot) # 7328 gene sets pass

# Calculate proportion of overlap with background
setSizes_motr_prot <- lengths(comb.c5.c8.Hs.entrez_filt_motr_prot)
setSizesOld_motr_prot <- lengths(comb.c5.c8.Hs.entrez)[names(comb.c5.c8.Hs.entrez_filt_motr_prot)]
overlapProp_motr_prot <- setSizes_motr_prot / setSizesOld_motr_prot

#plot(setSizesOld_motr_prot, overlapProp_motr_prot, main = "Set Size vs. Overlap")

# clusterSets with default arguments
clusterDF_motr_prot <- clusterSets(x = comb.c5.c8.Hs.entrez_filt_motr_prot, 
                             type = "jaccard", 
                             cutoff = 0.85, 
                             method = "complete",
                             h = 0.9)
## Use clusterSets output to select sets to retain for analysis
clusterDF_motr_prot$overlap_prop <- overlapProp_motr_prot[clusterDF_motr_prot$set] # overlap proportion
clusterDF_motr_prot$n_char <- nchar(clusterDF_motr_prot$set) # length of set description

# Reorder rows so that the first set in each cluster is the one to keep
o_motr_prot <- with(clusterDF_motr_prot, 
          order(cluster, set_size, overlap_prop, n_char, set,
                decreasing = c(FALSE, TRUE, TRUE, FALSE, TRUE),
                method = "radix"))
clusterDF_motr_prot <- clusterDF_motr_prot[o_motr_prot, ]

comb.c5.c8.Hs.entrez_filt_motr_prot <- comb.c5.c8.Hs.entrez_filt_motr_prot[with(clusterDF_motr_prot, set[!duplicated(cluster)])]

#length(comb.c5.c8.Hs.entrez_filt_motr_prot) # 7072 gene sets

## cameraPR
# female
res_motr_f_prot <- cameraPR(statistic = as.matrix(motr_female_prot[1]), 
                      index = comb.c5.c8.Hs.entrez_filt_motr_prot)
# male
res_motr_m_prot <- cameraPR(statistic = as.matrix(motr_male_prot[1]), 
                      index = comb.c5.c8.Hs.entrez_filt_motr_prot)

# present results
res_motr_f_prot %>% 
  dplyr::filter(str_detect(GeneSet, "^GO")) %>% 
  dplyr::mutate(across(where(is.numeric), ~round(.x, 3))) %>% 
  DT::datatable(
    caption = "MotrPac Female (cameraPR results)"
  )

res_motr_m_prot %>% 
  dplyr::filter(str_detect(GeneSet, "^GO")) %>% 
  dplyr::mutate(across(where(is.numeric), ~round(.x, 3))) %>% 
  DT::datatable(
    caption = "MotrPac Male (cameraPR results)"
  )
```

### MotrPac RNAseq vs Proteomics (cameraPR)

```{r}
#| label: motr-rnaseq-vs-prot
#| echo: false
#| message: false
#| warning: false


motr_rna_v_prot_f <- res_motr_f %>% 
  dplyr::inner_join(
    res_motr_f_prot,
    join_by(GeneSet),
    suffix = c("_rnaseq", "_prot")
  ) %>% 
  dplyr::filter(str_detect(GeneSet, "^GO"))

# cor(motr_rna_v_prot_f$ZScore_rnaseq, motr_rna_v_prot_f$ZScore_prot) 
# 0.67 overall
# 0.66 filtering on GO terms

motr_rna_v_prot_f_plot <- motr_rna_v_prot_f %>% 
  ggplot(aes(ZScore_rnaseq, ZScore_prot, label = GeneSet)) +
  geom_point() +
  geom_abline(linetype = "dashed") +
  theme_bw() +
  labs(title = "Motrpac Female - RNAseq vs Proteomics (camera)",
       subtitle = "Correlation coefficient: 0.67")
ggplotly(motr_rna_v_prot_f_plot)

motr_rna_v_prot_m <- res_motr_m %>% 
  dplyr::inner_join(
    res_motr_m_prot,
    join_by(GeneSet),
    suffix = c("_rnaseq", "_prot")
  ) %>% 
  dplyr::filter(str_detect(GeneSet, "^GO"))

# cor(motr_rna_v_prot_m$ZScore_rnaseq, motr_rna_v_prot_m$ZScore_prot) 
# 0.48 overall
# 0.47 filtering on GO terms

motr_rna_v_prot_m_plot <- motr_rna_v_prot_m %>% 
  ggplot(aes(ZScore_rnaseq, ZScore_prot, label = GeneSet)) +
  geom_point() +
  geom_abline(linetype = "dashed") +
  theme_bw() +
  labs(title = "Motrpac Male - RNAseq vs Proteomics (camera)",
       subtitle = "Correlation coefficient: 0.48")
ggplotly(motr_rna_v_prot_m_plot)
```
